### 2012-10-12 formalism/paper.tex
added updated version of the FOOL paper including discussion of the new type system amendments. Implementation includes join constraints but stubs them for the solver.

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

\usepackage{amsmath}
\usepackage{listings}
\usepackage{proof}
\usepackage{supertabular}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{theorem}
\usepackage{url}
\usepackage{graphicx}
\usepackage{mathtools}

\newtheorem{lem}{Lemma}
\newtheorem{thm}{Theorem}
\newtheorem{defn}{Definition}
\newtheorem{cor}{Corollary}

\newcommand{\Tr}{\mathit{Tr}}
\newcommand{\secref}[1]{Section \ref{#1}}
\newcommand{\figref}[1]{Figure \ref{#1}}
\newcommand{\lstref}[1]{Listing \ref{#1}}
\newcommand{\appref}[1]{Appendix \ref{#1}}
\newcommand{\lemref}[1]{Lemma \ref{#1}}
\newcommand{\corref}[1]{Corollary \ref{#1}}

\newcommand{\rlett}{\textsc{r\_let\_term} }
\newcommand{\rletv}{\textsc{r\_let\_value} }
\newcommand{\rseqlt}{\textsc{r\_seq\_left\_term} }
\newcommand{\rseqlv}{\textsc{r\_seq\_left\_value} }
\newcommand{\rfunc}{\textsc{r\_fun\_call} }
\newcommand{\rmethc}{\textsc{r\_meth\_call} }
\newcommand{\rifexpr}{\textsc{r\_if\_term} }
\newcommand{\riftrue}{\textsc{r\_if\_true} }
\newcommand{\riffalse}{\textsc{r\_if\_false} }

\newcommand{\tunit}{\textsc{t\_unit} }
\newcommand{\ttrue}{\textsc{t\_true} }
\newcommand{\tfalse}{\textsc{t\_false} }
\newcommand{\tobj}{\textsc{t\_object} }
\newcommand{\tfundef}{\textsc{t\_fun\_def} }
\newcommand{\tlet}{\textsc{t\_let} }
\newcommand{\tif}{\textsc{t\_if} }
\newcommand{\tseq}{\textsc{t\_seq} }
\newcommand{\tfunc}{\textsc{t\_fun\_call} }
\newcommand{\tmethc}{\textsc{t\_meth\_call} }
\newcommand{\tdrop}{\textsc{t\_drop} }
\newcommand{\tsub}{\textsc{t\_sub} }
\newcommand{\tnarrow}{\textsc{t\_narrow} }
\newcommand{\twiden}{\textsc{t\_widen} }

\newcommand{\subrefl}{\textsc{sub\_refl}}
\newcommand{\subtrans}{\textsc{sub\_trans}}
\newcommand{\subfn}{\textsc{sub\_fn}}
\newcommand{\subobj}{\textsc{sub\_obj}}

\newcommand{\remapunit}{\textsc{remap\_unit}}
\newcommand{\remapbool}{\textsc{remap\_bool}}
\newcommand{\remapfunc}{\textsc{remap\_func}}
\newcommand{\remapobj}{\textsc{remap\_obj}}

\newcommand{\remapapp}[3]{\ensuremath{\mathbf{remap}(#1, #2 \gg #3)}}

\newcommand{\typerule}[4]{#1 \triangleright #2 : #3 \triangleleft #4}
\newcommand{\oprule}[4]{#1 \mid #2\;\longrightarrow\;#3 \mid #4}
\newcommand{\inferrule}[4]{\left\llbracket #1 \triangleright #2 : #3 \triangleleft #4 \right\rrbracket}
\newcommand{\inferlhs}[1]{\left\llbracket \gamma \triangleright #1 : \alpha \triangleleft \gamma' \right\rrbracket}
\newcommand{\subst}[3]{#3 \{\overline{^{#1}/_{#2}}\}}
\newcommand{\substxy}[1]{\subst{x_i}{y_i}{#1}}
\newcommand{\unitv}{\mathtt{unit}}
\newcommand{\unitt}{\mathbf{Unit}}
\newcommand{\truev}{\mathtt{true}}
\newcommand{\falsev}{\mathtt{false}}
\newcommand{\boolt}{\mathbf{Bool}}
\newcommand{\ot}[2]{#1@\overline{#2}}
\newcommand{\funv}[4]{\lambda(\overline{#1 : #2 \gg #3}).#4}
\newcommand{\funt}[3]{(\overline{#1 \gg #2}) \rightarrow #3}
\newcommand{\lett}[3]{\mathtt{let}\:#1\:\mathtt{=}\:#2\:\mathtt{in}\:#3}
\newcommand{\ift}[3]{\mathtt{if} \; #1 \; \mathtt{then} \; #2 \; \mathtt{else} \; #3}
\newcommand{\cand}{\:\wedge\:}
\newcommand{\free}[1]{\mathbf{free}\:#1}
\newcommand{\fresh}[1]{\mathbf{fresh}\:#1}

\newcommand{\qed}{$\blacksquare$}

\newenvironment{proof}{\vspace{1ex}\noindent{\bf Proof}\hspace{0.5em}}
  {\hfill\qed\vspace{1ex}}

\lstset{
    basicstyle=\ttfamily \small,
    numbers=left,
    frame=single,
    tabsize=2,
    xleftmargin=2.0em,
    xrightmargin=0.4em
}

\input{ts_defns}

\begin{document}

\conferenceinfo{TBD}{TBD} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

\titlebanner{IN-PROGRESS WORK}        % These are ignored unless
\preprintfooter{DO NOT REDISTRIBUTE WITH PERMISSION}   % 'preprint' option specified.

\title{Typestate Inference in an Imperative Calculus}

\authorinfo{Iain McGinniss \and Simon Gay}
           {University of Glasgow}
           {\{iainmcgin, simon\}@dcs.gla.ac.uk}

\maketitle

\begin{abstract}
Current typestate research is primarily focused on studying
type systems that can support particular use cases for typestate;
the type systems are often complex, and programs in such languages 
require large type annotations to allow for full static type checking. Little
work has been done to determine whether inference of such annotations is 
feasible. We present a small imperative calculus which
models interactions with typestate-constrained objects and a type inference
technique which can derive the minimal state machine that governs interactions
with the observed objects, and therefore eliminate the need for
explicit type annotations.
\end{abstract}

\category{D.2.4}{Software Engineering}{Programming by contract}
\category{F.3.3}{Studies of Program Constructs}{Type structure}
\category{D.1.5}{Object-oriented Programming}{}
\category{D.3.1}{Programming Languages}{Formal Definitions and Theory}

\keywords
typestate, type inference, constraint solving, alias control

\section{Introduction}

Typestate is ubiquitous in object oriented programs \cite{Beckman2011} and yet
very few languages provide direct support for it. Stateful objects are a source
of subtle runtime errors when their contracts are not respected, as the
point of failure is often separated from the cause in both source location
and time. As such the use of immutable objects and the 
borrowing of ideas from functional programming has become much more prevalent, 
such as in languages like Scala and Rust \cite{Rust}.

Immutability runs contrary to what objects were intended to be --- objects in
their most fundamental form can be viewed as entities with private state and 
associated protocols, which specify the messages the entities can respond to.
An immutable object, meaning an object with constant private state and
a stateless protocol, has much more in common with a type constructor in 
a functional language.

An object may have a very simple 
protocol, where the set of messages (or equivalently, method calls) it is 
willing to accept is constant throughout its lifecycle --- this is what the 
type system of programming languages such as Java can express. This is
insufficient to correctly model the object's behaviour in many cases, as
the set of messages the object can correctly respond to is dependent upon an
abstraction of its private state. State change is typically synchronous, in 
response to message receipt, and safety requires that 
any message sent to an object is within the set which can be received for
the currently known state.

Describing the allowed sequence of messages which a class of objects can
receive, their typestate, produces significantly more complex types than
in an object oriented language without typestate constraints. Functions which
take objects as parameters must describe the type of object they accept,
in which state, and what state transformation will occur (which may not
be deterministic). This type annotation burden is particularly pronounced
in a language with function literals, which must be succinct
in order to be readable.

Additionally, static checking of typestate requires
that the current state of an object always be known at a point of invocation.
This is only possible when an alias control strategy is employed, such as linear types, 
but such types prohibit programming in a style that most programmers are 
familiar with. Attempting to relax linear typing through the use of fractional 
permissions as in Plural and Plaid \cite{Aldrich2009,Bierhoff2009}
adds a significant amount of complexity to the type system. This results in 
large type annotations within programs to express both typestate and aliasing 
constraints when modular static type checking is desired.
If it were possible to eliminate the need for such type annotations, code
could be more concise and type annotations would only be necessary for
documentation purposes, and could be checked that they are simply more specific
versions of the principal typing produced by the type inference algorithm.

\section{Modelling typestate constraints}
\label{sec:tsmodel}

In previous work \cite{McGinniss2011}, we have taken the view that most use 
cases for typestate can be adequately modelled by a deterministic
finite state machine. The state machine explicitly describes the
states of the object, the methods which can be called in each state and
the transition triggered by a method call in a specific state.
Fields on the object can either be treated as private or as a special case
of method invocation \cite{Abadi1996}. For simplicity
we will only consider methods with no parameters, as there are very few
use cases where the values of parameters have an effect on state transition 
decisions which could be described without the integration of much more
complex features into the model, such as range types, conditional transitions
or a full logical framework.

The type of a given object at any point in time can be described as the
combination of a state machine (which we will refer to as the 
{\it object protocol}) and a particular state from within that protocol.
We write this as $O@S$, where $O$ is the protocol and $S$ is the state.
Formally, an object protocol $O$ is defined as a triple
$\langle M, \Sigma, \Delta \rangle$ where $M$ is a set of all
method names, $\Sigma$ is the set of state labels and
$\Delta \subseteq \Sigma \times M \times \Sigma$ is the state
transition relation. Transitions for a given state must be unique, such that 
$(S, m, S') \in \Delta \implies \nexists S'' \:\mathrm{st.}\: S' \neq S'' 
\wedge (S, m, S'') \in \Delta$.

The subtyping relation $<:$ is defined in terms of a
{\it simulation preorder}. Given $O = \langle M, \Sigma, \Delta \rangle$ and
$O_2 = \langle M_2, \Sigma_2, \Delta_2 \rangle$, the simulation preorder
$\sim$
is defined to be the largest relation such that if $O@S \sim O_2@S_2$ then for all
$m$ and $S_4$ such that $(S_2, m, S_4) \in \Delta_2$ there exists an $S_3$ such that
$(S, m, S_3) \in \Delta$ and that $O@S_3 \sim O_2@S_4$.
Subtyping is defined such that $O@S <: O_2@S_2 \iff O@S \sim O_2@S_2$.

A perhaps more intuitive basis for subtyping
uses the notion of {\it trace inclusion}. The set of interaction traces 
$\Tr(O@S)$ of a type is the smallest set such that:

\[
\begin{tabular}{cc}
\infer{\epsilon \in \Tr(O@S)}{}
&
\infer{m . t \in \Tr(O@S)}
{(S, m, S') \in \Delta & t \in \Tr(O@S')}
\end{tabular}
\]

where $\epsilon$ is the empty trace, and $m . t$ is the 
prefixing of an existing trace with a method call.
Subtyping is therefore defined such that
$O@S <: O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$.

It is easy to
show that $O@S \sim O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$, so these
interpretations of subtyping can be used interchangeably wherever necessary.

Either view of subtyping is also easy to relate to the Liskov substitution 
principle \cite{Liskov1987} --- a type will accept every sequence of
method calls that a supertype would, therefore it is a {\it behavioural}
subtype.

\section{TS - an imperative calculus with typestate}

TS is a small imperative calculus which models interactions with 
typestate-constrained objects, supports function literals without 
implicit capture, and employs structural subtyping for function and object
types.
The grammar of the language is shown 
in \figref{fig:grammar}, with operational semantics in \figref{fig:opsemantics} 
and typing rules in \figref{fig:typerules}.
The language is designed primarily
to study typestate inference in imperative code, rather than to 
be practical, as it uses syntactic control of interference \cite{Reynolds1978} 
to avoid the issue of alias control entirely - there is no means within the
language to produce an alias to an object.

The language has the following values:

\begin{itemize}
\item {\tt unit}, the sole value of the {\tt Unit} type, which is used as
the result type for terms which have no meaningful result.
\item Function literals, which do not allow for implicit capture
of variables into their scopes --- everything must be passed as a parameter.
Tracking effects on implicitly captured references is not very well
studied in the current literature outside of strict linear types where
function literals that capture linear references must themselves be linear.
Parameters are passed by reference. 
Requiring that all references be explicitly passed to the function allows for
effects to be explicitly declared. The notation $T \gg T'$ is used to
describe an {\it effect type}, which declares a required type for the parameter 
on the left, and the new type of the parameter after application of the function 
on the right. Not all effect types are legal, as will be discussed later.
\item Object literals. These are composed of a state machine description
and a current state. The state machine description contains a set of
distinct state labels $\overline{S_i}$, within which is a set of methods $\overline{m_{ij}}$ 
which can be called in that state. The language does
not attempt to fully model the behaviour of objects, only interactions with
them. As such, it is sufficient for each method to be defined as a pair 
consisting of the return value of the method, and the next state of the object 
if the method is invoked. Non-determinism in the production of a return value 
could be simulated by specifying a set of possible return values and allowing 
the operational semantics to select one of these randomly.
All values would have to have the same type, or types for which a least upper 
bound exists. Such non-determinism would have no effect on the type system,
which is the main focus of this paper.
\end{itemize}

\noindent
The language has the following terms:

\begin{itemize}
\item A conditional of form $\ift{t_c}{t_t}{t_f}$, which evaluates
$t_c$ first to a boolean result which decides which of $t_t$ 
(the \"true\" branch) or $t_f$ (the \"false\" branch) is
evaluated.
\item Function calls of form $x(\overline{x_i})$, which extract the function
literal stored in $x$ and apply it to the set of parameters $\overline{x_i}$,
which must match the declared parameter types. Each $x_i$ must be a distinct
variable, and is passed by reference, allowing the type of $x_i$ to be changed
as a result of the function application.
\item Method invocations on objects of form $x.m$. A method call is only
valid when the method is available in the current state of the object stored
in $x$. After a valid invocation, the state of the object will have changed
as specified by the object's state machine.
\item A let-bind construct $\lett{x}{t}{t'}$, which creates a new variable
$x$ within the scope of the body term $t'$ which is a reference to the value
derived from the term $t$. Rebinding a variable name is not allowed.
\item Sequencing of form $\tseq{t}{t'}$, which evaluates terms in 
left-to-right order.
\end{itemize}

While variables are references to a value, there is no construct which
can duplicate a reference. This very simple mechanism
ensures that references are unique, allowing us to focus our attention on the
typestate properties of objects without the added complication of alias
control annotations in the types. The restriction that each variable passed to
a function be unique is for this reason, to prevent implicit duplication
by passing the same variable more than once to the same function.

\begin{figure}

\ottgrammartabular{
\ottt\ottinterrule
\ottv\ottinterrule
\otto\ottinterrule
\ottT\ottinterrule
\ottO\ottinterrule
}

\caption{\label{fig:grammar} Grammar of TS0}
\end{figure}

The example in \figref{fig:exampleprog} shows a short TS0 program
that TODO.

\begin{figure}
\begin{lstlisting}
TODO
\end{lstlisting}
\caption{\label{fig:exampleprog} An example program in TS0}
\end{figure}

\subsection{Stores}

Each variable is a unique reference to a value in a store $\mu$, which is 
defined as $\mu\:\mathrm{::=}\:\overline{x_i \mapsto v_i}$ where each $x_i$ is
distinct. A store location
is created (or overwritten) upon entry to the body of a let-binding, written as
$\mu[x \mapsto v]$, which is defined as follows:

\[
\begin{tabular}{cc}
\infer
{\mu[x \mapsto v] = \mu, x \mapsto v}
{\mu = \overline{x_i \mapsto v_i} & x \notin \overline{x_i}}
&
\infer
{\mu[x \mapsto v] = \mu, x \mapsto v}
{\mu = \overline{x_i \mapsto v_i}, x \mapsto v'}
\end{tabular}
\]

As there is no way for a reference to be duplicated, the mapping
from variable names to store locations is injective and therefore we have
no need for explicit ``location values'' as in the typical treatment of 
languages with references (such as in \cite{Pierce2002}). There is no need
for a separate store typing, as a well-typed store can be defined directly
in relation to a context as in \figref{fig:storetype}. A well-typed store
may mappings for variables which are not in the context as shown in
rule \textsc{st\_extra}.

\begin{figure}

\ottdefnreduce

\caption{\label{fig:opsemantics} Operational semantics}
\end{figure}

\subsection{Type rules and soundness}

As the language is imperative and the types of variables pointing to objects can
change, it is necessary for the typing judgements to express changes to the 
environment. We must then be able
to prove that the changes in the environment correspond to the changes in the
store during evaluation. A context $\Gamma$ is a set of unique variable names 
mapped to types, defined as $\Gamma\:\mathrm{::=}\:\overline{x_i : T_i}$,
with each $x_i$ distinct. \figref{fig:storetype} defines what it means for
a store to be well-typed with respect to a
context. \figref{fig:typerules} defines the rules for well-typed terms
in the language, while \figref{fig:subtyperules} defines the subtyping
relation.

\begin{figure}
\begin{ottdefnblock}
{$\Gamma  \vdash  \mu$}
{\ottcom{store $\mu$ is well typed against context $\Gamma$}}
\end{ottdefnblock}

\[
\begin{tabular}{cc}
\infer[\textsc{st\_empty}]
{\emptyset \vdash \emptyset}
{}
&
\infer[\textsc{st\_vars}]
{\overline{x_i : T_i} \vdash \overline{x_i \mapsto v_i}}
{\typerule{\Gamma}{v_i}{T_i}{\Gamma'}}
\\\\
\multicolumn{2}{c}{
\infer[\textsc{st\_extra}]
{\Gamma \vdash \mu, \overline{y_j \mapsto v_j}}
{\Gamma \vdash \mu & \overline{y_j \notin dom(\mu)}}
}
\end{tabular}
\]

\caption{\label{fig:storetype} Store typing judgements}
\end{figure}

\begin{figure}
\ottdefnmethtype

\ottdefntype

\caption{\label{fig:typerules} Type rules}
\end{figure}

\begin{figure}

\ottdefnsubtype

\caption{\label{fig:subtyperules} Subtyping rules}
\end{figure}

Corresponding to the values of the language, there are four main kinds
in the type system: Unit, Bool, function types and object types. 
Two modifications are made to the definition of object protocols described
in \secref{sec:tsmodel}: methods have return types, and an object type
is the combination of an object protocol and a {\it set} of states.
An object protocol $O$ is therefore formally defined as the tuple 
$O = \langle M, \Sigma, \Delta \rangle$
where the transition relation $\Delta \subseteq \Sigma \times M \times T \times \Sigma$, 
where $T$ is the set of all types in the language.

While an object value in TS is still deterministic and has a single state at any 
given
point in time, we may not be able to determine in the type system which state 
exactly it is in --- conditional evaluation readily produces this situation.
With an if-then-else statement where the branches
interact with an object differently, it is common for the state of the object
at the end of each branch to be different. As
knowing precisely which branch will be taken is rarely possible in a
decidable type system, we must either insist that the state of the object
be the same upon exit from both branches, or devise a means to cope with
the object being in one of multiple potential states. We choose the latter
approach, as it is both more flexible and allows for intuitive principal
types to be chosen, as will be discussed later.

% abandoned attempt to define join/meet
% \begin{figure}

% \[
% \begin{array}{c}
% \infer[\sqcap-state]
%   {S \sqcup S_2 = \{ \overline{ m : T'' \Rightarrow (S_3 \sqcap S_4)} \} }
%   {
%   m : T \Rightarrow S_3 \in O@S
%   &
%   m : T' \Rightarrow S_4 \in O_2@S_2
%   &
%   T'' = T \sqcup T'
%   }
% \\
% \infer[\sqcup-omg]
%   {S \sqcup S_2 = 
%   \{ \overline{ m : T'' \Rightarrow (S_3 \sqcup S_4} \} } 
%   \cup
%   \{ \overline{ m' : T_3 \Rightarrow S_4 \} }
%   {}
% \]

% \caption{\label{fig:joinmeet} Join/meet definition}
% \end{figure}

An object type in TS, written $O@\overline{S}$, is equivalent to
$\bigsqcup_{S \in \overline{S}} O@S$, where $T \sqcup T'$ is the least upper
bound (or {\it join}) of $T$ and $T'$. Deriving a least upper bound for
two arbitrary object types (with a single current state) 
involves constructing the intersection of the state machines for each object 
type, using the current state of each object type as the initial state.
Similarly, deriving a greatest lower bound is based upon the union of
the two state machines in the same fashion. Join is always defined for
object types, while meet is not --- where the return types of methods are
incompatible (i.e. no join/meet exists respectively), a join operation will 
discard the method (it is unsafe to
call it) while a meet operation would require that the method be callable,
however no valid return type can be assigned to the method. Therefore
the kind of object types is a join semi-lattice, with the object type 
$\{ S_1 \{\} \}@S_1$ serving as the least element.

The intuition here is that the only sequences of method calls which can be 
safely invoked on an object of type $O@\overline{S}$ are those which could be 
safely invoked on each individual $O@S$, where $S \in \overline{S}$, and that
a least upper bound of the return types exists.

In general, the set of
types in TS is a disjoint union of join semi-lattices.

\subsection{Extra notational conventions}

Given the formal definition of an object type $O@\overline{S}$, to save
referring to the $\Sigma$, $M$ and $\Delta$ that $O$ is composed of we shall
use the following notational conventions:

\begin{itemize}
\item $S \in O$, a predicate asserting $S$ is a state of $O$, is equivalent to
$S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$, the method existence predicate, 
states the method $m$ with return type $T$ exists in state $S$ and triggers a 
transition to state $S'$ when called. Formally, $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the previous to state sets, meaning that for all $S \in \overline{S}$
there exists a state $S'$ and type $T'$ where 
$(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \sim O_2@\overline{S_2}$, the extension of the
simulation relation to support state sets, can be interpreted as
as $\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

\subsection{Effect types}

An ``effect type'' $T \gg T'$ describes how the type of a variable changes
in response to function application. In the TS language, the only type
changes that are those which change the state of an object instance; the
values of the other types are immutable. If a boolean is passed as a parameter,
the effect type will always be $\boolt \gg \boolt$, similarly for
$\unitt$ and $\funt{T}{U}{V}$. An effect type of $\boolt \gg \unitt$ is not
possible. In general, effect types are kind-preserving.

\begin{figure}
\includegraphics[width=0.49\textwidth]{sm1.pdf}
\caption{\label{fig:sm1} Operational semantics}
\end{figure}

In the case of an object parameter, not all effect types are possible
either. Consider
the object protocol in \figref{fig:sm1}, which we will refer to as $O_1$.
An effect type of $O_1@B \gg O_1@A$ is impossible for this protocol, as there
is no sequence of method calls which would result in a transition from state
$B$ to state $A$. An effect type of $O_1@B \gg O_1@\{B,D\}$ is possible, 
as there is at least one sequence of method calls on an object in
state $B$ to both state $B$ and $D$.
An effect type $O_1@{B,D} \gg O_1@{B,C}$ which has a state set on
the left hand side is possible as there exists a sequence of method calls 
that would take an object in state $B$ to $C$ that would also
take an object in state $D$ to $B$ (specifically, a single method
call to $b$).

In the general case, an effect type concerning an object protocol $O$
of form $O@\overline{S} \gg O@\overline{S_2}$ is
legal if:

\begin{itemize}
\item Given a sequence of method calls that takes an object in state $S \in \overline{S}$
to state $S_2 \in \overline{S_2}$, the same sequence must also take an object
in any other $S' \in \overline{S}$ to a state $S_2' \in \overline{S_2}$.
\item All $S_2 \in \overline{S_2}$ are reachable by at least one sequence
of method calls from an object in a state $S \in \overline{S}$.
\end{itemize}

Checking the legality of an effect type is straightforward: by computing the
intersection of all the states, we can associate with each state in the 
intersection a set of states in the original machine
it was derived from (as shown in \figref{fig:sm1}. 
Should an associated state set contain an end state,
if all other states in the set are also end states then this represents a
``candidate'' end state within the intersection machine. The union of the state 
sets associated with all candidate end states must be equal to the
desired end state set for the effect type to be legal. This would correctly rule 
out the effect type $O@\{B,D\} \gg O@\{B\}$ (there is no sequence of method
calls that would take an instance in either state $B$ or $D$ and result in
the object exclusively being in state $B$) while accepting effect types
$O@\{B,D\} \gg O@\{B,D\}$ (corresponding to no operations occurring) and
$O@\{B,D\} \gg O@\{B,C,D\}$ (corresponding to either no operations occuring
or a single call to $b$).

\subsection{Principal types}

A principal type is one which makes the smallest commitment about 
the structure of a value based on its usage; due to the presence of subtyping 
in TS, a principal type is an upper bound on the type of the value.

\begin{figure}
\includegraphics[width=0.49\textwidth]{principal.pdf}
\caption{\label{fig:principalex}}
\end{figure}

Following from this general definition, a principal object type in TS 
is an object protocol and initial state set which permits all sequences of 
method calls that could be generated by an associated term $t$, and no more. 
In other words,
the language accepted by the state machine is the smallest possible, while
the return types of methods are as general as possible. Consider the example
function $f$ shown in \figref{lst:principalex}. It is clear from the usage
of $y$ that it must be of type $\boolt$ as demanded by the type judgement
of an if-then-else statement, while $x$ is an object of some sort.
We could ascribe any of the object types $O_1@A$, $O_2@A$ and $O_3@A$ to the parameter 
$x$, and after evaluation the object types would be $O_1@A$, $O_2@\{E,H\}$
and $O_2@F$ respectively.
$O_1@A$ is not principal as it will admit many more sequences of method calls 
than strictly required by the body of function $f$, while $O_2@A$ and $O_3@A$ 
are principal.

$O_2$ contains more states than are strictly necessary, i.e. it
is not a {\tt minimal} state machine, while $O_3$ is minimal.
We do not require that a principal type be minimal. If necessary, any
principal state machine we produce can be minimised, so this can be done
on demand where there is a specific need for it. One such need may be when
presenting the type to a user, as a minimal state machine may be easier to 
understand. Regardless, the state machines $O_2$ and $O_3$ can be shown to be 
isomorphic in the language that they accept (starting at state $A$), and that 
language is minimal, so we assert that either are acceptable principal types.

Similarly, it is possible for a principal effect type to have more than
one state in the state set of the input (left) object type. It is
always possible to restructure such an object type to have a single initial
state. It is not, however, possible to have both a single input state and
single output state for a state machine accepting an arbitrary language --- 
consider the language accepted by the regular expression $m^* | n^*$. It is 
necessary for a state machine which accepts this language, and only this 
language, to
have at least three exit states. In general, it is more intuitive to have
a single input state in the effect type but this is not required for a
type to be considered principal.

A principal type we can therefore ascribe to the function $f$
is $\forall T_1 \ldots T_6 . (O_3@A \gg O_3@F, \boolt) \rightarrow T_6$, 
or equivalently 
$\forall T_1 \ldots T_6 . (O_2@A \gg O_2@\{E,H\}, \boolt) \rightarrow T_6$.

There are occasional advantages to not always minimising principal state
machines. The additional structure of $O_2$ retains additional information
in the effect type which can be helpful when interpreting an effect type
when a subtype is passed to the function, as will be described in the next
section.

\subsection{Interpreting effect types}

The typing judgement for a function call permits a passed parameter to
be a subtype of the input portion of the associated effect type.
Determining a precise upper bound for the type of the parameter after the
function call, in particular where the parameter is an object, 
requires some thought.

Consider a function call $f(x)$ where
$f : (O_2@\overline{S_2} \gg O@\overline{S_4}) \rightarrow T$ and
$x : O_1@\overline{S} <: O_2@\overline{S_2}$, where $O_1$ and $O_2$
are the state machines from \figref{fig:principalex}, 
$\overline{S} = \{ A \}$, $S_2 = \{ A \}$ and $S_4 = \{E, H\}$. 
Based on the effect type,
we can assert that the function will have interacted with the $x$ using
one of the possible sequences that would be accepted by $O_2@\overline{S_2}$,
ending with the object in the state set $\overline{S_4}$.

A straightforward but questionable choice would 
be to replace the type of the variable
with that of the output portion of the effect type after the call. However,
this will often result in an unacceptable loss of type information,
reducing the number of valid programs which can be typed within the system.

It is clear that as $\ot{O}{S} \sim \ot{O_2}{S_2}$ there must exist some
$\overline{S_3}$ such that $\ot{O}{S_3} \sim \ot{O_2}{S_4}$.
One possible solution would be to choose
$\overline{S_3} = \{ S | O@S \sim \ot{O_2}{S_4} \}$. This is not specific
enough however --- given the object protocols from \figref{fig:principalex},
$\overline{S_3} = \{ A, B, C, D \}$ as there are no possible method calls
on states $E$ or $H$ of $O_2$.

It is clear that our choice of $S_3$ must relate the set of possible method
call sequences between $\ot{O_2}{S_2}$ and $\ot{O_2}{S_4}$. For this we
define a relation between pairs of object types
$(O@\overline{S}, O_2@\overline{S_2} \lesssim 
(O@\overline{S'}, O_2@\overline{S_2'})$, which we call the
{\it simulation predecessor relation}. The definition of this relation
is given in \figref{fig:simreach}. Essentially, if $O@S \sim O_2@S_2$ and there
exists a sequence of method calls that would transition an object value of type
$O_2@\overline{S_2}$ to $O_2@\overline{S_4}$, then this sequence of method calls
would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
such that $O@\overline{S_3} \sim O@\overline{S_4}$. This is abbreviated to
$(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.

Using this relation, we can choose $\overline{S_3} = 
\bigcup \{ \overline{S'} | 
  (\ot{O}{S} , \ot{O_2}{S_2}) \lesssim (\ot{O}{S'}, \ot{O_2}{S_4})
\}$. The solution using the object protocols from \figref{fig:principalex}
will therefore be $\overline{S_3} = \{ A \}$.

\begin{figure}

\[
\begin{array}{c}
\\
\infer[\lesssim\textsc{-adj}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})}
  {
    \begin{array}{cc}
    O@\overline{S} \sim O_2@\overline{S_2}
    \\
    m : T_i \Rightarrow \overline{S_4} \in O@\overline{S_2}
    \\
    m : T_i' \Rightarrow \overline{S_3} \in O@\overline{S}
    \end{array}
  }
\\\\
\infer[\lesssim\textsc{-refl}]
  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S}, O_2@\overline{S_2})}
  {O@\overline{S} \sim O_2@\overline{S_2}}
\\\\
\infer[\lesssim\textsc{-trans}]
  {
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  }
  {
  \begin{array}{c}
  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})
  \\
  (O@\overline{S_3}, O_2@\overline{S_4}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
  \end{array}
  }
\end{array}
\]
\caption{\label{fig:simreach} Definition of the simulation predecessor relation}
\end{figure}

The type system of TS uses a partial function named {\it remap}, defined in
\figref{fig:remaprules} to formalise
this selection of the type of a parameter after a function call. The cases
for Bool, Unit and function types are trivial as these types are immutable,
while for objects the simulation predecessor relation is used.

\begin{figure}
\ottdefnremap

\caption{\label{fig:remaprules} Definition of remap}
\end{figure}

\subsection{Soundness}

The soundness of the type system is stated formally in the following
theorem:

\begin{thm}[Progress and Preservation]
Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
there exist $t'$ and $\mu'$ such that
$\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
\end{thm}

\begin{proof}
in appendix \appref{app:soundness}.
\end{proof}

\section{Type inference for TS0}

Throughout much of the preceding text we have avoided writing object
types using the full syntax, instead preferring to use diagrams. The full
syntax for object types is verbose and difficult for a user to parse.
Having to write out object types in full would therefore obfuscate the
intent of the code, particularly when there is such intuitively a
strong correlation between the code and the principal object types of the
variables it interacts with. As such, the ability to infer principal object
types is essential for any language which includes typestate, especially
where structural rather than nominal subtyping is employed.

The main source of type verbosity in TS0 is the requirement for
type annotations on function literals. 
Despite there being no need for alias control
annotations, the effect declarations for each parameter are still verbose
and the ability to define type aliases would only help partially.

We therefore desire a general type inference strategy that can infer a
minimal $\Gamma$, $\Gamma'$ and $\alpha$ for a term $t$ such that 
$\typerule{\Gamma}{t}{\alpha}{\Gamma'}$. If $t$ is the body of a function
literal, we can extract the effect types for the named parameters (of which
the free variables of $t$ should be a subset) from $\Gamma$ and $\Gamma'$.

Our approach to type inference is inspired by \cite{Pottier2005} which advocates
separate constraint generation and constraint solving phases for
type inference. Our constraint grammar is shown in \figref{fig:congrammar}.

\begin{figure}
\ottgrammartabular{
\ottC\ottinterrule
\ottctxc\ottinterrule
\otttyc\ottinterrule
\otttinf\ottinterrule
}
\caption{\label{fig:congrammar} Constraint grammar}
\end{figure}

\begin{figure}
\[
\begin{array}{l}

% unit value
\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma = \gamma' 
\\\\

% true value
\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
\\\\

% false value
\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
\\\\

% object value
\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@S}
\triangleq
\exists \overline{\alpha_{ij}} . \\
\alpha = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\}@S
\cand
\gamma = \gamma'
\cand \\
\overline{\inferrule{\gamma}{\triangleright v_{ij}}{\alpha_{ij}}{\gamma}}
\\\\

% function value
\inferlhs{\lambda ( \overline{x_i} ) . t} \triangleq
\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_{ai}, \alpha_{bi}}, \alpha_r . \\
\alpha = (\overline{\alpha_{ai} \gg \alpha_{bi}}) \rightarrow \alpha_r
\cand
\gamma = \gamma'
\cand
\gamma_r = \emptyset
\cand \\
\gamma_{in} = \gamma_r, \overline{x_i : \alpha_{ai}}
\cand
\overline{x_i : \alpha_{bi} \in \gamma_{out}}
\cand
\\
\inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
\\\\

% let binding
\inferlhs{\lett{x}{t'}{t}} \triangleq \exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
\gamma_{b1} = \gamma_v, x : \alpha'
\cand
\gamma' = \gamma_{b2} - \{ x \}
\cand \\
\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}
\cand
\inferrule{\gamma_{b1}}{t}{\alpha}{\gamma_{b2}}
\\\\

% sequence
\inferlhs{t ; t'} \triangleq
\exists \gamma'', \alpha' . \\
\inferrule{\gamma}{t}{\alpha'}{\gamma''}
\cand
\inferrule{\gamma''}{t'}{\alpha}{\gamma'}
\\\\

% function call
\inferlhs{x ( \overline{x_i} )} \triangleq
\exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
\overline{x_i : \alpha_i \in \gamma}
\cand 
x : ( \overline{\alpha_i' \gg \alpha_i''} ) \rightarrow \alpha \in \gamma
\cand
\\
\gamma' = \gamma [ \overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''} } ]
\\\\

% method call
\inferlhs{x.m} \triangleq
\exists \alpha_x . \\
x : \alpha_x \in \gamma
\cand 
\cand  \\
\gamma' = \gamma[ x \mapsto \remapapp{\alpha_x}{O@S}{O@S'} ]
\\
\mathrm{where } O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
\\\\

% if-then-else
\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq
\\
\exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}
\cand
\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}
\cand 
\\
\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}
\cand
\alpha_c = \boolt
\cand
\\
\alpha = \alpha_t \sqcup \alpha_f
\cand
\gamma' = \gamma_t \sqcup \gamma_f
\\\\

\end{array}
\]
\caption{\label{fig:constraintgen} Constraint generation}
\end{figure}

There are three categories of constraints:

\begin{itemize}
\item {\bf Context constraints} --- these define relationships between contexts.
It is important that type changes be correctly observed by the inference
approach and that context constraints expressed in terms of equality, extension,
modification and removal exactly mirror the typing judgements. 
Extension, replacement and removal constraints modify the solution of
another context variable. Following these context variable dependencies,
ultimately every context variable depends upon a {\it root} variable which
is either of form $\gamma = \emptyset$ or $\free{\gamma}$. The root
context is {\it free} when it is the input context to the top-level term being
processed, while
the root is {\it empty} when it is the input context for a function body term.
A free context can be arbitrarily extended with new variables, while an
empty context cannot --- this difference allows the detection of unbound
variable usage within function bodies and the collection of free variables
in the root term.

Additionally, we have {\bf variable usage constraints} of form 
$x : \mathbb{T} \in \gamma$, which are generated for each location in the
program where a variable is used and what type that usage implies. These
are used to generate additional type variable
constraints by matching the asserted type $\mathbb{T}$ with what is known
about the type of $x$ in $\gamma$.

\item {\bf Type expressions constraints} --- these express 
constraints between two or three types: either two types are equal,
or one type is the join of two others. The latter form of type constraint
is generated when dealing with if-then-else terms.
\end{itemize}

Type variables can appear at any position within a type's structure, as shown
in the grammar for $\mathbb{T}$. Objects are treated as a special case, where
we use variables as placeholders for the object definition and state parts.
The states and transitions of the object structure are determined at a later
stage in the type inference process.

A constraint generation expression $\inferlhs{t}$, which relates the term
to an input context variable, a type variable and an output context variable, 
is expanded using the rules shown in \figref{fig:constraintgen}.
The constraint generation process for a root term $t$ begins with
$\exists \gamma, \gamma', \alpha. \free{\gamma} \cand \inferlhs{t}$.

The constraint generation rules correspond to the typing judgements for 
well-typed terms as one would expect:

\begin{itemize}
\item All value terms produce a $\gamma = \gamma'$ constraint as values do
not change the context. Additionally, a type variable constraint of form
$\alpha = \mathbb{T}$ where $\mathbb{T}$ is of the appropriate type to match
the value, with type variables in all positions relating to parameters and return
values.
\item The context variable $\gamma_a$ used as the input context for
function bodies has an empty root $\gamma_r$, to allow for detection of 
used unbound variables within
the body. This matches the use of a context in type inference
composed only of the parameters, and prevents implicit binding.
\item Let-bind terms generates a new context variable $\gamma_v$ for the
output context of the value term. This is extended by $\gamma_{b1}$, the
input context for the body term, with variable $x$ of the type inferred
for the value. The output context of the whole let-bind term removes
$x$ from the output context $\gamma_{b2}$ of the body term.
\item Update terms generate replacement constraints based on the inferred
type $\alpha'$ of the right-hand side of the expression.
\item Sequencing terms relate the output of the left term to the input of the
right term through a new context variable $\gamma''$. The type variable
$\alpha'$ must be generated to capture the type of the left term for
consistency, but it is not needed.
\item Function call terms generate context variable constraints to relate the
usage of the variables to whatever information has been inferred about the
parameter types of the function. All variables $\overline{x_i}$ used as 
parameters in the function call are replaced in the output context with
new type bindings to $\overline{\alpha_i'}$, which represents the effect
of the function.
\item Method call terms generate type variables to capture that $x$ must
be an object, that the method $m$ must exist in the current state $\alpha_s$
and that after the method call $x$ must be in the new state $\alpha_s'$.
\end{itemize}

\subsection{Context constraint solving}

Constraint solving is performed in phases. In the first phase, the context 
constraints
are normalised using the rules in \figref{fig:contextsolve} --- where the set of 
constraints in the premise exists, they are replaced with those in the 
conclusion. Normalisation means that every context variable will have a single 
defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
$\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$, where $\gamma'$ is
either empty or free. A context $\Gamma$ for type checking can be easily
constructed from constraints in this form, as once the type variable
constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
every $\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$ where $T_i$ is the
solved form of $\mathbb{T}_i$.

The context normalisation rules do the following:

\begin{itemize}
\item \textsc{ceq} --- Where $\gamma$ is directly equivalent
to $\gamma'$, which is defined in terms of extension of
$\gamma''$, rewrite the constraints such that
$\gamma$ directly copies the extension of $\gamma''$.
\item \textsc{crem} --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
which is defined in terms of extension of $\gamma''$, rewrite the
constraints such that $\gamma$ copies the extension of $\gamma''$ with
the variable $x$ removed from the extension. There is no need for an
error-producing variant of \textsc{crem}, where the variable $x$ is missing, 
as context removal constraints are always generated in a pair with a context 
extension constraint for let binding.
\item \textsc{cext} --- Where $\gamma$ extends $\gamma'$ with
a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
constraints such that $\gamma$'s extension set includes both the $y_j$
definitions and $x_j$ definitions.
\item $\textsc{cext}_2$ --- Where $\gamma$ extends $\gamma'$ with
a new variable $x$, and $\gamma'$ is defined in terms of extension of
some $\gamma''$ and the extension set already contains $x$, this
represents an attempt to redeclare $x$. As such, generate an error
$\mathbf{bound}\:x$, terminating the constraint solving process.
\item \textsc{crep} --- where $\gamma$ replaces the type of some
variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
such that $\gamma$ copies the extension set of $\gamma'$ with $x$
replaced as specified.
\item $\textsc{crep}_2$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of extension of $\gamma''$
which is free and $x$ is not in the extension set, $x$ is a
free variable in the term being processed. To introduce $x$ in such a way that
it will be included in all contexts, $\gamma''$ is redefined in
terms of extension of a new context $\gamma_r$ with variable $x$
in the extension set mapped to a new type variable $\alpha$. Following
this, \textsc{cext} can be applied to $\gamma'$, then \textsc{crep} can
be applied to $\gamma$.
\item $\textsc{crep}_3$ --- where $\gamma$ replaces variable $x$ in
$\gamma'$ which is defined in terms of an empty root $\gamma''$,
and $x$ is not in $\gamma$'s extension set, this indicates the usage
of an unbound variable within a function body. As such, generate an
error $\mathbf{unbound}\:x$ indicating that x is unbound within this
context, terminating the contraint solving process.
\item \textsc{cvar} --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of
extension of $\gamma_r$ which is free, and $x$ is not in the extension
set of $\gamma$, $x$ is a free variable in the term being processed. 
Similar to $\textsc{crep}_2$, redefine $\gamma_r$
in terms of extension of a new root $\gamma_r'$, with $x$ in
the extension set with the observed type $\mathbb{T}$.
\item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
for context variable $\gamma$, which is defined in terms of $\gamma'$
which is empty, and $x$ is not in the extension set of $\gamma$, 
$x$ is an unbound variable within a function body.
\end{itemize}

The rules are repeatedly applied to any matching sets of constraints
until no further rewriting is possible, producing either an error
or a set of context constraints which are fully normalised, and
can be converted into contexts for type checking easily.

The way in which the context constraints are generated guarantees
that the dependencies between contexts form a directed acyclic graph. For
this reason, we believe that that the first phase will always terminate,
but this has not yet been proven.

\subsection{Variable usage \& type variable constraint solving}

In the second phase, additional type variable 
constraints are generated by following the rule in \figref{fig:tvargen}, which
compares the observed type from the usage of a variable to the type
known in the context. Repeated application of this rule until no more matches
are found produces a constraint set composed exclusively of normalised
context variable constraints, type variable constraints and protocol
constraints. A first order unification algorithm can then be used to solve
the system of type variable constraints, producing a most general unifier
for the type variables if unification is possible.

First order unification will identify a set of type variables (and their synonyms)
which represent distinct object types. For each distinct object type, there
will also be a set of type variables which represent distinct states for
the object type. By matching these variables to the protocol constraints
within the constraint set, we can generate the state machine for each
object type in a straightforward manner. To produce a minimal description of 
the object protocol, this state machine can be minimised by constructing a 
bisimulation relation between the states within each object and collapsing 
equivalent states.

The typing rules of TS0 do not permit polymorphism, so
if the most general unifier does not map all type variables for
variables that occur in the input and output contexts of the term being
processed, then the term has no valid type in the language. Otherwise,
substituting the fully explicit type inferred for each type variable that
occurs in the inferred input and output context variables, as well as the
type variable for the input term, should produce a valid typing judgement
where the domain of the input context is the set of free variables in
the term. This has not yet been proven.

\begin{figure}

\begin{center}
\begin{tabular}{c}
	\infer[\textsc{ceq}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
\\\\
	\infer[\textsc{crem}]
	{
	\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
	}
	{
	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
	}
\\\\
	\infer[\textsc{cext}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
	{
		\gamma = \gamma', \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
		& \overline{y_j} \cap \overline{x_i} = \emptyset
	}
\\\\
	\infer[\textsc{cext\_2}]
	{\mathbf{bound}\:x}
	{
		\gamma = \gamma', \overline{y_j : \mathbb{T}_j}, x : \mathbb{T} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
		& x \in \overline{x_i}
	}
\\\\
	\infer[\textsc{crep}]
	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T} \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
	{\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
\\\\
	\infer[\textsc{crep\_2}]
	{
	\begin{array}{c}
	\exists \alpha, \gamma_r . 
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\\
	\gamma'' = \gamma_r, x : \alpha \cand
	\free{\gamma_r}
	\end{array}
	}
	{
	\begin{array}{c}
	\begin{array}{ccc}
	x \notin \overline{x_i}
	&
	\fresh{\gamma_r}
	&
	\fresh{\alpha}
	\end{array}
	\\
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\free{\gamma''}
	\end{array}
	}
\\\\
	\infer[\textsc{crep\_3}]
	{\mathbf{unbound}\:x}
	{
	\begin{array}{c}
	x \notin \overline{x_i}
	\\
	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
	\gamma'' = \emptyset
	\end{array}
	}
\\\\
	\infer[\textsc{cvar}]
	{\exists \gamma_r' . 
	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand
	\gamma_r = \gamma_r', x : \mathbb{T} \cand
	\free{\gamma_r'}
	}
	{
	\begin{array}{c}
	\begin{array}{cc}
	x \notin \overline{x_i}
	&
	\fresh{\gamma_r'}
	\end{array}
	\\
	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand 
	x : \mathbb{T} \in \gamma \cand
	\free{\gamma_r}
	\end{array}
	}
\\\\
	\infer[\textsc{cvar\_2}]
	{\mathbf{unbound}\:x}
	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \emptyset
	&
	x \notin \overline{x_i}}
\end{tabular}
\end{center}
\caption{\label{fig:contextsolve} Context normalisation}
\end{figure}

\begin{figure}
\[
	\infer[\textsc{tvar}]
	{\mathbb{T} = \mathbb{T}' \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
\]
\caption{\label{fig:tvargen} Refining type variable constraints}
\end{figure}

\subsection{Implementation}

A Scala implementation of a type checker for TS0 which uses the type inference
technique described can be found at \url{http://github.com/iainmcgin/ts0-lang}.
A topological order is built from the dependency graph between context variables
to efficiently normalise the context constraints. The first order unification 
algorithm presented in \cite{Martelli1982} is used to solve the type variable 
constraints, allowing the effects for the parameters to be extracted from
the context variables.

\section{Conclusion}

TS0 is a small, simple language that has proven useful
for the study of the inference of object protocols, and the theoretical work
done to date holds promise for future expansion and interesting results. 
To our knowledge, this is also the first attempt to study the inference
of generalised typestate in an object oriented context, rather than
more specific inference problems such as session type inference
\cite{Collingbourne2010}. The
constraint generation approach closely aligns with the typing judgements,
which we believe makes the approach intuitive.

\appendix

\section{Soundness}

\begin{thm}[Progress and Preservation]
Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
there exist $t'$ and $\mu'$ such that
$\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
\end{thm}

\begin{proof}
by induction on the structure of $t$. Assume $t$ is not a value. It is
therefore of one of the following forms:

\begin{itemize}
\item $t = \lett{x}{t''}{t'}$. 
It follows by rule \tlet that there exist $\Gamma_1, T', T''$
such that
$\typerule{\Gamma}{t''}{T'}{\Gamma_1}$ and that
$\typerule{\Gamma_1, x : T'}{t'}{T}{\Gamma', x : T''}$. There are two
possibilities:

  \begin{itemize}
  \item If $t''$ is some value $v$, then by \lemref{lem:valuectx} 
  we have that $\Gamma = \Gamma_1$ and reduction can occur
  by \rletv, such that 
  $\oprule{\lett{x}{v}{t'}}{\mu}{t'}{\mu [ x \mapsto v ]}$.
  Let $\mu' = \mu [ x \mapsto v ]$. Let $\Gamma'' = \Gamma, x : T'$.
  It follows that that $\Gamma'' \vdash \mu'$, with
  $\typerule{\Gamma''}{t'}{T}{\Gamma', x : T'}$.
  The requirements of rule \tdrop are satisfied by this, therefore we can 
  type $t'$ such that $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.

  \item If $t''$ is a term, 
  by induction we have that $\oprule{t''}{\mu}{t'''}{\mu'}$, with
  some $\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'''}{T'}{\Gamma_1}$.
  This satisfies the requirements of \rlett, meaning $t$ itself can reduce
  such that $\oprule{\lett{x}{t''}{t'}}{\mu}{\lett{x}{t'''}{t'}}{\mu'}$.
  Additionally, the conditions of \tlet are satisfied such that
  $\typerule{\Gamma''}{\lett{x}{t'''}{t'}}{T}{\Gamma'}$.

  \end{itemize}

\item $t = t_l ; t_r$. It follows by rule
\tseq that $\typerule{\Gamma}{t_l}{T_l}{\Gamma_{mid}}$ and
$\typerule{\Gamma_{mid}}{t_r}{T}{\Gamma'}$.

  \begin{itemize}
  \item If $t_l$ is some value $v$, then
  reduction can occur by \\ 
  \rseqlv such that
  $t_l; t_r \mid \mu \longrightarrow t_r \mu$. By \lemref{lem:valuectx}, 
  $\Gamma'' = \Gamma = \Gamma_{mid}$
  and $\mu' = \mu$, therefore $\Gamma'' \vdash \mu'$ and 
  $\typerule{\Gamma''}{t_r}{T}{\Gamma'}$.

  \item If $t_l$ is a term, by induction $t_l \mid \mu \longrightarrow t_l' \mid \mu'$
  with some $\Gamma'' \vdash \mu'$ and
  $\typerule{\Gamma''}{t_l'}{T_l}{\Gamma_{mid}}$.
  This satisfies the requirements of \rseqlt
  such that $t_l; t_r \mid \mu \longrightarrow t_l'; t_r \mid \mu'$.
  Additionally, the requirements of \tseq are satisfied such that
  $\typerule{\Gamma''}{t_l'; t_r}{T}{\Gamma'}$.

  \end{itemize}

\item $t = x ( \overline{x_i} )$. It follows by rule \tfunc that
$\Gamma = \Gamma_1, \overline{x_i : T_i}$ 
with $\Gamma_1(x) = \funt{U_i}{V_i}{T}$
such that $\overline{T_i <: U_i}$ 
and that $\Gamma' = \Gamma_1, \overline{x_i : T_i'}$
where $T_i' = \mathbf{remap} ( T_i, U_i >> V_i )$.

As $\Gamma \vdash \mu$, 
it follows that $\mu(x) = \funv{y_i}{U_i'}{V_i'}{t'}$
where $\typerule{\overline{y_i : U_i'}}{t'}{T'}{\overline{y_i : V_i'}}$,
with $\funt{U_i'}{V_i'}{T'} <: \funt{U_i}{V_i}{T}$. It
follows by \subfn that for each $i$, $U_i <: U_i'$, $V_i' <: V_i$
and $T' <: T$.
Additionally for each $x_i$, $\mu(x_i) = v_i$ such that 
$\typerule{\emptyset}{v_i}{T_i''}{\emptyset}$ 
where $T_i'' <: T_i$.

Let $t_{sub} = \subst{x_i}{y_i}{t'}$.
The term $t$ can be reduced by \rfunc such that
$x ( \overline{x_i} ) \mid \mu \longrightarrow t_{sub} \mid \mu$.
By the substitution lemma, 
$\typerule{\overline{x_i : U_i'}}{t_{sub}}{T'}{\overline{x_i : V_i'}}$.
By lemma \lemref{lem:upgrading}, it follows that
for each $i$, there exists a $V_i'' = \remapapp{T_i''}{U_i'}{V_i'}$
such that 
$\typerule{\overline{x_i : U_i}}{T_{sub}}{T'}{\overline{x_i : V_i''}}$.
By \lemref{lem:remap} for each $i$ that $V_i'' <: V_i'$.

Therefore, by given that $T_i <: U_i$ and that $V_i'' <: V_i' <: V_i$
and that $T' <: T$ we can apply \tnarrow, \twiden and \tsub
such that $\typerule{\overline{x_i : T_i}}{t'}{T}{\overline{x_i : V_i}}$.

Let $\mu' = \mu$. Let $\Gamma'' = \Gamma$. It follows that 
$\Gamma'' \vdash \mu'$. By applying the weakening lemma, we can type 
$t_{sub}$ such that $\typerule{\Gamma''}{t''}{T}{\Gamma'}$.

\item $t = x.m$. It follows by rule \tmethc that
$\Gamma = \Gamma_1, x : O@\overline{S}$ and that 
$\Gamma' = \Gamma_1, x : O@\overline{S'}$, with
$\overline{S'} = \{ S_n | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \}$.
As
$\Gamma \vdash \mu$, we must have $\mu = \mu_1, x \mapsto o'@S_c$ with
$\Gamma_1 \vdash \mu_1$ and where
$\typerule{\emptyset}{o'@S_c}{O'@S_c}{\emptyset}$ such that
$O'@S_c <: O@\overline{S}$.

By the definition of subtyping for object types,
we must have that
$o' = [ S' \{ m = (v,S_n) ; ... \} ... ]$, where $O'@S_n <: O@\overline{S'}$. 
Where the value $v$ is concerned, $\typerule{\emptyset}{v}{T'}{\emptyset}$
for some $T'$ such that $T' <: T = \bigsqcup{ \{ T_r | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \} }$.

Therefore reduction can occur by \rmethc such that
$x.m \mid \mu_1, x \mapsto o@S \longrightarrow v \mid \mu_1, x \mapsto o@S_n$.
Let $\mu' = \mu_1, x \mapsto o@S'$. Let $\Gamma'' = \Gamma'$. It follows that
$\Gamma'' \vdash \mu'$, as $O'@S_n <: O@\overline{S'}$. 
By application of the weakening lemma and rule \tsub, it follows that
$\typerule{\Gamma''}{v}{T}{\Gamma'}$.

\item $t = \ift{t_c}{t_t}{t_f}$. It follows by rule \tif that there exists
$\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$ and $T_f$ such that
$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
$\typerule{\Gamma_2}{t_f}{T_f}{\Gamma_3}$ with
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$. There are
three possibilities for reduction:

  \begin{itemize}
  \item $t_c = \truev$. By \lemref{lem:valuectx} we have that
  $\Gamma = \Gamma_1$. Reduction can occur by \riftrue such that
  $\oprule{t}{\mu}{t_t}{\mu}$. By \tsub we have that
  $\typerule{\Gamma_1}{t_t}{T}{\Gamma_2}$ as $T_t <: T$. Additionally, for
  all $x \in \Gamma_2$, $\Gamma_2(x) <: \Gamma'(x)$. Therefore by
  repeatedly applying \twiden for all $x$, we have that
  $\typerule{\Gamma_1}{t_t}{T}{\Gamma'}$.
  Let $\Gamma'' = \Gamma$, $\mu' = \mu$ and 
  $t' = t_t$. Therefore, $\typerule{\Gamma''}{t'}{T}{\Gamma'}$ with
  $\Gamma'' \vdash \mu'$.

  \item $t_c = \falsev$. By \lemref{lem:valuectx} we have that
  $\Gamma = \Gamma_1$. Reduction can occur by \riffalse such that
  $\oprule{t}{\mu}{t_f}{\mu}$. By \tsub we hhave that
  $\typerule{\Gamma_1}{t_f}{T}{\Gamma_3}$ as $T_f <: T$. Additionally, for
  all $x \in \Gamma_3$, $\Gamma_3(x) <: \Gamma'(x)$. Therefore by
  repeatedly applying \twiden for all $x$, we have that
  $\typerule{\Gamma_1}{t_f}{T}{\Gamma'}$. Let $\Gamma'' = \Gamma$,
  $\mu' = \mu$ and $t' = t_f$. Therefore, 
  $\typerule{\Gamma''}{t'}{T}{\Gamma'}$ with $\Gamma'' \vdash \mu'$.

  \item $t_c$ is not a value. By induction, there exists $t_c'$,
  $\mu'$, $\Gamma''$ such that 
  $\oprule{t_c}{\mu}{t_c'}{\mu'}$ and 
  $\typerule{\Gamma''}{t_c'}{\boolt}{\Gamma_1}$ with $\Gamma'' \vdash \mu'$. 
  Reduction of $t$ can therefore occur by \rifexpr such that
  $\oprule{t}{\mu}{\ift{t_c'}{t_t}{t_f}}{\mu'}$. It follows
  by \tif that $\typerule{\Gamma''}{\ift{t_c'}{t_t}{t_f}}{T}{\Gamma'}$.
  \end{itemize}

\end{itemize}
\end{proof}

\section{Supporting lemmas}
\label{app:proofs}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:valuectx %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:valuectx}
If $v$ is a value and $\typerule{\Gamma}{v}{T}{\Gamma'}$, then $\Gamma = \Gamma'$.
\end{lem}
\begin{proof}
Directly from the typing judgements. $v$ is either {\tt unit}, {\tt true},
{\tt false}, a function literal or an object literal.
Correspondingly, by rules \tunit,
\ttrue, \tfalse, 
\tfundef and \tobj, $\Gamma = \Gamma'$.
\end{proof}

\begin{lem}[Object state sets]
Given an object state set $\overline{S}$ and a state $S'$, if
$S' \in \overline{S}$ then $O@S' <: O@\overline{S}$.
\end{lem}
\begin{proof}
$O@\overline{S}$ is defined as equivalent to
$\bigsqcup_{S \in \overline{S}}{O@S}$. Given that $S \in \overline{S}$,
we must have that $\overline{S} = \overline{S'} \cup \{ S \}$ for
some $\overline{S'}$.
it follows that $O@\overline{S} = O@S' \sqcup \left( 
\bigsqcup_{S \in \overline{S'}}{O@S} \right)$.
By the definition of join, 
$O@S' <: O@S' \sqcup \bigsqcup_{S \in \overline{S}}{O@S}$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:ssetinc  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:ssetinc}
Given an object protocol $O = \left\{ \overline{ S_i \{ ... \} } \right\}$ and two state sets 
$\overline{S} \subseteq \overline{S'} \subseteq \overline{S_i}$, it follows
that $O@\overline{S} <: O@\overline{S'}$.
\end{lem}
\begin{proof}
From the definition of state sets, we have that 
$O@\overline{S} = \bigsqcup_{S \in \overline{S}}{O@S}$ and that
$O@\overline{S'} = \bigsqcup_{S \in \overline{S'}}{O@S}$.
As $\overline{S} \subseteq \overline{S'}$ there exists an
$\overline{S''}$ such that $\overline{S'} = \overline{S} \cup \overline{S''}$
where $\overline{S} \cap \overline{S''} = \emptyset$. Therefore
$O@\overline{S'} = \left(\bigsqcup_{S \in \overline{S}}{O@S}\right) \sqcup \left(\bigsqcup_{S' \in \overline{S'} - \overline{S}}{O@S'}\right)$.

As $A <: A \sqcup B$ for all $A$ and $B$ where $A \sqcup B$ is defined, 
it follows that $O@\overline{S} <: O@\overline{S'}$.
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}
\label{lem:remap}
Given types $T$, $U$, and $V$, such that
$\remapapp{T}{U}{V}$ is defined, it follows that $\remapapp{T}{U}{V} <: V$.
\end{lem}
\begin{proof}
By case analysis of V.

\begin{itemize}
\item $V = \unitt$ or $V = \boolt$. By either \remapunit or \remapbool
it must be the case that $U = V$. It must also be the case that
$T = U$, as $\unitt$ and $\boolt$ do not have any subtypes other than
themselves. Therefore $V = \remapapp{T}{U}{V} <: V$ by \subrefl.

\item $V = \funv{V_i}{V_i'}{V'}$ for some $\overline{V_i}$, $\overline{V_i'}$
and $V'$. By \remapfunc, $T <: U$ and $T = \remapapp{T}{U}{V}$. Therefore,
$T <: V$.

\item $V = O@\overline{S'}$ for some $O$ and $\overline{S'}$. By
\remapobj, $O'@\overline{S'''} = \remapapp{T}{U}{V}$ where 
$T = O'@\overline{S''}$ and $U = O@\overline{S}$ and
$\overline{S'''} = \{ S | O'@S \sim O@\overline{S'}, S \in O' \}$ 
for some $O'$, $\overline{S}$ and $\overline{S''}$. By \subobj,
$T = O'@\overline{S'''} <: O@\overline{S'} = V$ as for each $S \in \overline{S'''}$,
$O'@S <: O@\overline{S'}$.
\end{itemize}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapcv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}[Remap is covariant]
\label{lem:remapcv}
Given types $T, T', U$ and $V$ such that $T <: T'$, 
$\remapapp{T}{U}{V} <: \remapapp{T'}{U}{V}$.
\end{lem}

\begin{lem}
\label{lem:remapeff}
Given types $T$, $U$ and $V$ such that $U <: V$ and $T = \remapapp{U}{V}{V}$, 
it follows that $U <: T$. \\
(TODO: object portion needs reworked due to new definition of remap)
\end{lem}
\begin{proof}
By case analysis of the type $V$.

\begin{itemize}
\item $V = \unitt$ or $V = \boolt$. 
It must be the case that $V = U$ also, as $\unitt$ and $\boolt$
have no subtypes other than themselves by rule \subrefl.
Therefore, by rule \remapunit or \remapbool, $T = V$,
and $U <: T$ by rule \subrefl.

\item $V = \funv{T_i}{T_i'}{T_r}$ for some $\overline{T_i}$, $\overline{T_i'}$
and $T_r$. As $U <: V$, it follows by rule \remapfunc that
$T = U$, and $U <: T$ by \subrefl.

\item $V = O@\overline{S}$ for some $O$ and $\overline{S}$. 
It follows that $U = O'@\overline{S'}$ 
for some $O'$ and $\overline{S'}$ 
such that $O'@\overline{S'} \sim O@\overline{S}$ 
by rule \subobj. 
By rule \remapobj, it follows that $T = O'@\overline{S''}$ 
such that $\overline{S''} = \{ S | O'@S \sim O@\overline{S}, S \in O' \}$.
By \lemref{lem:ssetinc} it follows that 
$O'@\overline{S'} <: O'@\overline{S''}$.

\end{itemize}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{cor}
\label{lem:remapsame}
Given a type $T$, $T <: \remapapp{T}{T}{T}$. \\
(TODO: in fact, with the new definition of remap it should be possible to
state that $T = \remapapp{T}{T}{T}$).
\end{cor}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:upgrade %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}[remap]
\label{lem:upgrade}
If a term $t$ can be typed such that 
$\typerule{\Gamma, x : T}{t}{U}{\Gamma', x : V}$, and if
$T' <: T$, then there exists a $V' <: \remapapp{T'}{T}{V}$ 
such that $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
\end{lem}
\begin{proof}
by induction on the structure of $t$.
The term $t$ is of one of the following forms:

\begin{itemize}
\item $t$ is a value, so by \lemref{lem:valuectx} it follows that
$\Gamma = \Gamma'$ and $T = V$. 
The environment can be arbitrary in the typing rules for values,
therefore we can type $t$ such that 
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : T'}$.
Let $V' = T'$.
By \lemref{lem:remapeff}, we have that $V' <: \remapapp{T'}{T}{V}$.
Therefore, $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\item $t = \lett{y}{t'}{t''}$. It follows from the rule \tlet that
$x \neq y$ (as variables cannot be rebound) and that there exists 
$\Gamma_1, \Gamma_2, T_2, T_y, T_y'$ such that
$\typerule{\Gamma, x : T}{t'}{T_y}{\Gamma_1, x : T_2}$ and that
$\typerule{\Gamma_1, x : T_2, y : T_y}{t''}{U}{\Gamma', x : V, y : T_y'}$.
By induction, it follows that there exists a type $V_2 <: \remapapp{T'}{T}{T_2}$ such that
$\typerule{\Gamma, x : T'}{t'}{T_y}{\Gamma_1, x : V_2}$.
By \lemref{lem:remap} we have that $V_2 <: T_2$.
Therefore it also follows by induction that
there exists a type $V_3 <: \remapapp{V_2}{T_2}{V}$
such that
$\typerule{\Gamma_1, x : V_2, y : T_y}{t''}{U}{\Gamma', x : T_3, y : V_3}$.
Let $V' = V_3$. The rules of \tlet are satisfied such that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\item $t = y ( \overline{z_i} )$. 
It follows from the rule \tfunc that there exists
$\Gamma_1, \overline{T_i}, \overline{U_i}, \overline{V_i}, \overline{T_i'}$ 
such that 
$\Gamma, x : T = \Gamma_1, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i}$
and that
$\Gamma', x : V = \Gamma_1, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i'}$
such that for each $i$, $T_i <: U_i$, $T_i' <: V_i$ 
and $U <: U'$.

There are three cases to consider:
\begin{itemize}
\item $x \neq y$ and $x \notin \overline{x_i}$, meaning that
$\Gamma_1 = \Gamma_2, x : T$ for some $\Gamma_2$, that
$\Gamma = \Gamma_2, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i}$
and that $\Gamma' = \Gamma_2, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i'}$.
Additionally, $T = V$. 
It follows that $t$ can be typed without $x$
such that $\typerule{\Gamma}{t}{U}{\Gamma'}$. Let $V' = T'$.
By the weakening lemma, $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.

\item $x = y$, meaning $\Gamma = \Gamma_1, \overline{z_i : T_i}$ and
$\Gamma' = \Gamma_1, \overline{z_i : T_i'}$. 
Additionally, $T = V = \funt{U_i}{V_i}{U}$.
$T'$ must therefore be of form $\funt{U_i'}{V_i'}{U'}$ 
such that for each $i$, $U_i <: U_i'$, $V_i' <: V_i$, $U' <: U$ by the
definition of subtyping. Let $V' = T'$.
By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.
It follows by rule \tfunc, and \tsub to replace $U'$ with $U$, that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\item $x \in \overline{z_i}$, meaning that $T = T_i$ and 
$V = T_i' = \remapapp{T}{U_i}{V_i}$ for some $i$.
Let $V' = \remapapp{T'}{U_i}{V_i}$. 
By \lemref{lem:remapcv}, $V' <: V$.
As $T' <: T$, we can type $t$ by rule \tfunc such that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma, x : V'}$.

\end{itemize}

\item $t = y.m$. By the rule \tmethc it follows that
$\Gamma, x : T = \Gamma_1, y : O@\overline{S}$
and that
$\Gamma', x : V = \Gamma_1, y : O@\overline{S'}$
for some $\Gamma_1, O, \overline{S}$ and $\overline{S'}$
where
$\overline{S'} = \{ S_n | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \}$.
Additionally, 
$U = \bigsqcup{ \{ T_r | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \} }$.
There are two cases to consider:

\begin{itemize}
\item $x = y$. It follows that $\Gamma = \Gamma_1 = \Gamma'$ and
that $T = O@\overline{S}, V = O@\overline{S'}$.
$T'$ must be of form $O'@\overline{S}''$ such that
$O'@\overline{S}'' \sim O@\overline{S}$ by rule \subobj.

Let $V' = O'@\overline{S}''' = \remapapp{T'}{T}{V}$
where
$S''' = \{ S_n' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}''\}$.
Let $U' = \bigsqcup{ \{ T_r' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}'' \} }$.
As $O'@\overline{S}'' \sim O@\overline{S}$, it follows that
$O'@\overline{S}''' \sim O@\overline{S}'$.
Therefore, $U' <: U$ and $V' <: V$. By the rule \tmethc, and \tsub
to replace $U'$ with $U$, it follows that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\item $x \neq y$. It follows that $\Gamma = \Gamma'', y : O@\overline{S}$
and that $\Gamma' = \Gamma'', y : O@\overline{S'}$
for some $\Gamma''$, $O$ and $\overline{S}$, and that
$T = V$.
By rule \tmethc we have that $\typerule{\Gamma}{t}{U}{\Gamma'}$.
Let $V' = T'$. By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.
By the weakening lemma, 
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\end{itemize}

\item $t = t' ; t''$. It follows by rule \tlet that there exists
$\Gamma'', V_{mid}, U'$ 
such that $\typerule{\Gamma, x : T}{t'}{U'}{\Gamma'', x : V_{mid}}$
and that $\typerule{\Gamma'', x : V_{mid}}{t''}{U}{\Gamma', x : V}$.
By induction, it follows that there exists $V_{mid}' <: \remapapp{T'}{T}{V_{mid}}$ 
such that $\typerule{\Gamma, x : T'}{t'}{U'}{\Gamma'', x : V_{mid}'}$.
By \lemref{lem:remap} we have that $V_{mid}' <: V_{mid}$.
There it also follows by induction that 
there exists $V_{end} <: \remapapp{V_{mid}'}{V_{mid}}{V}$ such that
$\typerule{\Gamma'', x : V_{mid}'}{t''}{U}{\Gamma'', x : V_{end}}$.
Let $V' = V_{end}$. By rule \tlet it follows that
$\typerule{\Gamma, x : T'}{t' ; t''}{U}{\Gamma'', x : V'}$.

\item $t = \ift{t_c}{t_t}{t_f}$. It follows from rule \tif that there
exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $U_t$, $U_f$, $V_c$, $V_t$ and
$V_f$
such that
$\typerule{\Gamma, x : T}{t_c}{\boolt}{\Gamma_1, x : V_c}$ and that
$\typerule{\Gamma_1, x : V_c}{t_t}{U_t}{\Gamma_2, x : V_t}$ and that
$\typerule{\Gamma_1, x : V_c}{t_f}{U_f}{\Gamma_3, x : V_f}$ with
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$, $U = U_t \sqcup U_f$ and
$V = V_t \sqcup V_f$.

By induction, there exists a $V_c' <: \remapapp{T'}{T}{V_c}$
such that $\typerule{\Gamma, x : T'}{t_c}{\boolt}{\Gamma_1, x : V_c'}$.
By \lemref{lem:remap} we also have that $V_c' <: V_c$.
Therefore by induction there also exists a 
$V_t' <: \remapapp{V_c'}{V_c}{V_t}$ and $V_f' <: \remapapp{V_c'}{V_c}{V_f}$
such that $\typerule{\Gamma_1, x : V_c'}{t_t}{T_t}{\Gamma_2, x : V_t'}$
and that $\typerule{\Gamma_1, x : V_c'}{t_f}{T_f}{\Gamma_3, x : V_f'}$.
By \lemref{lem:remap} we also have that $V_t' <: V_t$ and that
$V_f' <: V_f$.
Let $V' = V_t' \sqcup V_f'$. By the definition of join, $V' <: V$.
It follows by rule \tif that
$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.

\end{itemize}
\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:weakening %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}[Weakening]
\label{lem:weakening}
If $\typerule{\Gamma}{t}{T}{\Gamma'}$,
then for all $\Gamma''$ such that $dom(\Gamma) \cap dom(\Gamma'') = \emptyset$,
$t$ can also be typed such that
$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
\end{lem}
\begin{proof}
by induction on the structure of $t$.

If $t$ is some value $v$, it follows by \lemref{lem:valuectx} that 
$\Gamma = \Gamma'$ and that the typing derivation will still be valid as long
as the input and output contexts are identical. Therefore,
$\typerule{\Gamma , \Gamma''}{v}{T}{\Gamma', \Gamma''}$.

If $t$ is not a value, then it is of one of the following forms:

\begin{itemize}
\item $t = \lett{x}{t'}{t''}$. It follows from rule \tlet that
there exists $\Gamma_1, \Gamma_2, T', T''$ such that
$\typerule{\Gamma}{t'}{T'}{\Gamma_1}$ and
$\typerule{\Gamma_1, x : T'}{t''}{T}{\Gamma', x : T''}$.
By \tlet we have that $dom(\Gamma') = dom(\Gamma)$.
By induction, $t'$ can be typed such
that $\typerule{\Gamma,\Gamma''}{t'}{T'}{\Gamma_1,\Gamma''}$,
and $t''$ can be typed such that
$\typerule{\Gamma_1, \Gamma'', x : T'}{t''}{T}{\Gamma', \Gamma'', x : T''}$,
as we can assume by Barendregt's convention that $x$ can be made distinct from
any variable names in $\Gamma''$ by relabeling $x$.
Therefore,
$\typerule{\Gamma, \Gamma''}{\lett{x}{t'}{t''}}{T}{\Gamma', \Gamma''}$.

\item $t = x ( \overline{x_i} )$. It follows from the rule \tfunc that
there exists a $\Gamma_1, \overline{T_i}, \overline{U_i}, \overline{V_i}, \overline{T_i'}$ 
such that 
$\Gamma = \Gamma_1, x : (\overline{U_i \gg V_i}) \rightarrow T, \overline{x_i : T_i}$
and that
$\Gamma' = \Gamma_1, x : (\overline{U_i \gg V_i}) \rightarrow T, \overline{x_i : T_i'}$.
As $\Gamma_1$ can be arbitrary in the rule \tfunc, it follows that we
can extend the input and output contexts such that
$\typerule{\Gamma, \Gamma''}{x ( \overline{x_i} )}{T}{\Gamma', \Gamma''}$.

\item $t = x.m$. It follows from the rule \tmethc that there exists
a $\Gamma_1$ such that $\Gamma = \Gamma_1, x : O@\overline{S}$ and that 
$\Gamma' = \Gamma_1, x : O@\overline{S'}$.
As $\Gamma_1$ can be arbitrary in rule \tmethc, it follows that we can
extend the input and out contexts such that
$\typerule{\Gamma, \Gamma''}{x.m}{T}{\Gamma', \Gamma''}$.

\item $t = t' ; t''$. It follows from the rule \tseq that there exists
$\Gamma_1$ and $T'$ such that $\typerule{\Gamma}{t'}{T'}{\Gamma_1}$
and $\typerule{\Gamma_1}{t''}{T}{\Gamma'}$. By induction
$\typerule{\Gamma, \Gamma''}{t'}{T'}{\Gamma_1, \Gamma''}$
as $dom(\Gamma_1) \subseteq dom(\Gamma)$ and therefore
$\Gamma_1$ and $\Gamma''$ are disjoint.
Similarly, as $dom(\Gamma_1) = dom(\Gamma')$, by induction
$\typerule{\Gamma_1, \Gamma''}{t''}{T}{\Gamma', \Gamma''}$.
Therefore $\typerule{\Gamma, \Gamma''}{t'}{T'}{\Gamma_1, \Gamma''}$.

\item $t = \ift{t_c}{t_t}{t_f}$. It follows from rule \tif that there
exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$ and $T_f$ such that
$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
$\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ with
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$.
By induction each of $\Gamma_1$, $\Gamma_2$ and $\Gamma_3$ can
be extended by $\Gamma''$ in the typing judgements for $t_c$,
$t_t$ and $t_f$. As $\Gamma'' \sqcup \Gamma'' = \Gamma''$, it
follows that $(\Gamma_2, \Gamma'') \sqcup (\Gamma_3, \Gamma'') = 
\Gamma_2 \sqcup \Gamma_3, \Gamma''$ and therefore $t$ can be
typed such that $\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.

\end{itemize}

\end{proof}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:subst %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{lem}[Substitution]
\label{lem:subst}
If $\typerule{\Gamma}{t}{T}{\Gamma'}$, then
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t}}
{T}
{\subst{x_i}{y_i}{\Gamma'}}$
where each $x_i$ and $y_i$ is distinct, and $\overline{x_i} \cap \overline{y_i} = \emptyset$.
\end{lem}
\begin{proof}
by induction on the structure of $t$.

If $t$ is some value $v$, then $\Gamma = \Gamma'$, therefore
$\subst{x_i}{y_i}{\Gamma} = \subst{x_i}{y_i}{\Gamma'}$
and substitution has no effect on the value, which has no specific requirements
of either $\Gamma$ or $\Gamma'$ other than that they are equal.
Therefore, $\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{v}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.

If $t$ is a term, then it is of one of the following forms:

\begin{itemize}
\item $t = \lett{x}{t''}{t'}$. 
Substitution is defined on this term such that \\
$\subst{x_i}{y_i}{t} \equiv 
\lett{x}{( \subst{x_i}{y_i}{t''} )}{(\subst{x_j}{y_j}{t'})}$
, where $\overline{y_j} = \overline{y_i} - \{ x \}$.

By rule \tlet, it follows
that there exists $T', T'', \Gamma_1$ such that
$\typerule{\Gamma}{t''}{T''}{\Gamma_1}$ and that
$\typerule{\Gamma_1, x : T''}{t''}{T}{\Gamma', x : T'}$.
Additionally, $x \notin dom(\Gamma)$.

By induction, $t''$ can be substituted such that 
$\typerule
{\subst{x_i}{y_i}{\Gamma}}
{\subst{x_i}{y_i}{t''}}
{\subst{x_i}{y_i}{\Gamma_1}}
$.
Additionally, $t'$ can be substituted such that
$(\subst{x_j}{y_j}{\Gamma_1}) , x : T'' \triangleright \subst{x_j}{y_j}{t''} \triangleleft \{ (\overline{x_j / y_j} \} \Gamma') , x : T'$ as it is guaranteed that $x \notin \overline{y_j}$.
As $dom(\Gamma') = dom(\Gamma)$, and that $x \notin dom(\Gamma)$, it follows that $\subst{x_i}{y_i}{\Gamma_1} = (\subst{x_j}{y_j}{\Gamma_1})$ and
that $\overline{x_i / y_i} \} \Gamma' = \overline{x_j / y_j} \} \Gamma'$.

Therefore, the requirements of \tlet are satisfied such that
$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.

\item $t = t_l ; t_r$.
Substitution is defined on this term such that
$\{ \overline{x_i / y_i} \} ( t_l ; t_r ) \equiv \subst{x_i}{y_i}{t_l} ; \subst{x_i}{y_i}{t_r}$.

By rule \tseq, there exists $\Gamma'', T'$ such that
$\typerule{\Gamma}{t_l}{T'}{\Gamma''}$ and
$\typerule{\Gamma''}{t_r}{T}{\Gamma'}$. By induction,
$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t_l}}{T'}{\subst{x_i}{y_i}{\Gamma''}}$ and
$\typerule{\subst{x_i}{y_i}{\Gamma''}}{\subst{x_i}{y_i}{t_r}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
Therefore, $t$ can by typed \tseq after the substitution.

\item $t = x_0 ( \overline{x_k} )$.
Substition is defined on this term such that
$\subst{x_i}{y_i}{x_0} ( \overline{x_k} ) \equiv y_0 ( \overline{y_k} )$
where for each $x \in \{ x_0, \overline{x_k} \}$,
$x = y_j \implies y = x_j$, otherwise $y$ is the original value.

For each substitution that occurs in the term, the variable must exist in
both the input and output contexts by rule \tfunc, and will be substituted
for the same name. Therefore, $t$ can be typed by \tfunc after the
substitution.

\item $t = x.m$.
Substition is defined on this term such that
$\subst{x_i}{y_i}{x.m} \equiv y.m$
where $y = x_j$ if $x = y_j$ for some $y_j \in \overline{y_i}$.
Otherwise, $y = x$.

Rule \tmethc requires that $x$ exist in both $\Gamma$ and $\Gamma'$ and
therefore any substitution of x in the term will be matched with the
same substitution in the contexts. Therefore, $t$ can be typed by
\tmethc after the substitution.

\item $t = \ift{t_c}{t_t}{t_f}$. Substitution is defined on this term such
that \\
$\subst{x_i}{y_i}{(\ift{t_c}{t_t}{t_f})} \equiv 
\ift{\substxy{t_c}}{\substxy{t_t}}{\substxy{t_f}}$.
By rule \tif there exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$, $T_f$
such that
$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
$\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ where
$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$. By induction
substitution can be performed on $t_c$, $t_t$ and $t_f$ such that
$\typerule{\substxy{\Gamma}}{\substxy{t_c}}{\boolt}{\substxy{\Gamma_1}}$ and that
$\typerule{\substxy{\Gamma_1}}{\substxy{t_t}}{T_t}{\substxy{\Gamma_2}}$ and
$\typerule{\substxy{\Gamma_1}}{\substxy{t_f}}{T_f}{\substxy{\Gamma_3}}$.
By the definition of substitution on contexts, 
$\substxy{\Gamma_2} \sqcup \substxy{\Gamma_3} = \substxy{\Gamma'}$. Therefore,
by rule \tif, 
$\typerule{\substxy{\Gamma}}{\substxy{\ift{t_c}{t_t}{t_f}}}{T}{\substxy{\Gamma'}}$.
\end{itemize}
\end{proof}

\bibliographystyle{abbrvnat}
\bibliography{refs}

% The bibliography should be embedded for final submission.

%\begin{thebibliography}{}
%\softraggedright

%\bibitem[Smith et~al.(2009)Smith, Jones]{smith02}
%P. Q. Smith, and X. Y. Jones. ...reference text...

%\end{thebibliography}

\end{document}

### 2012-10-12 src/main/scala/ConstraintGenerator.scala

@@ -293,8 +293,7 @@ object ConstraintGenerator {
   def ifConstraints(t : If) =
     (ConstraintSet() +
       EqualityConstraint(VarTE(t.condition->typeVar), BoolTE) +
-      SubtypeConstraint(VarTE(t.whenTrue->typeVar), VarTE(t->typeVar)) +
-      SubtypeConstraint(VarTE(t.whenFalse->typeVar), VarTE(t->typeVar)) +
+      JoinConstraint(VarTE(t->typeVar), VarTE(t.whenTrue->typeVar), VarTE(t.whenFalse->typeVar)) +
       ContextConstraint(t.condition->inContextVar, sameAs(t->inContextVar)) +
       ContextConstraint(t.whenTrue->inContextVar, sameAs(t.condition->outContextVar)) +
       ContextConstraint(t.whenFalse->inContextVar, sameAs(t.condition->outContextVar)) +

### 2012-10-12 src/main/scala/ConstraintSolver.scala

@@ -81,9 +81,13 @@ class ConstraintSolver(t : Term) {
 
     val (contexts, eqConstraints, subConstraints) = 
       reduceToTypeConstraints(constraints)
+
+    val stubbedJoins = constraints.jcs.flatMap(jc => 
+      Seq(EqualityConstraint(jc.a, jc.left), 
+        EqualityConstraint(jc.a, jc.right)))
     
     solveTypeConstraints(contexts, 
-      eqConstraints, 
+      eqConstraints ++ stubbedJoins, 
       subConstraints,
       constraints.mcs)
   }

### 2012-10-12 src/main/scala/Constraints.scala

@@ -27,6 +27,10 @@ case class SubtypeConstraint(a : TypeExpr, b : TypeExpr) extends Constraint {
   override def toString = a + " <: " + b
 }
 
+case class JoinConstraint(a : TypeExpr, left : TypeExpr, right : TypeExpr) extends Constraint {
+  override def toString = a + " = " + left + " ∨ " + right
+}
+
 /** 
  * Represents evidence that the specified context variable must be
  * composed of some other context and some explicit modifications,
@@ -131,12 +135,14 @@ case class ConstraintSet(
     cvcs : Seq[ContextVarConstraint] = Seq.empty,
     tecs : Seq[EqualityConstraint] = Seq.empty,
     scs : Seq[SubtypeConstraint] = Seq.empty,
+    jcs : Seq[JoinConstraint] = Seq.empty,
     mcs : Seq[MethodConstraint] = Seq.empty) {
 
   def +(cc : ContextConstraint) = this.copy(ccs = cc +: ccs)
   def +(cvc : ContextVarConstraint) = this.copy(cvcs = cvc +: cvcs)
   def +(tec : EqualityConstraint) = this.copy(tecs = tec +: tecs)
   def +(sc : SubtypeConstraint) = this.copy(scs = sc +: scs)
+  def +(jc : JoinConstraint) = this.copy(jcs = jc +: jcs)
   def +(mc : MethodConstraint) = this.copy(mcs = mc +: mcs)
   
 
@@ -146,11 +152,12 @@ case class ConstraintSet(
       cvcs ++ others.cvcs,
       tecs ++ others.tecs,
       scs ++ others.scs,
+      jcs ++ others.jcs,
       mcs ++ others.mcs)
 
   override def toString = {
     val sortedCcs = ccs.sortBy(_.context.v)
-    val typeCs = tecs ++ scs
+    val typeCs = tecs ++ scs ++ jcs
     ("context constraints:\n\t" +
       (if (sortedCcs.isEmpty) "none" else sortedCcs.mkString("\n\t")) +
       "\nvar constraints:\n\t" +
@@ -164,7 +171,7 @@ case class ConstraintSet(
 
   def toStringSimple = {
     val sortedCcs = ccs.sortBy(_.context.v)
-    (sortedCcs ++ cvcs ++ tecs ++ scs ++ mcs).mkString("; ")
+    (sortedCcs ++ cvcs ++ tecs ++ scs ++ jcs ++ mcs).mkString("; ")
   }
 }
 

### 2012-11-05 formalism/paper.tex
addition of new join-of-separate-types type constraint to formalism (still needs full discussion).

@@
 \usepackage{mathtools}
+
+\DeclareSymbolFont{symbolsC}{U}{ntxsyc}{m}{n}
+\SetSymbolFont{symbolsC}{bold}{U}{ntxsyc}{b}{n}
+\DeclareMathSymbol{\sqcupplus}{\mathbin}{symbolsC}{124}
 
@@
 \infer
-{\mu[x \mapsto v] = \mu, x \mapsto v}
+{\mu[x \mapsto v] = \overline{x_i \mapsto v_i} , x \mapsto v}
 {\mu = \overline{x_i \mapsto v_i}, x \mapsto v'}
@@
 in relation to a context as in \figref{fig:storetype}. A well-typed store
-may mappings for variables which are not in the context as shown in
+may contain mappings for variables which are not in the context as shown in
 rule \textsc{st\_extra}.
@@
 {\overline{x_i : T_i} \vdash \overline{x_i \mapsto v_i}}
-{\typerule{\Gamma}{v_i}{T_i}{\Gamma'}}
+{\overline{\typerule{\Gamma}{v_i}{T_i}{\Gamma'}}}
 \\\\
@@
 \infer[\textsc{st\_extra}]
-{\Gamma \vdash \mu, \overline{y_j \mapsto v_j}}
-{\Gamma \vdash \mu & \overline{y_j \notin dom(\mu)}}
+{\Gamma \vdash \mu, \overline{x_i \mapsto v_i}}
+{\Gamma \vdash \mu & \overline{x_i \notin dom(\mu)}}
 }
@@
 \begin{itemize}
-\item $S \in O$, a predicate asserting $S$ is a state of $O$, is equivalent to
-$S \in \Sigma$.
-\item $m : T \Rightarrow S' \in O@S$, the method existence predicate, 
-states the method $m$ with return type $T$ exists in state $S$ and triggers a 
-transition to state $S'$ when called. Formally, $(S,m,T,S') \in \Delta$.
+\item $S \in O$ is the {\it state existence predicate} asserting 
+$S$ is a state of $O$. 
+Formally, this is equivalent to $S \in \Sigma$.
+\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
+asserting the existence
+of method $m$ with return type $T$ in state $S$ which triggers a 
+transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
+Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
 \item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
-of the previous to state sets, meaning that for all $S \in \overline{S}$
-there exists a state $S'$ and type $T'$ where 
-$(S,m,T',S') \in \Delta$ and that
+of the method existence predicate to state sets, meaning that for all 
+$S \in \overline{S}$ there exists a state and type 
+$T'$ where $(S,m,T',S') \in \Delta$ and that
 $T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
 and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
-\item $O@\overline{S} \sim O_2@\overline{S_2}$, the extension of the
-simulation relation to support state sets, can be interpreted as
-as $\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
+\item $O@\overline{S} \sim O_2@\overline{S_2}$ is the extension of the
+simulation relation to support state sets, and can be interpreted as
+$\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
 \end{itemize}
@@
 \otttinf\ottinterrule
+\ottoinf\ottinterrule
+\ottsinf\ottinterrule
 }
@@
 \[
-\begin{array}{l}
+\begin{array}{ll}
 
@@
 % object value
-\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@S}
-\triangleq
-\exists \overline{\alpha_{ij}} . \\
-\alpha = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\}@S
+\inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@\overline{S}}
+\triangleq \\
+\exists \alpha_o, \alpha_s, \overline{\alpha_{ij}}, \gamma_{\emptyset} . \\
+\gamma_{\emptyset} = \emptyset
 \cand
+\alpha = \alpha_o@\alpha_s
+\cand
+\overline{\inferrule{\gamma_{\emptyset}}{v_{ij}}{\alpha_{ij}}{\gamma_{\emptyset}}}
+\cand
+\\
 \gamma = \gamma'
-\cand \\
-\overline{\inferrule{\gamma}{\triangleright v_{ij}}{\alpha_{ij}}{\gamma}}
+\cand
+\alpha_o @ \alpha_s = \{ \overline{ S_i \{ \overline{ m_{ij} : \alpha_{ij} >> S_{ij} } \} }\} @ \overline{S}
 \\\\
@@
 % function value
-\inferlhs{\lambda ( \overline{x_i} ) . t} \triangleq
-\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_{ai}, \alpha_{bi}}, \alpha_r . \\
-\alpha = (\overline{\alpha_{ai} \gg \alpha_{bi}}) \rightarrow \alpha_r
+\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \triangleq \\
+\exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_i, \alpha_i'}, \alpha_r . \\
+\alpha = (\overline{\alpha_i \gg \alpha_i'}) \rightarrow \alpha_r
 \cand
@@
 \cand \\
-\gamma_{in} = \gamma_r, \overline{x_i : \alpha_{ai}}
+\gamma_{in} = \gamma_r, \overline{x_i : \alpha_i}
 \cand
-\overline{x_i : \alpha_{bi} \in \gamma_{out}}
+\overline{x_i : \alpha_i' \in \gamma_{out}}
 \cand
@@
 \inferrule{\gamma_{in}}{t}{\alpha_r}{\gamma_{out}}
+\\
+\cand
+\overline{T_i <: \alpha_i}
+\cand
+\overline{\alpha_i' <: T_i'}
 \\\\
@@
 % let binding
-\inferlhs{\lett{x}{t'}{t}} \triangleq \exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
+\inferlhs{\lett{x}{t'}{t}} \triangleq \\
+\exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
 \gamma_{b1} = \gamma_v, x : \alpha'
@@
 % sequence
-\inferlhs{t ; t'} \triangleq
+\inferlhs{t ; t'} \triangleq \\
 \exists \gamma'', \alpha' . \\
@@
 % function call
-\inferlhs{x ( \overline{x_i} )} \triangleq
+\inferlhs{x ( \overline{x_i} )} \triangleq \\
 \exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
@@
 % method call
-\inferlhs{x.m} \triangleq
-\exists \alpha_x . \\
-x : \alpha_x \in \gamma
-\cand 
+\inferlhs{x.m} \triangleq \\
+\exists \alpha_o, \alpha_s . \\
+x : \alpha_o@\alpha_s \in \gamma
 \cand \\
-\gamma' = \gamma[ x \mapsto \remapapp{\alpha_x}{O@S}{O@S'} ]
+\gamma' = \gamma[ x \mapsto \remapapp{\alpha_o@\alpha_s}{O@S}{O@S'} ]
 \\
-\mathrm{where } O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
+\mathbf{where}\; O = \{ S \{ m : \alpha \Rightarrow S' \} S' \{ \} \}
 \\\\
@@
 % if-then-else
-\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq
-\\
+\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq \\
 \exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
@@
 \\
-\alpha = \alpha_t \sqcup \alpha_f
+\alpha = \alpha_t \sqcupplus \alpha_f
 \cand

### 2012-11-27 formalism/paper.tex
proof of context normalisation almost complete, interim commit before final pass

@@
 \newcommand{\subst}[3]{#3 \{\overline{^{#1}/_{#2}}\}}
+\newcommand{\substo}[3]{#3 \{^{#1}/_{#2}\}}
 \newcommand{\substxy}[1]{\subst{x_i}{y_i}{#1}}
@@
 \newcommand{\ift}[3]{\mathtt{if} \; #1 \; \mathtt{then} \; #2 \; \mathtt{else} \; #3}
+
 \newcommand{\cand}{\:\wedge\:}
@@
 \newcommand{\fresh}[1]{\mathbf{fresh}\:#1}
+\newcommand{\partialo}[2]{\mathbf{partial}\:\overline{\mathbb{S}_{#1}{#2}}}
+\newcommand{\fixedo}[2]{\mathbf{fixed}\:\overline{\mathbb{S}_{#1}{#2}}}
+\newcommand{\os}[2]{\omega_{#1}@\sigma_{#2}}
+
+\newcommand{\tinf}{\mathbb{T}}
 
@@
 % unit value
-\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma = \gamma' 
+\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma' = \gamma
 \\\\
@@
 % true value
-\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
+\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
 \\\\
@@
 % false value
-\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma = \gamma' 
+\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
 \\\\
@@
 \inferlhs{\lett{x}{t'}{t}} \triangleq \\
-\exists \gamma_v, \gamma_{b1}, \gamma_{b2}, \alpha' . \\
-\gamma_{b1} = \gamma_v, x : \alpha'
+\exists \gamma_v, \gamma_{in}, \gamma_{out}, \alpha' . \\
+\gamma_{in} = \gamma_v, x : \alpha'
 \cand
-\gamma' = \gamma_{b2} - \{ x \}
+\gamma' = \gamma_{out} - \{ x \}
 \cand \\
@@
 \cand
-\inferrule{\gamma_{b1}}{t}{\alpha}{\gamma_{b2}}
+\inferrule{\gamma_{in}}{t}{\alpha}{\gamma_{out}}
 \\\\
@@
 Additionally, we have {\bf variable usage constraints} of form 
-$x : \mathbb{T} \in \gamma$, which are generated for each location in the
+$x : \tinf \in \gamma$, which are generated for each location in the
 program where a variable is used and what type that usage implies. These
 are used to generate additional type variable
-constraints by matching the asserted type $\mathbb{T}$ with what is known
+constraints by matching the asserted type $\tinf$ with what is known
 about the type of $x$ in $\gamma$.
@@
 Type variables can appear at any position within a type's structure, as shown
-in the grammar for $\mathbb{T}$. Objects are treated as a special case, where
+in the grammar for $\tinf$. Objects are treated as a special case, where
 we use variables as placeholders for the object definition and state parts.
@@
 not change the context. Additionally, a type variable constraint of form
-$\alpha = \mathbb{T}$ where $\mathbb{T}$ is of the appropriate type to match
+$\alpha = \tinf$ where $\tinf$ is of the appropriate type to match
 the value, with type variables in all positions relating to parameters and return
@@
 
+\subsection{Constraint solving}
+
+Constraint solving is performed in phases. First, generated context constraints
+for a term $t$ aresolved, which determines the contents of each context associated to
+each context variable in the typing derivation of $t$, and produces additional
+type equality constraints (of form $\mathbb{T} = \mathbb{T}'$) based on
+observed usages of variables. If a variable either used in a context where
+it is unbound (the variable is {\it free} in $t$) or it is rebound
+(a subterm within $t$ is of form $let x = ... in t'$ where a subterm of $t'$
+is also of form $let x = ... in t''$), then no principal typing exists and
+the phase reports the offending variable.
+
+Then, the set of type constraints is solved, which determines a lower and
+upper bound for each type variable, if such bounds can be determined.
+Additionally, if the constraints lead to a contradiction (i.e. demanding the
+existence of an invalid type like $\boolt \sqcup \unitt$) then no solution
+will be reported.
+
 \subsection{Context constraint solving}
 
-Constraint solving is performed in phases. In the first phase, the context 
-constraints
-are normalised using the rules in \figref{fig:contextsolve} --- where the set of 
-constraints in the premise exists, they are replaced with those in the 
-conclusion. Normalisation means that every context variable will have a single 
+Solving context constraints involves {\it normalising} the constraints. 
+Normalisation means that every context variable will have a single 
 defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
-$\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$, where $\gamma'$ is
+$\gamma = \gamma', \overline{x_i : \tinf_i}$, where $\gamma'$ is
 either empty or free. A context $\Gamma$ for type checking can be easily
@@
 constraints are solved we can say $\Gamma = \overline{x_i : T_i}$ for
-every $\gamma = \gamma', \overline{x_i : \mathbb{T}_i}$ where $T_i$ is the
-solved form of $\mathbb{T}_i$.
+every $\gamma = \gamma', \overline{x_i : \tinf_i}$ where $T_i$ is the
+solved form of $\tinf_i$.
+
+We say that a constraint $C$ is {\it normalisable} if, by
+applying the rules in \figref{fig:contextsolve}, it can be transformed to
+a constraint $C'$ which is normalised or the rules produce $\mathbf{bound}\:x$
+or $\mathbf{unbound}\:x$ for some variable $x$.
 
@@
 in terms of extension of a new root $\gamma_r'$, with $x$ in
-the extension set with the observed type $\mathbb{T}$.
+the extension set with the observed type $\tinf$.
 \item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
@@
 
-\subsection{Variable usage \& type variable constraint solving}
-
-In the second phase, additional type variable 
-constraints are generated by following the rule in \figref{fig:tvargen}, which
-compares the observed type from the usage of a variable to the type
-known in the context. Repeated application of this rule until no more matches
-are found produces a constraint set composed exclusively of normalised
-context variable constraints, type variable constraints and protocol
-constraints. A first order unification algorithm can then be used to solve
-the system of type variable constraints, producing a most general unifier
-for the type variables if unification is possible.
-
-First order unification will identify a set of type variables (and their synonyms)
-which represent distinct object types. For each distinct object type, there
-will also be a set of type variables which represent distinct states for
-the object type. By matching these variables to the protocol constraints
-within the constraint set, we can generate the state machine for each
-object type in a straightforward manner. To produce a minimal description of 
-the object protocol, this state machine can be minimised by constructing a 
-bisimulation relation between the states within each object and collapsing 
-equivalent states.
-
-The typing rules of TS0 do not permit polymorphism, so
-if the most general unifier does not map all type variables for
-variables that occur in the input and output contexts of the term being
-processed, then the term has no valid type in the language. Otherwise,
-substituting the fully explicit type inferred for each type variable that
-occurs in the inferred input and output context variables, as well as the
-type variable for the input term, should produce a valid typing judgement
-where the domain of the input context is the set of free variables in
-the term. This has not yet been proven.
-
 \begin{figure}
@@
 \begin{tabular}{c}
-	\infer[\textsc{ceq}]
-	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
-	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
+	\infer[\textsc{c\_eq}]
+	{\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
+	{\gamma = \gamma' \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
 \\\\
-	\infer[\textsc{crem}]
+	\infer[\textsc{c\_eq\_2}]
+	{\gamma = \gamma', \emptyset}
+	{C \cand \gamma = \gamma' & \free{\gamma'} \in C \vee \gamma' = \emptyset \in C}
+\\\\
+	\infer[\textsc{c\_rem}]
 	{
-	\gamma = \gamma'', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
+	\gamma = \gamma'', \overline{x_i : \tinf_i} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
 	}
 	{
-	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}
+	\gamma = \gamma' - \{x\} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf
 	}
 \\\\
-	\infer[\textsc{cext}]
-	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}}
+	\infer[\textsc{c\_ext}]
+	{\gamma = \gamma'', \overline{x_i : \tinf_i}, \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}}
 	{
-		\gamma = \gamma', \overline{y_j : \mathbb{T}_j} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
+		\gamma = \gamma', \overline{y_j : \tinf_j} \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
 		& \overline{y_j} \cap \overline{x_i} = \emptyset
@@
 \\\\
-	\infer[\textsc{cext\_2}]
+	\infer[\textsc{c\_ext\_2}]
 	{\mathbf{bound}\:x}
 	{
-		\gamma = \gamma', \overline{y_j : \mathbb{T}_j}, x : \mathbb{T} \cand \gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}
+		\gamma = \gamma', \overline{y_j : \tinf_j}, x : \tinf \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}
 		& x \in \overline{x_i}
@@
 \\\\
-	\infer[\textsc{crep}]
-	{\gamma = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T} \cand
-	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
-	{\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
-	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
+	\infer[\textsc{c\_rep}]
+	{\gamma = \gamma'', \overline{x_i : \tinf_i}, x : \tinf \cand
+	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
+	{\gamma = \gamma'[x \mapsto \tinf] \cand
+	\gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf'}
 \\\\
-	\infer[\textsc{crep\_2}]
+	\infer[\textsc{c\_rep\_2}]
 	{
 	\begin{array}{c}
-	\exists \alpha, \gamma_r . 
-	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
-	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
-	\\
-	\gamma'' = \gamma_r, x : \alpha \cand
-	\free{\gamma_r}
+	\exists \alpha . 
+	\gamma = \gamma'[x \mapsto \tinf] \cand
+	\gamma' = \gamma'', \overline{x_i : \tinf_i} \cand
+	\\
+	x : \alpha \in \gamma''
 	\end{array}
-	}
-	{
-	\begin{array}{c}
-	\begin{array}{ccc}
-	x \notin \overline{x_i}
-	&
-	\fresh{\gamma_r}
-	&
-	\fresh{\alpha}
-	\end{array}
-	\\
-	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
-	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
-	\free{\gamma''}
-	\end{array}
-	}
-\\\\
-	\infer[\textsc{crep\_3}]
-	{\mathbf{unbound}\:x}
-	{
-	\begin{array}{c}
-	x \notin \overline{x_i}
-	\\
-	\gamma = \gamma'[x \mapsto \mathbb{T}] \cand
-	\gamma' = \gamma'', \overline{x_i : \mathbb{T}_i} \cand
-	\gamma'' = \emptyset
-	\end{array}
-	}
-\\\\
-	\infer[\textsc{cvar}]
-	{\exists \gamma_r' . 
-	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand
-	\gamma_r = \gamma_r', x : \mathbb{T} \cand
-	\free{\gamma_r'}
 	}
@@
 	&
-	\fresh{\gamma_r'}
+	\fresh{\alpha}
 	\end{array}
 	\\
-	\gamma = \gamma_r, \overline{x_i : \mathbb{T}_i} \cand 
-	x : \mathbb{T} \in \gamma \cand
-	\free{\gamma_r}
+	\gamma = \gamma'[x \mapsto \tinf] \cand
+	\gamma' = \gamma'', \overline{x_i : \tinf_i}
 	\end{array}
@@
 \\\\
+	\infer[\textsc{crep\_3}]
+	{C \cand \gamma = \gamma'[x \mapsto \tinf] \cand 
+	 x : \alpha \in \gamma'}
+	{
+	C \cand \gamma = \gamma'[x \mapsto \tinf]
+	&
+	\free{\gamma'} \in C \vee \gamma' = \emptyset \in C
+	}
+\\\\
+	\infer[\textsc{cvar\_1}]
+	{\exists \gamma' . \free{\gamma'} \cand \gamma = \gamma', x : \tinf}
+	{\free{\gamma} \cand x : \tinf \in \gamma}
+\\\\
 	\infer[\textsc{cvar\_2}]
 	{\mathbf{unbound}\:x}
-	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i} \cand \gamma' = \emptyset
+	{\gamma = \emptyset \cand x : \tinf \in \gamma}
+\\\\
+	\infer[\textsc{cvar\_3}]
+	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma'}
+	{\gamma = \gamma', \overline{x_i : \tinf_i} \cand x : \tinf \in \gamma
 	&
-	x \notin \overline{x_i}}
+	x \notin \overline{x_i}
+	}
+\\\\
+	\infer[\textsc{c\_join}]
+	{
+	\begin{array}{c}
+	\gamma = \overline{x_i : \tinf_i''} \cand 
+	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
+	\\
+	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'} \cand
+	\overline{\tinf_i'' = \tinf_i \sqcup \tinf_i'}
+	\end{array}
+	}
+	{
+	\begin{array}{c}
+	\gamma = \gamma' \sqcup \gamma'' \cand 
+	\gamma' = \gamma_r, \overline{x_i : \tinf_i} \cand 
+	\\
+	\gamma'' = \gamma_r, \overline{x_i : \tinf_i'}
+	\end{array}
+	}
+\\\\
+	\infer[\textsc{cerr\_1}]
+	{\mathbf{unbound}\:x}
+	{C \cand \mathbf{unbound}\:x}
+\\\\
+	\infer[\textsc{cerr\_2}]
+	{\mathbf{bound}\:x}
+	{C \cand \mathbf{bound}\:x}
 \end{tabular}
@@
 
+\subsection{Additional type variable constraint solving}
+
+In the second phase, additional type variable 
+constraints are generated by following the rule in \figref{fig:tvargen}, which
+compares the observed type from the usage of a variable to the type
+known in the context. Repeated application of this rule until no more matches
+are found produces a constraint set composed exclusively of normalised
+context variable constraints, and type variable constraints. At this point, 
+we can focus on the type variable constraints exclusively as the context
+variable constraints have no further contribution to the overall solution.
+
 \begin{figure}
@@
 	\infer[\textsc{tvar}]
-	{\mathbb{T} = \mathbb{T}' \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
-	{x : \mathbb{T} \in \gamma \cand \gamma = \gamma', \overline{x_i : \mathbb{T}_i}, x : \mathbb{T}'}
+	{\tinf = \tinf' \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
+	{x : \tinf \in \gamma \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
 \]
 \caption{\label{fig:tvargen} Refining type variable constraints}
+\end{figure}
+
+\subsection{Type variable constraint solving}
+
+\subsection{Object variable constraint solving}
+
+Normalised object variable constraints are of the forms listed in
+\figref{fig:objconstraints}. Given a set of such normalised
+constraints $C_\omega$, solving proceeds by application of the rules in
+\figref{fig:objsolving}. Where there is no
+explicit partial or fixed solution for a given object variable $\omega$ in
+a constraint set, the solution for the object is equivalent to an empty partial solution, $\mathbf{partial}\:\emptyset$.
+
+\begin{figure}
+\[
+\begin{array}{ll}
+\omega = \partialo{}{} 
+	& \text{partial object solution}
+	\\
+\omega = \fixedo{}{}
+	& \text{fixed object solution} 
+	\\
+\sigma = \mathbf{unknown}
+	& \text{undefined state set}
+	\\
+\sigma = \overline{S} 
+	& \text{state set solution} 
+	\\
+\os{1}{1} = \os{2}{2} 
+	& \text{equality} 
+	\\
+\os{1}{1} <: \os{2}{2} 
+	& \text{subtyping} 
+	\\
+\os{1}{1} = \\
+\:\:\remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}} 
+	& \text{remap definition}
+	\\
+\os{1}{1} = \os{2}{2} \sqcup \os{3}{3}
+	& \text{outer join}
+	\\
+\os{1}{1} = \os{1}{2} \veebar \os{1}{3}
+	& \text{inner join}
+\end{array}
+\]
+\caption{\label{fig:objconstraints} Normalised object variable constraints}
+\end{figure}
+
+\begin{figure*}
+
+\begin{center}
+\begin{tabular}{cc}
+
+\infer[\textsc{$=$-solve}]
+{
+	\substo{\omega}{\omega_1}{
+	\substo{\omega}{\omega_2}{
+	\substo{\sigma}{\sigma_1}{
+	\substo{\sigma}{\sigma_2}{C}}}}, \,
+	\omega = \Omega, \,
+	\sigma = \Sigma
+}
+{
+	\begin{array}{c}
+	C, \, 
+	\os{1}{1} = \os{2}{2}, \,
+	\omega_1 = \Omega_1, \,
+	\omega_2 = \Omega_2, \,
+	\sigma_1 = \Sigma_1, \,
+	\sigma_2 = \Sigma_2
+	\\
+	\sigma = min(\sigma_1, \sigma_2) 
+	\\
+	\omega = min(\omega_1, \omega_2) 
+	\\
+	(\Omega, \Sigma) = \mathbf{connect}(\Omega_1, \Sigma_1, \Omega_2, \Sigma_2)
+	\end{array}
+}
+
+\\
+\\
+
+\infer[\textsc{remap-solve}]
+{
+	C, \,
+	\omega_1 = \partialo{}{'}, \,
+	\sigma_1 = \overline{S}_{out}, \,
+	\sigma_2 = \overline{S}_{in}
+}
+{
+	\begin{array}{c}
+	C, \,
+	\os{1}{1} = \remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}}, \,
+	\omega_1 = \partialo{1}{}, \,
+	\sigma_1 = \Sigma_1, \,
+	\sigma_2 = \Sigma_2
+	\\
+	\omega_2 = \fixedo{2}{} \in C
+	\\
+	\sigma_3 = \overline{S}_3 \in C
+	\\
+	\sigma_4 = \overline{S}_4 \in C
+	\\
+	(\mathbb{S}, \overline{S}_{in}, E) = \mathbf{refine}(\mathbb{S}_1, \Sigma_2, \mathbb{S}_2, \overline{S_3})
+	\\
+	(\mathbb{S}', \Sigma') = \mathbf{connect}(\partialo{}{}, E(\overline{S_4}), \partialo{}{}, \overline{S}_2)
+	\end{array}
+}
+
+\\
+\\
+
+\infer[\textsc{$<:$-solve}]
+{
+	C, \,
+	\omega_1 = \mathbb{S}', \,
+	\sigma_1 = \overline{S}' \,
+}
+{
+	\begin{array}{c}
+	C, \,
+	\os{1}{1} <: \os{2}{2}, \,
+	\omega_1 = \partialo{1}{}, \,
+	\sigma_1 = \Sigma_1
+	\\
+	\omega_2 = \fixedo{2}{} \in C
+	\\
+	\sigma_2 = \overline{S}_2 \in C
+	\\
+	(\mathbb{S}', \overline{S}', E) = \mathbf{overlay}(\mathbb{S}_1, \Sigma_1, \mathbb{S}_2, \overline{S}_2)
+	\end{array}
+}
+
+\\
+\\
+
+\infer[\textsc{$\sqcup$-solve}]
+{
+	C, \,
+	\omega_1 = \Omega',
+	\sigma_1 = \Sigma'
+}
+{
+	\begin{array}{c}
+	C, \,
+	\os{1}{1} = \os{2}{2} \sqcup \os{3}{3} \,
+	\omega_1 = \Omega_1,
+	\sigma_1 = \Sigma_1
+	\\
+	\omega_2 = \fixedo{2} \in C
+	\\
+	\omega_3 = \fixedo{3} \in C
+	\\
+	\sigma_2 = \overline{S}_2 \in C
+	\\
+	\sigma_3 = \overline{S}_3 \in C
+	\\
+	(\mathbb{S}_\cup, \overline{S}'_\cup) = \mathbf{union}(\mathbb{S}_2, \overline{S}_2, \mathbb{S}_3, \overline{S}_3)
+	\\
+	(\Omega', \Sigma') = \mathbf{connect}(\Omega_1, \Sigma_1, \partialo{\cup}{}, \overline{S}_\cup)
+	\end{array}	
+}
+
+\\
+\\
+
+\infer[\textsc{$\veebar$-solve}]
+{
+	C, \,
+	\omega = \Omega, \,
+	\sigma_1 = \Sigma
+}
+{
+	\begin{array}{c}
+	C, \,
+	\os{1}{1} = \os{1}{2} \veebar \os{1}{3}, \,
+	\omega_1 = \Omega_1,
+	\sigma_1 = \Sigma_1
+	\\
+	\sigma_2 = \overline{S}_2 \in C
+	\\
+	\sigma_3 = \overline{S}_3 \in C
+	\\
+	(\Omega, \Sigma) = \mathbf{connect}(\Omega_1, \Sigma_1, \fixedo{2}{}, \overline{S}_2 \cup \overline{S}_3)
+	\end{array}
+}
+
+\end{tabular}
+\end{center}
+\caption{\label{fig:objsolving} Object constraint solving rules}
+\end{figure*}
+
+\begin{figure}
+
+\caption{\label{fig:connectdefn} Definition of $\mathbf{connect}$}
+\end{figure}
+
+\begin{figure}
+
+\caption{\label{fig:overlaydefn} Definition of $\mathbf{overlay}$}
+\end{figure}
+
+\begin{figure}
+
+\caption{\label{fig:overlaydefn} Definition of $\mathbf{union}$}
 \end{figure}
@@
 
-\section{Supporting lemmas}
+\subsection{Supporting lemmas}
 \label{app:proofs}
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% normalisation  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\section{Context constraint normalisation}
+
+For the purposes of the following proofs, existential quantification of
+context variables will be omitted. We shall assume that all introduced
+context variables are globally fresh.
+
+\begin{defn}[Context sub-constraint]
+Given a constraint $C$, the function $ctx(C)$ returns the sub-constraint
+of $C$ composed of only context constraints, i.e. those of class $\mathcal{C}$
+in \figref{fig:congrammar}.
+\end{defn}
+
+\begin{defn}[Context variable definition]
+A context $C$ {\it defines} context variable $\gamma$, written
+$\gamma \in C$, if $C$ contains a sub-constraint $\free{\gamma}$,
+$\gamma = \emptyset$ or $\gamma = \gamma', \overline{x_i : \tinf_i}$.
+\end{defn}
+
+\begin{defn}[Context root]
+Given a normalised constraint $C$ where $\gamma \in C$, the 
+function $root(\gamma, C)$ is defined as follows:
+
+\begin{itemize}
+\item If $\free{\gamma}$ is a sub-constraint of $C$, then $root(\gamma, C) = \gamma$.
+\item If $\gamma = \emptyset$ is a sub-constraint of $C$, then $root(\gamma, C) = \gamma$.
+\item If $\gamma = \gamma', \overline{x_i : \tinf_i}$ is a sub-constraint of $C$, then
+$root(\gamma, C) = \gamma'$.
+\end{itemize}
+\end{defn}
+
+\begin{defn}
+Given a normalized constraint $C$ where $\gamma \in C$,
+the {\it domain} of $\gamma$ (written $dom(\gamma, C)$) is defined as follows:
+
+\begin{itemize}
+\item If $\free{\gamma}$ is a sub-constraint of $C$, 
+then $dom(\gamma, C) = \emptyset$.
+\item If $\gamma = \emptyset$ is a sub-constraint of $C$,
+then $dom(\gamma, C) = \emptyset$.
+\item If $\gamma = \gamma', \overline{x_i : \tinf_i}$ is a sub-constraint of $C$
+then $dom(\gamma, C) = \overline{x_i}$.
+\end{itemize}
+\end{defn}
+
+\begin{defn}
+Given two normalised context constraints $C$ and $C'$ we say that
+$C'$ is {\it stronger} than $C$ (written $C \leq C'$) if:
+
+\begin{itemize}
+\item The domains of the context variables monotonically increase: 
+$\forall \gamma \in C, dom(\gamma, C) \subseteq dom(\gamma, C')$
+\item Root relationships do not change: 
+$\forall \gamma_a, \gamma_b \in C, root(\gamma_a, C) = root(\gamma_b, C) \implies root(\gamma_a, C') = root(\gamma_b, C')$.
+\item The domains of related contexts change consistently:
+$\forall \gamma_a, \gamma_b \in C, root(\gamma_a, C) = root(\gamma_b, C) \cand dom(\gamma_a, C) = dom(\gamma_b, C) \implies dom(\gamma_a, C') = dom(\gamma_b, C')$.
+\end{itemize}
+
+$C \leq C'$ is a partial order.
+
+\end{defn}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% thm:ctx_norm  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{thm}[Normalisation]
+\label{thm:ctx_norm}
+Let $t$ be an arbitrary term. Let $C_t = ctx(\inferlhs{t})$.
+Given a normalised constraint $C$ where $\gamma \in C$ and $\gamma' \notin C$,
+$C \cand C_t$ is normalisable. Where an error does not occur, the normalised
+output $C'$ will have the following properties:
+
+\begin{enumerate}
+\item $C \leq C'$
+\item $dom(\gamma, C') = dom(\gamma', C')$.
+\item $root(\gamma, C') = root(\gamma', C')$.
+\end{enumerate}
+
+\end{thm}
+
+\begin{proof}
+By induction on the structure of t.
+
+\begin{itemize}
+\item $t = unit$. It follows that 
+$C_t \triangleq \gamma' = \gamma$. 
+By \lemref{lem:norm_eq}, $C \cand \gamma' = \gamma$ is normalisable, with
+normalised output $C'$ which has all the required properties.
+
+\item $t = true$. Same as $t = unit$.
+
+\item $t = false$. Same as $t = unit$.
+
+\item $t = \left[ \overline{ S_i \{ \overline{m_ij = (v_ij, S_ij)} \} }  \right]$.
+It follows that 
+$C_t \triangleq 
+\gamma' = \gamma \cand
+\overline{\gamma_{ij} = \emptyset} \cand
+\overline{\inferrule{\gamma_{ij}}{v_ij}{\alpha_ij}{\gamma_{ij}'}}.
+$.
+
+By \lemref{lem:norm_eq} the constraint $C \cand \gamma' = \gamma$ is
+normalisable, with normalised output $C_2$ such that $C \leq C_2$ with
+$dom(\gamma, C_2) = dom(\gamma', C_2)$ and $root(\gamma, C_2) = root(\gamma', C_2)$.
+
+Let $C_3 = C_2 \cand \overline{\gamma_{ij} = \emptyset}$. 
+$C_3$ is normalised by definition, with $C_2 \leq C_3$.
+
+By repeated application of the induction hypothesis, the constraint
+$C_3 \cand \overline{\inferrule{\gamma_{ij}}{v_ij}{\alpha_ij}{\gamma_{ij}'}}$
+is normalisable. If no error is generated, the normalised output $C_4$ will be such that $C_3 \leq C_4$. Transitively, $C \leq C_4$, and the domains and roots
+of $\gamma$ and $\gamma'$ will still be such that $dom(\gamma, C_4) = dom(\gamma', C_4)$ and $root(\gamma, C_4) = root(\gamma', C_4)$.
+
+\item $t = \\(\overline{x_i : T_i \gg T_i'}).t'$. 
+It follows that 
+$C_t \triangleq 
+\gamma' = \gamma \cand \gamma_r = \emptyset \cand 
+\gamma_in = \gamma_r, \overline{x_i : \alpha_i} \cand
+\overline{x_i : \alpha_i' \in \gamma_out} \cand ctx(\inferrule{\gamma_{in}}{t'}{\alpha'}{\gamma_{out}})$.
+
+By \lemref{lem:norm_eq}, the constraint $C \cand \gamma' = \gamma$
+is normalisable. If no error is generated, the normalised output $C_2$ will be such that $C \leq C_2$ with $dom(\gamma, C_2) = dom(\gamma', C_2)$ and
+$root(\gamma, C_2) = root(\gamma', C_2)$.
+
+Let $C_3$ be the constraint 
+$C_2 \cand \gamma_r = \emptyset \cand \gamma_{in} = \gamma_r, \overline{x_i : \alpha_i}$.
+$C_3$ is normalised by definition, with $C_2 \leq C_3$.
+
+The constraint $C_3 \cand \inferrule{\gamma_{in}}{t'}{\alpha}{\gamma_{out}}$
+is normalisable by induction. Let $C_4$ be the normalised output, where
+$C_3 \leq C_4$.
+
+Let $C_5$ be the constraint 
+$C_4 \cand \overline{x_i : \alpha_i' \in \gamma_out}$.
+This constraint is normalisable by repeated application of 
+\lemref{lem:norm_usage}. If no error is generated, the normalised constraint 
+$C_5$ will be produced, where $C_4 \leq C_5$.
+
+Transitively, $C \leq C_5$. It is also the case that $C_2 \leq C_5$, therefore 
+$dom(\gamma, C_5) = dom(\gamma', C_5)$ and
+$root(\gamma, C_5) = root(\gamma', C_5)$.
+
+\item $t = \lett{x}{t'}{t''}$.
+It follows that 
+$C_t \triangleq \gamma_{in} = \gamma_v, x : \alpha' \cand 
+\gamma' = \gamma_{out} - \{x\} \cand
+ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}) \cand
+ctx(\inferrule{\gamma_{in}}{t''}{\alpha}{\gamma_{out}})$.
+
+By induction, the constraint 
+$C \cand ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma_v})$ is normalisable.
+If no error is generated, the normalised output $C_2$ will be such that
+$C \leq C_2$ with $dom(\gamma, C_2) = dom(\gamma_v, C_2)$ and
+$root(\gamma, C_2) = root(\gamma_v, C_2)$.
+
+The constraint $C_2 \cand \gamma_{in} = \gamma_v, x : \alpha'$ is
+normalisable by \lemref{lem:norm_ext}. If no error is generated, the
+normalised output $C_3$ will be such that $C_2 \leq C_3$ with 
+$dom(\gamma_{in}, C_3) = dom(\gamma_v, C_2) + \{ x \}$.
+
+By induction, the constraint
+$C_3 \cand ctx(\inferrule{\gamma_{in}}{t''}{\alpha}{\gamma_{out}})$
+is normalisable. If no error is generated, the normalised output $C_4$
+will be such that
+$C_3 \leq C_4$ with $dom(\gamma_{in}, C_4) = dom(\gamma_{out}, C_4)$
+and $root(\gamma_{in}, C_4) = root(\gamma_{out}, C_4)$.
+
+As $x \in dom(\gamma_{in}, C_3)$, it must therefore be the case that
+$x \in dom(\gamma_{in}, C_4)$, and that $x \in dom(\gamma_{out}, C_4)$.
+
+By rule $\textsc{c\_rem}$, the constraint
+$C_4 \cand \gamma' = \gamma_{out} - \{ x \}$ can be transformed
+to $C_5 = C_4 \cand \gamma' = \gamma_r, \overline{x_i : \tinf_i}$, where
+$\gamma_r = root(\gamma_{out}, C_4)$, and 
+$\overline{x_i} = dom(\gamma_{out}, C_4)$. $C_5$ is normalised by definition,
+with $C_4 \leq C_5$.
+
+Transitively, $C \leq C_5$.
+
+% relate dom(\gamma) and dom(\gamma')
+% we currently preserve equality relationships between domains,
+% can we preserve equality plus variable set?
+
+\item $t = t' ; t''$.
+It follows that
+$C_t \triangleq 
+ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma''}) \cand
+ctx(\inferrule{\gamma''}{t''}{\alpha}{\gamma'})$.
+
+By induction, the constraint 
+$C \cand ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma''})$ is normalisable.
+If no error is generated, the normalised output $C_2$ will be such that
+$C \leq C_2$ with $dom(\gamma, C_2) = dom(\gamma'', C_2)$ and
+$root(\gamma, C_2) = root(\gamma'', C_2)$.
+
+By induction, the constraint
+$C_2 \cand ctx(\inferrule{\gamma''}{t''}{\alpha'}{\gamma'})$ is normalisable.
+If no error is generated, the normalised output $C_3$ will be such that
+$C_2 \leq C_3$ with $dom(\gamma'', C_3) = dom(\gamma', C_3)$ and
+$root(\gamma'', C_3) = root(\gamma', C_3)$.
+
+As $C_2 \leq C_3$, it also follows that $dom(\gamma, C_3) = dom(\gamma'', C_3)$
+and $root(\gamma, C_3) = root(\gamma'', C_3)$. Therefore,
+$dom(\gamma, C_3) = dom(\gamma', C_3)$ and 
+$root(\gamma, C_3) = root(\gamma', C_3)$. Transitively, $C \leq C_3$.
+
+\item $t = x(\overline{x_i})$.
+It follows that
+$C_t \triangleq \overline{x_i : \alpha_i \in \gamma} \cand 
+x : \tinf_x \cand
+\gamma' = \gamma[\overline{x_i \mapsto \tinf_i}]
+$.
+
+The constraint $C \cand x : \alpha_x \in \gamma$ is normalisable by 
+\lemref{lem:norm_usage}. If no error is generated, the normalised output $C_2$ will be such that $C \leq C_2$ with $x \in dom(\gamma, C_2)$.
+
+The constraint $C_2 \cand \overline{x_i : \alpha_i \in \gamma}$ is normalisable
+by repeated application of \lemref{lem:norm_usage}. If no error is generated, the normalised output $C_3$ will be such that $C_2 \leq C_3$ with
+$x_i \in dom(\gamma, C_3)$, for all $i$.
+
+The constraint $C_3 \cand \gamma' = \gamma[\overline{x_i \mapsto \remapapp{\alpha_i}{\alpha_i'}{\alpha_i''}}]$
+is normalisable by \lemref{lem:norm_rep}. If no error is generated, the
+normalised output $C_4$ will be such that $C_3 \leq C_4$ with
+$dom(\gamma, C_4) = dom(\gamma', C_4)$ and $root(\gamma, C_4) = root(\gamma', C_4)$. Transitively, $C \leq C_4$.
+
+\item $t = x.m$.
+It follows that
+$C_t \triangleq x : \alpha_o@\alpha_s \in \gamma \cand
+\gamma' = \gamma[x \mapsto \alpha_x]
+$.
+
+The constraint $C \cand x : \alpha_o@\alpha_s \in \gamma$ is normalisable
+by \lemref{lem:norm_usage}. If no error is generated, the normalised output
+$C_2$ will be such that $C \leq C_2$ with $x \in dom(\gamma, C_2)$.
+
+The constraint $C_2 \cand \gamma' = \gamma[x \mapsto \alpha_x]$ is normalisable
+by \lemref{lem:norm_rep}. If no error is generated, the normalised output $C_3$ will be such that $C_2 \leq C_3$ with
+$dom(\gamma', C_3) = dom(\gamma, C_3)$ and $root(\gamma', C_3) = root(\gamma, C_3)$. Transitively, $C \leq C_3$.
+
+\item $t = \ift{t_c}{t_t}{t_f}$.
+It follows that
+$C_t \triangleq
+ctx(\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}) \cand
+ctx(\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t}) \cand
+ctx(\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f}) \cand
+\gamma' = \gamma_t \sqcup \gamma_f
+$.
+
+By induction, the constraint
+$C \cand ctx(\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c})$ is normalisable.
+If no error is produced, the normalised output $C_2$ will be such that
+$C \leq C_2$ with $dom(\gamma, C_2) = dom(\gamma_c, C_2)$ and
+$root(\gamma, C_2) = root(\gamma_c, C_2)$.
+
+By induction, the constraint
+$C_2 \cand ctx(\inferrule{\gamma_c}{t_t}{\alpha_t}{\gamma_t})$ is normalisable.
+If no error is produced, the normalised output $C_3$ will be such that
+$C_2 \leq C_3$ with $dom(\gamma_c, C_3) = dom(\gamma_t, C_3)$ and
+$root(\gamma_c, C_3) = root(\gamma_t, C_3)$.
+
+By induction, the constraint
+$C_3 \cand ctx(\inferrule{\gamma_c}{t_f}{\alpha_f}{\gamma_f})$ is normalisable.
+If no error is produced, the normalised output $C_4$ will be such that
+$C_3 \leq C_4$ with $dom(\gamma_c, C_4) = dom(\gamma_f, C_4)$ and
+$root(\gamma_c, C_4) = root(\gamma_f, C_4)$.
+
+It follows that as $C \leq C_2 \leq C_3 \leq C_4$, that the domains
+of $\gamma$, $\gamma_c$, $\gamma_t$ and $\gamma_f$ are all equal in $C_4$,
+as are their roots.
+
+It follows by rule $\textsc{c\_join}$ that the constraint
+$C_4 \cand \gamma' = \gamma_t \sqcup \gamma_f$ can be transformed to
+$C_5 = C_4 \cand \gamma' = \gamma_r, \overline{x_i : T_i} \cand \overline{T_i = T_t \veeq T_f}$ 
+where $\overline{x_i} = dom(\gamma, C_4)$. Therefore $dom(\gamma, C_4) = dom(\gamma', C_4)$ and $root(\gamma, C_4) = root(\gamma', C_4)$. Trivially, $C_4 \leq C_5$ and transitively $C \leq C_5$.
+
+\end{itemize}
+\end{proof}
+
+\begin{cor}
+Given an arbitrary root term $t$, the generated constraint set for $t$,
+which is defined as
+$\exists \gamma, \gamma', \alpha . \free{\gamma} . \inferlhs{t}$,
+is normalisable.
+\end{cor}
+
+\subsection{Supporting Lemmas}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:norm_eq  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}[Equality normalisation]
+\label{lem:norm_eq}
+Given a normalised constraint $C$ where $\gamma \in C$ and $\gamma' \notin C$, 
+the constraint $C \cand \gamma' = \gamma$ is normalisable. Where an
+error is not generated, the normalised output $C'$ will have the following
+properties:
+
+\begin{enumerate}
+\item $dom(\gamma', C') = dom(\gamma, C')$
+\item $root(\gamma', C') = root(\gamma, C')$
+\item $C \leq C'$.
+\item $\forall \gamma'' \in C, dom(\gamma'', C'') = dom(\gamma'', C)$
+\end{enumerate}
+
+\end{lem}
+\begin{proof}
+There are two cases to consider:
+
+\begin{itemize}
+\item The sub-constraint $\free{\gamma}$ or $\gamma = \emptyset$ exists
+in $C$. By rule $\textsc{c\_eq\_2}$ the constraint $C \cand \gamma' = \gamma$
+will be transformed to $C_2 = C \cand \gamma' = \gamma, \emptyset$, which is
+normalised and has all the required properties.
+
+\item The sub-constraint $\gamma = \gamma'', \overline{x_i : \tinf_i}$ exists
+in $C$. By rule $\textsc{c\_eq}$ the constraint $C \cand \gamma' = \gamma$
+will be transformed to 
+$C_2 = C \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}$,
+which is normalised and has all the required properties.
+\end{itemize}
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:norm_ext  %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}[Extension normalisation]
+\label{lem:norm_ext}
+Given a normalised constraint $C$ where $\gamma \in C$ and $\gamma' \notin C$, 
+the constraint $C' = C \cand \gamma' = \gamma, \overline{x_i : \tinf_i}$
+is normalisable. When an error is not generated, the normalised output
+$C''$ will have the following properties:
+
+\begin{enumerate}
+\item $dom(\gamma', C'') = dom(\gamma, C'') + \overline{x_i}$
+\item $root(\gamma', C'') = root(\gamma, C'')$
+\item $C \leq C''$.
+\item $\forall \gamma'' \in C, dom(\gamma'', C'') = dom(\gamma'', C)$
+\end{enumerate}
+
+\end{lem}
+\begin{proof}
+There are two cases to consider:
+
+\begin{enumerate}
+\item The sub-constraint $\free{\gamma}$ or $\gamma = \emptyset$ exists in $C$.
+The constraint $C'$ is therefore already normalised, meaning $C'' = C'$.
+$C''$ has all the required properties.
+
+\item The sub-constraint $\gamma = \gamma_r, \overline{x_j : \tinf_j}$ exists
+in $C$, along with either $\free{\gamma_r}$ or $\gamma_r = \emptyset$ (as
+$C$ is normalised).
+
+If $\overline{x_i}$ and $\overline{x_j}$ are disjoint, then by rule
+$\textsc{c\_ext}$ the constraint $C'$ can be transformed to
+$C'' = C \cand \gamma' = \gamma_r, \overline{x_i : \tinf_i}, \overline{x_j : \tinf_j}$.
+$C''$ is therefore normalised, with all the required properties.
+
+If $\overline{x_i} \cap \overline{x_j} \neq \emptyset$, then by rule
+$\textsc{c\_ext\_2}$ a $\mathbf{bound}\:x$ error will be generated.
+\end{enumerate}
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:norm_rep %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}[Replacement normalisation]
+\label{lem:norm_rep}
+Given a normalised constraint $C$ where $\gamma \in C$ and $\gamma' \notin C$,
+the constraint $C' = C \cand \gamma' = \gamma[x \mapsto \tinf]$ is normalisable. When
+an error is not generated, the normalised output $C''$ will have the
+following properties:
+
+\begin{enumerate}
+\item $dom(\gamma', C'') = dom(\gamma, C'')$
+\item $root(\gamma', C'') = root(\gamma, C'')$
+\item $C \leq C''$.
+\end{enumerate}
+
+\end{lem}
+\begin{proof}
+
+There are three cases to consider:
+
+\begin{enumerate}
+
+\item The sub-constraint $\free{\gamma}$ or $\gamma = \emptyset$ exists in
+$C$. By rule $\textsc{c\_rep\_3}$, the constraint $C'$ can be transformed
+to 
+$C_2 = C \cand \gamma' = \gamma[x \mapsto \tinf] \cand x : \alpha \in \gamma$.
+
+By \lemref{lem:norm_usage}, the constraint $C \cand x : \alpha \in \gamma$ is 
+normalisable. If no
+error is generated, the normalised constraint $C_3$ will be produced such
+that $C \leq C_3$ and $x \in dom(\gamma, C_3$. Therefore, by the 
+reasoning of case 3 below, $C_3 \cand \gamma' = \gamma[x \mapsto \tinf]$ is 
+normalisable. If no error is generated, the normalised constraint $C''$ will 
+be generated with all the required properties.
+
+\item The sub-constraint $\gamma = \gamma'', \overline{x_i : \tinf_i}$ exists
+in $C$, where $x \notin \overline{x_i}$.
+
+By rule $\textsc{c\_rep\_2}$, the
+constraint $C'$ can be transformed to $C_2 = C' \cand x : \alpha \in \gamma''$.
+Let $C_3 = C \cand x : \alpha \in \gamma''$. It follows that
+$C_2$ is equivalent to $C_3 \cand \gamma' = \gamma[x \mapsto \tinf]$. 
+By \lemref{lem:norm_usage}, $C_3$ is normalisable. If no
+error is generated, the normalised constraint $C_4$ will be produced such that
+$C \leq C_4$, and $x \in dom(\gamma'', C_4)$.
+
+The constraint $C_4 \cand \gamma' = \gamma[x \mapsto \tinf]$ is normalisable
+by the reasoning of case 3 below. 
+If no error occurs, normalised constraint $C''$ will be produced satisfying
+properties 1 and 2 and with $C_4 \leq C''$. Transitively, $C \leq C''$.
+
+\item The sub-constraint 
+$\gamma = \gamma'', \overline{x_i : \tinf_i}, x : \tinf_x$
+exists in $C$. It follows by rule $\textsc{c\_rep}$ that constraint $C'$ can 
+be transformed to 
+$C'' = C \cand \gamma' = \gamma'', \overline{x_i : \tinf_i}, x : \tinf$, 
+which is normalised and has all the required properties.
+
+\end{enumerate}
+
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:norm_usage %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}[Usage normalisation]
+\label{lem:norm_usage}
+Given a normalised constraint $C$ where $\gamma \in C$, the constraint 
+$C \cand x : \tinf \in \gamma$ is normalisable. When an error is not generated, 
+the constraint $C_2$ will be generated such that $C \leq C_2$ and
+$x \in dom(\gamma, C_2)$.
+\end{lem}
+\begin{proof}
+
+There are three cases to consider:
+
+\begin{enumerate}
+\item $C = C_2 \cand \free{\gamma}$.
+
+By rule $\textsc{c\_var\_1}$, the constraint
+$\free{\gamma} \cand x : \tinf \in \gamma$ can be transformed to 
+$C_3 = \free{\gamma_r} \cand \gamma = \gamma_r, x : \tinf$ where
+$\gamma_r$ is a fresh context variable.
+
+It remains to show that $C_4 = C_2 \cand C_3$ is normalisable.
+
+Let $\overline{\gamma_i}$ be the set of context variables defined in
+$C_2$ that have $\gamma$ as their root: for all $i$, $\gamma_i = \gamma, \overline{x_{ij}}$ is
+a sub-constraint of $C_2$ (as it was previously normalised).
+
+\begin{itemize}
+\item If $\overline{\gamma_i}$ is empty, then $C_4$ is normalised.
+It can be directly observed that $x \in dom(\gamma, C_4)$ and that
+$C \leq C_4$.
+
+\item If $\overline{\gamma_i}$ is not empty, then
+$C_2 = C_5 \cand \overline{\gamma_i = \gamma, \overline{x_{ij}}}$.
+
+If $x \in \overline{x_{ij}}$ for some $i$, then by rule $\textsc{c\_ext\_2}$
+the error $\mathbf{bound}\:x$ will be generated.
+
+Otherwise, by repeated application of rule $\textsc{c\_ext}$ the constraint
+$C_4$ can be transformed to $C_6 = C_3 \cand C_5 \cand \overline{\gamma_i = \gamma, \overline{x_{ij}}, x : \tinf}$. For every $\gamma' \in C$, 
+the following predicates hold:
+\begin{itemize}
+\item $root(\gamma', C) = \gamma \implies root(\gamma', C_6) = \gamma_r \cand dom(\gamma', C_6) = dom(\gamma', C_6) + \{x\}$
+\item $root(\gamma', C) \neq \gamma \implies root(\gamma', C_6) = root(\gamma', C) \cand dom(\gamma', C_6) = dom(\gamma', C)$
+\end{itemize}
+
+Therefore, $C \leq C_6$, and $x \in dom(\gamma, C_6)$.
+\end{itemize}
+
+\item $C = C_2 \cand \gamma = \emptyset$, where $C_2$ is normalised.
+
+By rule $\textsc{c\_var\_2}$, $\exists \gamma . C \cand x : \tinf \in \gamma$
+will be transformed to $\mathbf{unbound}\:x$.
+
+\item $C = C_2 \cand \gamma = \gamma_r, \overline{x_i : \tinf_i}$, where 
+$C_2$ is normalised.
+
+If $x = x_i$ for some $i$, then by rule $\textsc{tvar}$ the constraint
+$C \cand x : \tinf \in \gamma$ will be transformed to $C' = C \cand \tinf_i <: \tinf$. As
+the type constraint is irrelevant to normalisation, the tranformed constraint
+is normalised, and $C \leq C'$. Trivially, $x \in dom(\gamma, C')$.
+
+If $x \notin \overline{x_i}$, then by rule $\textsc{c\_var\_3}$ the
+constraint $C \cand x : \tinf \in \gamma$ will be transformed to
+$C \cand \overline{x_i : \tinf_i} \in \gamma_r$. As
+$\gamma_r$ must be defined in $C$ by a constraint $\free{\gamma_r}$ or
+$\gamma_r = \emptyset$, the reasoning of parts 1 and 2 above can be applied.
+Where an error is not generated, the normalised constraint $C'$ will be
+generated such that $C \leq C'$ and $x \in dom(\gamma, C')$.
+
+\end{enumerate}
+\end{proof}
+
 \bibliographystyle{abbrvnat}

### 2012-11-30 formalism/paper.tex
minor clean-ups of proofs of context normalisation, some initial descriptive work of the object constraint solving process.

@@
 \newcommand{\appref}[1]{Appendix \ref{#1}}
+\newcommand{\thmref}[1]{Theorem \ref{#1}}
 \newcommand{\lemref}[1]{Lemma \ref{#1}}
@@
 % unit value
-\inferlhs{\unitv} \triangleq \alpha = \unitt \cand \gamma' = \gamma
+\inferlhs{\unitv} \coloneqq \alpha = \unitt \cand \gamma' = \gamma
 \\\\
@@
 % true value
-\inferlhs{\truev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
+\inferlhs{\truev} \coloneqq \alpha = \boolt \cand \gamma' = \gamma 
 \\\\
@@
 % false value
-\inferlhs{\falsev} \triangleq \alpha = \boolt \cand \gamma' = \gamma 
+\inferlhs{\falsev} \coloneqq \alpha = \boolt \cand \gamma' = \gamma 
 \\\\
@@
 \inferlhs{[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij}, S_{ij}) } \} }]@\overline{S}}
-\triangleq \\
+\coloneqq \\
 \exists \alpha_o, \alpha_s, \overline{\alpha_{ij}}, \gamma_{\emptyset} . \\
@@
 % function value
-\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \triangleq \\
+\inferlhs{\lambda ( \overline{x_i : T_i \gg T_i'} ) . t} \coloneqq \\
 \exists \gamma_r, \gamma_{in}, \gamma_{out}, \overline{\alpha_i, \alpha_i'}, \alpha_r . \\
@@
 % let binding
-\inferlhs{\lett{x}{t'}{t}} \triangleq \\
+\inferlhs{\lett{x}{t'}{t}} \coloneqq \\
 \exists \gamma_v, \gamma_{in}, \gamma_{out}, \alpha' . \\
@@
 % sequence
-\inferlhs{t ; t'} \triangleq \\
+\inferlhs{t ; t'} \coloneqq \\
 \exists \gamma'', \alpha' . \\
@@
 % function call
-\inferlhs{x ( \overline{x_i} )} \triangleq \\
+\inferlhs{x ( \overline{x_i} )} \coloneqq \\
 \exists \overline{\alpha_i}, \overline{\alpha_i'}, \overline{\alpha_i''} . \\
@@
 % method call
-\inferlhs{x.m} \triangleq \\
+\inferlhs{x.m} \coloneqq \\
 \exists \alpha_o, \alpha_s . \\
@@
 % if-then-else
-\inferlhs{\ift{t_c}{t_t}{t_f}} \triangleq \\
+\inferlhs{\ift{t_c}{t_t}{t_f}} \coloneqq \\
 \exists \alpha_c, \alpha_t, \alpha_f, \gamma_c, \gamma_t, \gamma_f . \\
@@
 
-Additionally, we have {\bf variable usage constraints} of form 
+Additionally, we have {\it variable usage constraints} of form 
 $x : \tinf \in \gamma$, which are generated for each location in the
@@
 \item {\bf Type expressions constraints} --- these express 
-constraints between two or three types: either two types are equal,
-or one type is the join of two others. The latter form of type constraint
-is generated when dealing with if-then-else terms.
+relationships between types. Either two types are equal ($\tinf = \tinf'$),
+one type is a subtype of another ($\tinf <: \tinf'$) or one type
+is the join of two others ($\tinf \sqcup \tinf'$). As a special case of
+the latter, if it is known that the two types are derived from
+interactions with the same value, this is written as $\tinf \veebar \tinf'$.
+This additional information is relevant when solving object type constraints
+in particular, and is discussed later.
 \end{itemize}
@@
 is expanded using the rules shown in \figref{fig:constraintgen}.
-The constraint generation process for a root term $t$ begins with
-$\exists \gamma, \gamma', \alpha. \free{\gamma} \cand \inferlhs{t}$.
+The constraint generation process for a root term is defined
+by the function $gen(t) = \exists \gamma, \gamma', \alpha. \free{\gamma} \cand \inferlhs{t}$.
 
@@
 not change the context. Additionally, a type variable constraint of form
-$\alpha = \tinf$ where $\tinf$ is of the appropriate type to match
-the value, with type variables in all positions relating to parameters and return
-values.
-\item The context variable $\gamma_a$ used as the input context for
-function bodies has an empty root $\gamma_r$, to allow for detection of 
+$\alpha = \tinf$ is generated where $\tinf$ is of the appropriate type to 
+match the value. In the case of functions or object values, this type will
+contain additional type variables for parameters and return values, with
+the intention that these be solved to validate user type annotations.
+\item The context variable $\gamma_{in}$ used as the input context for
+function bodies has an empty root $\gamma_r$. This allows for the detection of 
 used unbound variables within
-the body. This matches the use of a context in type inference
-composed only of the parameters, and prevents implicit binding.
+the body term. The domain of the input context is strictly composed of the 
+parameters, preventing implicit binding, in order to match the properties
+of the typing judgement for functions.
 \item Let-bind terms generates a new context variable $\gamma_v$ for the
-output context of the value term. This is extended by $\gamma_{b1}$, the
+output context of the value term. This is extended by $\gamma_{in}$, the
 input context for the body term, with variable $x$ of the type inferred
-for the value. The output context of the whole let-bind term removes
-$x$ from the output context $\gamma_{b2}$ of the body term.
-\item Update terms generate replacement constraints based on the inferred
-type $\alpha'$ of the right-hand side of the expression.
+for the value. The output context $\gamma'$ of the whole let-bind term removes
+$x$ from the output context $\gamma_{out}$ of the body term.
 \item Sequencing terms relate the output of the left term to the input of the
 right term through a new context variable $\gamma''$. The type variable
-$\alpha'$ must be generated to capture the type of the left term for
-consistency, but it is not needed.
+$\alpha'$ is generated to infer the type of the left term, though this is not needed beyond confirming that it does in fact have a type.
 \item Function call terms generate context variable constraints to relate the
@@
 and that after the method call $x$ must be in the new state $\alpha_s'$.
+\item If-then-else terms relate the context variables used for the
+condition, true branch and false branch as expected. It is asserted that the
+type of the condition must be a boolean. The output context
+variable $\gamma'$ is expressed as the join of the outputs for the true
+and false branches.
 \end{itemize}
@@
 Constraint solving is performed in phases. First, generated context constraints
-for a term $t$ aresolved, which determines the contents of each context associated to
+for a term $t$ are solved, which determines the contents of each context associated to
 each context variable in the typing derivation of $t$, and produces additional
@@
 Solving context constraints involves {\it normalising} the constraints.
-Normalisation means that every context variable will have a single 
+A {\it normalised} constraint is one in which
+context variable has a single 
 defining constraint of form $\gamma = \emptyset$, $\free{\gamma}$ or
@@
 every $\gamma = \gamma', \overline{x_i : \tinf_i}$ where $T_i$ is the
-solved form of $\tinf_i$.
+solved form of $\tinf_i$, or $\Gamma = \emptyset$ for every $\free{\gamma}$
+or $\gamma = \emptyset$.
 
@@
 applying the rules in \figref{fig:contextsolve}, it can be transformed to
-a constraint $C'$ which is normalised or the rules produce $\mathbf{bound}\:x$
+a constraint $C'$ which is normalised, or the rules produce $\mathbf{bound}\:x$
 or $\mathbf{unbound}\:x$ for some variable $x$.
-
-The context normalisation rules do the following:
-
-\begin{itemize}
-\item \textsc{ceq} --- Where $\gamma$ is directly equivalent
-to $\gamma'$, which is defined in terms of extension of
-$\gamma''$, rewrite the constraints such that
-$\gamma$ directly copies the extension of $\gamma''$.
-\item \textsc{crem} --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
-which is defined in terms of extension of $\gamma''$, rewrite the
-constraints such that $\gamma$ copies the extension of $\gamma''$ with
-the variable $x$ removed from the extension. There is no need for an
-error-producing variant of \textsc{crem}, where the variable $x$ is missing, 
-as context removal constraints are always generated in a pair with a context 
-extension constraint for let binding.
-\item \textsc{cext} --- Where $\gamma$ extends $\gamma'$ with
-a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
-$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
-constraints such that $\gamma$'s extension set includes both the $y_j$
-definitions and $x_j$ definitions.
-\item $\textsc{cext}_2$ --- Where $\gamma$ extends $\gamma'$ with
-a new variable $x$, and $\gamma'$ is defined in terms of extension of
-some $\gamma''$ and the extension set already contains $x$, this
-represents an attempt to redeclare $x$. As such, generate an error
-$\mathbf{bound}\:x$, terminating the constraint solving process.
-\item \textsc{crep} --- where $\gamma$ replaces the type of some
-variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
-of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
-such that $\gamma$ copies the extension set of $\gamma'$ with $x$
-replaced as specified.
-\item $\textsc{crep}_2$ --- where $\gamma$ replaces variable $x$ in
-$\gamma'$ which is defined in terms of extension of $\gamma''$
-which is free and $x$ is not in the extension set, $x$ is a
-free variable in the term being processed. To introduce $x$ in such a way that
-it will be included in all contexts, $\gamma''$ is redefined in
-terms of extension of a new context $\gamma_r$ with variable $x$
-in the extension set mapped to a new type variable $\alpha$. Following
-this, \textsc{cext} can be applied to $\gamma'$, then \textsc{crep} can
-be applied to $\gamma$.
-\item $\textsc{crep}_3$ --- where $\gamma$ replaces variable $x$ in
-$\gamma'$ which is defined in terms of an empty root $\gamma''$,
-and $x$ is not in $\gamma$'s extension set, this indicates the usage
-of an unbound variable within a function body. As such, generate an
-error $\mathbf{unbound}\:x$ indicating that x is unbound within this
-context, terminating the contraint solving process.
-\item \textsc{cvar} --- where a variable constraint for $x$ exists
-for context variable $\gamma$, which is defined in terms of
-extension of $\gamma_r$ which is free, and $x$ is not in the extension
-set of $\gamma$, $x$ is a free variable in the term being processed. 
-Similar to $\textsc{crep}_2$, redefine $\gamma_r$
-in terms of extension of a new root $\gamma_r'$, with $x$ in
-the extension set with the observed type $\tinf$.
-\item $\textsc{cvar}_2$ --- where a variable constraint for $x$ exists
-for context variable $\gamma$, which is defined in terms of $\gamma'$
-which is empty, and $x$ is not in the extension set of $\gamma$, 
-$x$ is an unbound variable within a function body.
-\end{itemize}
-
-The rules are repeatedly applied to any matching sets of constraints
-until no further rewriting is possible, producing either an error
-or a set of context constraints which are fully normalised, and
-can be converted into contexts for type checking easily.
-
-The way in which the context constraints are generated guarantees
-that the dependencies between contexts form a directed acyclic graph. For
-this reason, we believe that that the first phase will always terminate,
-but this has not yet been proven.
 
@@
 \\\\
+\infer[\textsc{t\_var}]
+	{
+	\tinf' <: \tinf \cand 
+	\gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'
+	}
+	{
+	x : \tinf \in \gamma \cand 
+	\gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
+\\\\
 	\infer[\textsc{c\_join}]
@@
 
+The context normalisation rules do the following:
+
+\begin{itemize}
+\item $\textsc{c\_eq}$ --- Where $\gamma$ is directly equivalent
+to $\gamma'$, which is defined in terms of extension of
+$\gamma''$, rewrite the constraints such that
+$\gamma$ directly copies the definition of $\gamma'$.
+\item $\textsc{c\_eq\_2}$ --- Where $\gamma$ is directly equivalent
+to $\gamma'$, which is defined as either $\free{\gamma}$ or 
+$\gamma = \emptyset$, $\gamma'$ is defined to be an empty extension of 
+$\gamma'$.
+
+\item $\textsc{c\_rem}$ --- Where $\gamma$ removes a variable $x$ from $\gamma'$,
+which is defined in terms of extension of $\gamma''$ that is known to have
+$x$ in its domain, rewrite the
+constraints such that $\gamma$ copies the definition $\gamma''$ with
+the variable $x$ removed from the extension. There is no need for an
+error-producing variant of \textsc{c\_rem}, where the variable $x$ is missing, 
+as context removal constraints are always generated in a pair with a context 
+extension constraint for let binding.
+
+\item $\textsc{c\_ext}$ --- Where $\gamma$ extends $\gamma'$ with
+a variable set $\overline{y_j}$, and $\gamma'$ is defined in terms of extension of
+$\gamma''$ with a disjoint set of variables in its extension set, rewrite the
+constraint such that $\gamma$'s extension set includes both the $y_j$
+definitions and $x_j$ definitions.
+
+\item $\textsc{c\_ext}_2$ --- Where $\gamma$ extends $\gamma'$ with
+a new variable $x$, and $\gamma'$ is defined in terms of extension of
+some $\gamma''$ and the extension set already contains $x$, this
+represents an attempt to redeclare $x$. As such, generate an error
+$\mathbf{bound}\:x$, terminating the constraint solving process.
+
+\item $\textsc{c\_rep}$ --- where $\gamma$ replaces the type of some
+variable $x$ in $\gamma'$, where $\gamma$ is defined in terms of extension
+of $\gamma''$ and $x$ is in the extension set, rewrite the constraints
+such that $\gamma$ copies the extension set of $\gamma'$ with $x$
+replaced as specified.
+
+\item $\textsc{c\_rep}_2$ --- where $\gamma$ replaces variable $x$ in
+$\gamma'$ which is defined in terms of extension of $\gamma''$
+which is free and $x$ is not in the extension set, $x$ is a
+free variable in the term being processed. To introduce $x$ in such a way that
+it will be included in all contexts, a variable usage constraint
+$x : \alpha \in \gamma''$ is introduced, with the intention that this be
+solved by $c\_var\_1$ or $c\_var\_2$ first before the replacement can be
+solved using $c\_rep\_1$.
+
+\item $\textsc{crep}_3$ --- this performs the same function as
+$c\_rep\_2$, but for the case where $\gamma$ is directly defined as
+a replacement on a free or empty context variable $\gamma'$.
+
+\item $\textsc{c\_var\_1}$ --- where $x$ is asserted to exist in a context
+variable $\gamma$, which is defined as $\free{\gamma}$, the definition
+of $\gamma$ is replaced with one which extends a new free context variable
+$\gamma'$ with the variable $x$. This is intended to also have the effect
+that all other context variables which are defined as extensions of 
+$\gamma$ will have to be updated using \textsc{c\_ext}, or an error
+generated by \textsc{c\_ext} if they already contain $x$.
+
+\item $\textsc{c\_var\_2}$ --- where $x$ is asserted to exist in a
+context variable $\gamma$, which is defined as $\gamma = \emptyset$, this
+represents the attempt to use an unbound variable in a context which does
+not permit free variables (such as in the body of a function). As such,
+an $\mathbf{unbound}\:x$ error is generated.
+
+\item $\textsc{c\_var\_3}$ --- where $x$ is asserted to exist
+in a context variable $\gamma$, which is defined in terms of
+extension of $\gamma'$ which does not include $x$, the assertion
+can be rewritten as an assertion that $x$ must exist within $\gamma'$.
+This is intended to propagate to a root context variable and solved
+by \textsc{c\_var\_1} or \textsc{c\_var\_2}.
+
+\item $\textsc{c\_join}$ --- where a context variable is defined as the
+join of two other context variables $\gamma'$ and $\gamma''$, which are in turn defined as extensions
+of the same context variable $\gamma_r$, and which contain the same variables,
+the join can be solved by generating a new set of type variables, one for
+each $x_i$, defined as the join of the types known for $x_i$ in $\gamma'$
+and $\gamma''$.
+\end{itemize}
+
+The rules are repeatedly applied to any matching sets of constraints
+until no further rewriting is possible, producing either an error
+or a set of context constraints which are fully normalised, and
+can be converted into contexts for type checking easily.
+
+Given an arbitrary term $t$, the context constraints generated by $t$
+are normalisable (see \appref{app:ctx_norm}).
+This is less general than stating that the context constraint solving rules
+can solve any conjunction of context constraints, but is sufficient for
+the purposes of this language.
+
 \subsection{Additional type variable constraint solving}
 
-In the second phase, additional type variable 
-constraints are generated by following the rule in \figref{fig:tvargen}, which
-compares the observed type from the usage of a variable to the type
-known in the context. Repeated application of this rule until no more matches
-are found produces a constraint set composed exclusively of normalised
-context variable constraints, and type variable constraints. At this point, 
-we can focus on the type variable constraints exclusively as the context
-variable constraints have no further contribution to the overall solution.
-
-\begin{figure}
-\[
-	\infer[\textsc{tvar}]
-	{\tinf = \tinf' \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
-	{x : \tinf \in \gamma \cand \gamma = \gamma', \overline{x_i : \tinf_i}, x : \tinf'}
-\]
-\caption{\label{fig:tvargen} Refining type variable constraints}
-\end{figure}
+During context constraint solving, a number of additional type variable
+constraints are generated by rule \textsc{t\_var} and \textsc{c\_join}.
+These, in addition to the original type variable constraints generated
+directly from the term under consideration, form the constraint set that
+is to be solved by the type variable constraint solving phase.
 
@@
 
-\subsection{Object variable constraint solving}
-
-Normalised object variable constraints are of the forms listed in
-\figref{fig:objconstraints}. Given a set of such normalised
+Solving a set of type variable constraints consists of demonstrating that
+there exists at least one substitution $\sigma$ for all type variables that 
+turns all constraints into tautologies.
+
+For instance, the constraint
+$\alpha = \boolt \cand \alpha' <: \alpha$ has exactly one type variable 
+substitution that provides a valid solution, which is 
+$\sigma = \{ (\alpha, \boolt), (\alpha', \boolt) \}$.
+
+A constraint may have no solution, such as the constraint 
+$\unitt <: \alpha \cand \alpha <: \boolt$, where no type can satisfy the upper
+and lower bounds specified on $\alpha$.
+
+A constraint may also have an infinite number of solutions, such as 
+$\alpha <: \{ S \{ m : \alpha' \Rightarrow S' \} S' \{\} \}$, where
+$\alpha'$ can be any valid type and $\alpha$ can be any object type which 
+allows at least one call to the method $m$, which must return the type 
+$\alpha'$.
+
+The goal of the type constraint solver is to find the most general substitution
+such that all other possible substitutions are {\it refinements} of this base
+substitution, written $\sigma \sqsubset \sigma'$. This means that 
+$dom(\sigma) \subseteq dom(\sigma')$ and that 
+$\forall x \in dom(\sigma)$ . $\sigma'(x) <: \sigma(x)$.
+
+The type system of TS1 does not support type variables, therefore any
+set of constraints for which a concrete lower type bound cannot be found 
+for every type variable has no solution.
+
+The basic types $\boolt$, $\unitt$ and $\funt{T_i}{U_i}{V}$ exhibit 
+{\it structural subtyping}. However object types do not. If we treat object 
+type expressions as atoms,
+then a standard constraint solver which supports structural subtyping such as 
+\ref{Simonet2003} can be used. This solver will find solutions for all of the 
+variables that concern structural types, and leave a set of constraints between 
+object type variables which we must solve through some other mechanism. 
+This is the approach taken for the type constraints for TS1, which separates 
+the task of {\it object constraint solving} from the main type constraint 
+solving phase.
+
+\subsection{Object constraint solving}
+
+This phase of the algorithm is only concerned with constraints of the form
+listed in \figref{fig:objconstraints}. The two classes of variable, object
+and state variables, have no particular significance other than clarifying
+their roles during solving.
+
+The output from the structural subtyping constraint solver should contain
+a set of equality and subtyping constraints that concern object types.
+TODO: describe normalisation of object constraints.
+
+Given a conjunction of such normalised
 constraints $C_\omega$, solving proceeds by application of the rules in
-\figref{fig:objsolving}. Where there is no
-explicit partial or fixed solution for a given object variable $\omega$ in
-a constraint set, the solution for the object is equivalent to an empty partial solution, $\mathbf{partial}\:\emptyset$.
+\figref{fig:objsolving}.
+
+\begin{itemize}
+
+\item $\textsc{=-solve}$ ``connects'' two existing solutions in order to
+produce a single solution, which is as minimal as possible. $connect()$ is
+a variant of state machine union where the initial and final states are
+unknown. Some examples are shown in \figref{fig:connect_ex}
+
+\end{itemize}
 
@@
 \begin{figure*}
+\includegraphics[width=\textwidth]{connect_ex1.pdf}
+\caption{\label{fig:connect_ex} Examples of connecting solutions}
+\end{figure*}
+
+% \begin{figure}
+% \begin{center}
+% \begin{tabular}{c}
+% \infer[\textsc{eq\_norm}]
+% 	{
+% 	\substo{\omega@\sigma}{\alpha}{C}
+% 	}
+% 	{
+% 	C \cand \alpha = \omega@\sigma
+% 	}
+% \\\\
+% \infer[\textsc{sub\_norm\_1}]
+% 	{
+% 	\substo{\omega_2@\sigma_2}{\alpha}{C}
+% 	}
+% 	{
+% 	C \cand \alpha <: \omega@\sigma
+% 	&
+% 	\fresh{\omega_2}
+% 	&
+% 	\fresh{\sigma_2}
+% 	}
+% \\\\
+% \infer[\textsc{sub\_norm\_2}]
+% 	{
+% 	\substo{\omega_2@\sigma_2}{\alpha}{C}
+% 	}
+% 	{
+% 	C \cand \omega@\sigma <: \alpha
+% 	&
+% 	\fresh{\omega_2}
+% 	&
+% 	\fresh{\sigma_2}
+% 	}
+% \\\\
+% \infer[\textsc{join\_}]
+% \end{tabular}
+% \end{center}
+% \caption{\label{fig:objnorm} Object constraint normalisation rules}
+% \end{figure}
+
+\begin{figure*}
 
@@
 	\substo{\sigma}{\sigma_1}{
-	\substo{\sigma}{\sigma_2}{C}}}}, \,
-	\omega = \Omega, \,
+	\substo{\sigma}{\sigma_2}{C}}}} \cand
+	\omega = \Omega \cand
 	\sigma = \Sigma
@@
 	\begin{array}{c}
-	C, \, 
-	\os{1}{1} = \os{2}{2}, \,
-	\omega_1 = \Omega_1, \,
-	\omega_2 = \Omega_2, \,
-	\sigma_1 = \Sigma_1, \,
+	C \cand 
+	\os{1}{1} = \os{2}{2} \cand
+	\omega_1 = \Omega_1 \cand
+	\omega_2 = \Omega_2 \cand
+	\sigma_1 = \Sigma_1 \cand
 	\sigma_2 = \Sigma_2
 	\\
-	\sigma = min(\sigma_1, \sigma_2) 
-	\\
+	\sigma = min(\sigma_1, \sigma_2) \:\:\:
 	\omega = min(\omega_1, \omega_2) 
@@
 {
-	C, \,
-	\omega_1 = \partialo{}{'}, \,
-	\sigma_1 = \overline{S}_{out}, \,
+	C \cand
+	\omega_1 = \partialo{}{'} \cand
+	\sigma_1 = \overline{S}_{out} \cand
 	\sigma_2 = \overline{S}_{in}
@@
 	\begin{array}{c}
-	C, \,
-	\os{1}{1} = \remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}}, \,
-	\omega_1 = \partialo{1}{}, \,
-	\sigma_1 = \Sigma_1, \,
+	C \cand
+	\os{1}{1} = \remapapp{\os{1}{2}}{\os{2}{3}}{\os{2}{4}} \cand
+	\omega_1 = \partialo{1}{} \cand
+	\sigma_1 = \Sigma_1 \cand
 	\sigma_2 = \Sigma_2
 	\\
+		\begin{array}{ccc}
 		\omega_2 = \fixedo{2}{} \in C
-	\\
+		&
 		\sigma_3 = \overline{S}_3 \in C
-	\\
+		&
 		\sigma_4 = \overline{S}_4 \in C
+		\end{array}
 	\\
@@
 {
-	C, \,
-	\omega_1 = \mathbb{S}', \,
-	\sigma_1 = \overline{S}' \,
+	C \cand
+	\omega_1 = \mathbb{S}' \cand
+	\sigma_1 = \overline{S}' \cand
 }
@@
 	\begin{array}{c}
-	C, \,
-	\os{1}{1} <: \os{2}{2}, \,
-	\omega_1 = \partialo{1}{}, \,
+	C \cand
+	\os{1}{1} <: \os{2}{2} \cand
+	\omega_1 = \partialo{1}{} \cand
 	\sigma_1 = \Sigma_1
 	\\
+	\begin{array}{cc}
 	\omega_2 = \fixedo{2}{} \in C
-	\\
+	&
 	\sigma_2 = \overline{S}_2 \in C
+	\end{array}
 	\\
@@
 	\begin{array}{c}
-	C, \,
-	\os{1}{1} = \os{2}{2} \sqcup \os{3}{3} \,
-	\omega_1 = \Omega_1,
+	C \cand
+	\os{1}{1} = \os{2}{2} \sqcup \os{3}{3} \cand
+	\omega_1 = \Omega_1 \cand
 	\sigma_1 = \Sigma_1
 	\\
+	\begin{array}{cccc}
 	\omega_2 = \fixedo{2} \in C
-	\\
+	&
 	\omega_3 = \fixedo{3} \in C
-	\\
+	&
 	\sigma_2 = \overline{S}_2 \in C
-	\\
+	&
 	\sigma_3 = \overline{S}_3 \in C
+	\end{array}
 	\\
@@
 	\begin{array}{c}
-	C, \,
-	\os{1}{1} = \os{1}{2} \veebar \os{1}{3}, \,
-	\omega_1 = \Omega_1,
+	C \cand
+	\os{1}{1} = \os{1}{2} \veebar \os{1}{3} \cand
+	\omega_1 = \Omega_1 \cand
 	\sigma_1 = \Sigma_1
 	\\
+	\begin{array}{cc}
 	\sigma_2 = \overline{S}_2 \in C
-	\\
+	&
 	\sigma_3 = \overline{S}_3 \in C
+	\end{array}
 	\\
@@
 \section{Context constraint normalisation}
+\label{app:ctx_norm}
 
@@
 \item The domains of related contexts change consistently:
-$\forall \gamma_a, \gamma_b \in C, root(\gamma_a, C) = root(\gamma_b, C) \cand dom(\gamma_a, C) = dom(\gamma_b, C) \implies dom(\gamma_a, C') = dom(\gamma_b, C')$.
+$\forall \gamma_a, \gamma_b \in C, root(\gamma_a, C) = root(\gamma_b, C) \cand dom(\gamma_a, C) = dom(\gamma_b, C) + \overline{x_i} \implies dom(\gamma_a, C') = dom(\gamma_b, C') + \overline{x_i}$.
 \end{itemize}
@@
 Given a normalised constraint $C$ where $\gamma \in C$ and $\gamma' \notin C$,
-$C \cand C_t$ is normalisable. Where an error does not occur, the normalised
+$C \cand C_t$ is normalisable. Where an error is not generated, the normalised
 output $C'$ will have the following properties:
@@
 \item $t = unit$. It follows that 
-$C_t \triangleq \gamma' = \gamma$. 
-By \lemref{lem:norm_eq}, $C \cand \gamma' = \gamma$ is normalisable, with
-normalised output $C'$ which has all the required properties.
+$C_t \coloneqq \gamma' = \gamma$. 
+By \lemref{lem:norm_eq}, $C \cand \gamma' = \gamma$ is normalisable.
+If no error is generated, then the normalised output $C'$ has all the required 
+properties.
 
@@
 
-\item $t = \left[ \overline{ S_i \{ \overline{m_ij = (v_ij, S_ij)} \} }  \right]$.
-It follows that 
-$C_t \triangleq 
+\item $t = \left[ \overline{ S_i \{ \overline{m_{ij} = (v_{ij}, S_{ij})} \} }  \right]@S$.
+It follows that \\
+$C_t \coloneqq 
 \gamma' = \gamma \cand
 \overline{\gamma_{ij} = \emptyset} \cand
-\overline{\inferrule{\gamma_{ij}}{v_ij}{\alpha_ij}{\gamma_{ij}'}}.
+\overline{\inferrule{\gamma_{ij}}{v_{ij}}{\alpha_{ij}}{\gamma_{ij}'}}.
 $.
@@
 By \lemref{lem:norm_eq} the constraint $C \cand \gamma' = \gamma$ is
-normalisable, with normalised output $C_2$ such that $C \leq C_2$ with
+normalisable. If no error is generated, the normalised output $C_2$ will be such that $C \leq C_2$ with
 $dom(\gamma, C_2) = dom(\gamma', C_2)$ and $root(\gamma, C_2) = root(\gamma', C_2)$.
@@
 By repeated application of the induction hypothesis, the constraint
-$C_3 \cand \overline{\inferrule{\gamma_{ij}}{v_ij}{\alpha_ij}{\gamma_{ij}'}}$
+$C_3 \cand \overline{\inferrule{\gamma_{ij}}{v_{ij}}{\alpha_{ij}}{\gamma_{ij}'}}$
 is normalisable. If no error is generated, the normalised output $C_4$ will be such that $C_3 \leq C_4$. Transitively, $C \leq C_4$, and the domains and roots
@@
 
-\item $t = \\(\overline{x_i : T_i \gg T_i'}).t'$. 
+\item $t = \lambda(\overline{x_i : T_i \gg T_i'}).t'$. 
 It follows that 
-$C_t \triangleq 
-\gamma' = \gamma \cand \gamma_r = \emptyset \cand 
-\gamma_in = \gamma_r, \overline{x_i : \alpha_i} \cand
-\overline{x_i : \alpha_i' \in \gamma_out} \cand ctx(\inferrule{\gamma_{in}}{t'}{\alpha'}{\gamma_{out}})$.
+$C_t \coloneqq 
+\gamma' = \gamma \cand \\
+\gamma_r = \emptyset \cand \gamma_{in} = \gamma_r, \overline{x_i : \alpha_i} \cand
+\overline{x_i : \alpha_i' \in \gamma_{out}} \cand \\
+ctx(\inferrule{\gamma_{in}}{t'}{\alpha'}{\gamma_{out}})$.
 
@@
 
-The constraint $C_3 \cand \inferrule{\gamma_{in}}{t'}{\alpha}{\gamma_{out}}$
+The constraint $C_3 \cand ctx(\inferrule{\gamma_{in}}{t'}{\alpha}{\gamma_{out}})$
 is normalisable by induction. Let $C_4$ be the normalised output, where
@@
 Let $C_5$ be the constraint 
-$C_4 \cand \overline{x_i : \alpha_i' \in \gamma_out}$.
+$C_4 \cand \overline{x_i : \alpha_i' \in \gamma_{out}}$.
 This constraint is normalisable by repeated application of 
@@
 \item $t = \lett{x}{t'}{t''}$.
-It follows that 
-$C_t \triangleq \gamma_{in} = \gamma_v, x : \alpha' \cand 
-\gamma' = \gamma_{out} - \{x\} \cand
+It follows that \\
+$C_t \coloneqq \gamma_{in} = \gamma_v, x : \alpha' \cand 
+\gamma' = \gamma_{out} - \{x\} \cand \\
 ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma_v}) \cand
@@
 normalised output $C_3$ will be such that $C_2 \leq C_3$ with 
-$dom(\gamma_{in}, C_3) = dom(\gamma_v, C_2) + \{ x \}$.
+$dom(\gamma_{in}, C_3) = dom(\gamma_v, C_3) + \{ x \}$.
 
@@
 
-Transitively, $C \leq C_5$.
-
-% relate dom(\gamma) and dom(\gamma')
-% we currently preserve equality relationships between domains,
-% can we preserve equality plus variable set?
+As $C_3 \leq C_5$, it can be observed that $dom(\gamma_{in}, C_5) = dom(\gamma_v, C_5) + \{x\}$ (as this relationship holds for $C_3$) and that 
+$dom(\gamma_{in}, C_5) = dom(\gamma_{out}, C_5)$.
+Therefore, $dom(\gamma') = dom(\gamma_{out}, C_5) - \{x\} = dom(\gamma)$.
+Additionally, $root(\gamma, C_5) = root(\gamma', C_5)$ and transitively, $C \leq C_5$.
 
@@
 It follows that
-$C_t \triangleq 
+$C_t \coloneqq 
 ctx(\inferrule{\gamma}{t'}{\alpha'}{\gamma''}) \cand
@@
 It follows that
-$C_t \triangleq \overline{x_i : \alpha_i \in \gamma} \cand 
+$C_t \coloneqq \overline{x_i : \alpha_i \in \gamma} \cand 
 x : \tinf_x \cand
@@
 It follows that
-$C_t \triangleq x : \alpha_o@\alpha_s \in \gamma \cand
+$C_t \coloneqq x : \alpha_o@\alpha_s \in \gamma \cand
 \gamma' = \gamma[x \mapsto \alpha_x]
@@
 It follows that
-$C_t \triangleq
+$C_t \coloneqq
 ctx(\inferrule{\gamma}{t_c}{\alpha_c}{\gamma_c}) \cand
@@
 $C_4 \cand \gamma' = \gamma_t \sqcup \gamma_f$ can be transformed to
-$C_5 = C_4 \cand \gamma' = \gamma_r, \overline{x_i : T_i} \cand \overline{T_i = T_t \veeq T_f}$ 
+$C_5 = C_4 \cand \gamma' = \gamma_r, \overline{x_i : T_i} \cand \overline{T_i = T_t \veebar T_f}$ 
 where $\overline{x_i} = dom(\gamma, C_4)$. Therefore $dom(\gamma, C_4) = dom(\gamma', C_4)$ and $root(\gamma, C_4) = root(\gamma', C_4)$. Trivially, $C_4 \leq C_5$ and transitively $C \leq C_5$.
@@
 If $\overline{x_i} \cap \overline{x_j} \neq \emptyset$, then by rule
-$\textsc{c\_ext\_2}$ a $\mathbf{bound}\:x$ error will be generated.
+$\textsc{c\_ext\_2}$ a $\mathbf{bound}\:x$ error will be generated, for
+some $x$.
 \end{enumerate}
@@
 error is generated, the normalised constraint $C_3$ will be produced such
-that $C \leq C_3$ and $x \in dom(\gamma, C_3$. Therefore, by the 
+that $C \leq C_3$ and $x \in dom(\gamma, C_3)$. Therefore, by the 
 reasoning of case 3 below, $C_3 \cand \gamma' = \gamma[x \mapsto \tinf]$ is 
@@
 $C \cand x : \tinf \in \gamma$ is normalisable. When an error is not generated, 
-the constraint $C_2$ will be generated such that $C \leq C_2$ and
-$x \in dom(\gamma, C_2)$.
+the constraint $C'$ will be generated such that $C \leq C'$ and
+$x \in dom(\gamma, C')$.
 \end{lem}

### 2012-11-30 formalism/paper.tex
replaced \sim with \precsim to make it clearer that we mean simulation rather than bisimulation (that there is a direction).

@@
 $O_2 = \langle M_2, \Sigma_2, \Delta_2 \rangle$, the simulation preorder
-$\sim$
-is defined to be the largest relation such that if $O@S \sim O_2@S_2$ then for all
+$\precsim$
+is defined to be the largest relation such that if $O@S \precsim O_2@S_2$ then for all
 $m$ and $S_4$ such that $(S_2, m, S_4) \in \Delta_2$ there exists an $S_3$ such that
-$(S, m, S_3) \in \Delta$ and that $O@S_3 \sim O_2@S_4$.
-Subtyping is defined such that $O@S <: O_2@S_2 \iff O@S \sim O_2@S_2$.
+$(S, m, S_3) \in \Delta$ and that $O@S_3 \precsim O_2@S_4$.
+Subtyping is defined such that $O@S <: O_2@S_2 \iff O@S \precsim O_2@S_2$.
 
@@
 It is easy to
-show that $O@S \sim O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$, so these
+show that $O@S \precsim O_2@S_2 \iff \Tr(O_2@S_2) \subseteq \Tr(O@S)$, so these
 interpretations of subtyping can be used interchangeably wherever necessary.
@@
 In general, the set of
-types in TS is a disjoint union of join semi-lattices.
+types in TS is a join semi-lattice, with $\top$ as the greatest element.
 
@@
 and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
-\item $O@\overline{S} \sim O_2@\overline{S_2}$ is the extension of the
+\item $O@\overline{S} \precsim O_2@\overline{S_2}$ is the extension of the
 simulation relation to support state sets, and can be interpreted as
-$\bigsqcup_{S \in \overline{S}}O@S \sim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
+$\bigsqcup_{S \in \overline{S}}O@S \precsim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
 \end{itemize}
@@
 
-It is clear that as $\ot{O}{S} \sim \ot{O_2}{S_2}$ there must exist some
-$\overline{S_3}$ such that $\ot{O}{S_3} \sim \ot{O_2}{S_4}$.
+It is clear that as $\ot{O}{S} \precsim \ot{O_2}{S_2}$ there must exist some
+$\overline{S_3}$ such that $\ot{O}{S_3} \precsim \ot{O_2}{S_4}$.
 One possible solution would be to choose
-$\overline{S_3} = \{ S | O@S \sim \ot{O_2}{S_4} \}$. This is not specific
+$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. This is not specific
 enough however --- given the object protocols from \figref{fig:principalex},
@@
 {\it simulation predecessor relation}. The definition of this relation
-is given in \figref{fig:simreach}. Essentially, if $O@S \sim O_2@S_2$ and there
+is given in \figref{fig:simreach}. Essentially, if $O@S \precsim O_2@S_2$ and there
 exists a sequence of method calls that would transition an object value of type
@@
 would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
-such that $O@\overline{S_3} \sim O@\overline{S_4}$. This is abbreviated to
+such that $O@\overline{S_3} \precsim O@\overline{S_4}$. This is abbreviated to
 $(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.
@@
     \begin{array}{cc}
-    O@\overline{S} \sim O_2@\overline{S_2}
+    O@\overline{S} \precsim O_2@\overline{S_2}
     \\
@@
   {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S}, O_2@\overline{S_2})}
-  {O@\overline{S} \sim O_2@\overline{S_2}}
+  {O@\overline{S} \precsim O_2@\overline{S_2}}
 \\\\
@@
 then a standard constraint solver which supports structural subtyping such as 
-\ref{Simonet2003} can be used. This solver will find solutions for all of the 
+\cite{Simonet2003} can be used. This solver will find solutions for all of the 
 variables that concern structural types, and leave a set of constraints between 
@@
 \section{Soundness}
+\label{app:soundness}
 
@@
 $T = O'@\overline{S''}$ and $U = O@\overline{S}$ and
-$\overline{S'''} = \{ S | O'@S \sim O@\overline{S'}, S \in O' \}$ 
+$\overline{S'''} = \{ S | O'@S \precsim O@\overline{S'}, S \in O' \}$ 
 for some $O'$, $\overline{S}$ and $\overline{S''}$. By \subobj,
@@
 for some $O'$ and $\overline{S'}$ 
-such that $O'@\overline{S'} \sim O@\overline{S}$ 
+such that $O'@\overline{S'} \precsim O@\overline{S}$ 
 by rule \subobj. 
 By rule \remapobj, it follows that $T = O'@\overline{S''}$ 
-such that $\overline{S''} = \{ S | O'@S \sim O@\overline{S}, S \in O' \}$.
+such that $\overline{S''} = \{ S | O'@S \precsim O@\overline{S}, S \in O' \}$.
 By \lemref{lem:ssetinc} it follows that 
@@
 $T'$ must be of form $O'@\overline{S}''$ such that
-$O'@\overline{S}'' \sim O@\overline{S}$ by rule \subobj.
+$O'@\overline{S}'' \precsim O@\overline{S}$ by rule \subobj.
 
@@
 Let $U' = \bigsqcup{ \{ T_r' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}'' \} }$.
-As $O'@\overline{S}'' \sim O@\overline{S}$, it follows that
-$O'@\overline{S}''' \sim O@\overline{S}'$.
+As $O'@\overline{S}'' \precsim O@\overline{S}$, it follows that
+$O'@\overline{S}''' \precsim O@\overline{S}'$.
 Therefore, $U' <: U$ and $V' <: V$. By the rule \tmethc, and \tsub

### 2013-01-02 formalism/paper.tex
progress and preservation theorem reviewed and fixed

@@
 Current typestate research is primarily focused on studying
-type systems that can support particular use cases for typestate;
+type systems that can support particular use cases;
 the type systems are often complex, and programs in such languages 
@@
 models interactions with typestate-constrained objects and a type inference
-technique which can derive the minimal state machine that governs interactions
-with the observed objects, and therefore eliminate the need for
-explicit type annotations.
+technique which can derive principal typings for terms with some
+restrictions, and therefore eliminate the need for
+explicit type annotations in most cases.
 \end{abstract}
@@
 insufficient to correctly model the object's behaviour in many cases, as
-the set of messages the object can correctly respond to is dependent upon an
-abstraction of its private state. State change is typically synchronous, in 
+the set of messages the object can correctly respond to is dependent upon
+its private state. State change is typically synchronous, in 
 response to message receipt, and safety requires that 
 any message sent to an object is within the set which can be received for
-the currently known state.
+the current state. The private state can often be abstracted
+such that restrictions on sequences of messages can be expressed as a
+deterministic finite state machine.
 
@@
 in which state, and what state transformation will occur (which may not
-be deterministic). This type annotation burden is particularly pronounced
+be deterministic). This type annotation burden is particularly heavy
 in a language with function literals, which must be succinct
-in order to be readable.
-
-Additionally, static checking of typestate requires
+in order to be practical.
+
+Static checking of typestate requires
 that the current state of an object always be known at a point of invocation.
-This is only possible when an alias control strategy is employed, such as linear types, 
-but such types prohibit programming in a style that most programmers are 
+This is only possible when an alias control strategy is employed, such as linear
+types, but such types prohibit programming in a style that most programmers are 
 familiar with. Attempting to relax linear typing through the use of fractional 
 permissions as in Plural and Plaid \cite{Aldrich2009,Bierhoff2009}
-adds a significant amount of complexity to the type system. This results in 
+adds further complexity to the type system. This results in 
 large type annotations within programs to express both typestate and aliasing 
 constraints when modular static type checking is desired.
-If it were possible to eliminate the need for such type annotations, code
-could be more concise and type annotations would only be necessary for
-documentation purposes, and could be checked that they are simply more specific
-versions of the principal typing produced by the type inference algorithm.
+If it were possible to eliminate the need for such type annotations in
+most situations, code could be more concise and readable, with annotations
+reserved for situations where inference is not possible or are required
+for formal documentation purposes.
 
@@
 
-A perhaps more intuitive basis for subtyping
-uses the notion of {\it trace inclusion}. The set of interaction traces 
-$\Tr(O@S)$ of a type is the smallest set such that:
+A perhaps more intuitive basis for subtyping is {\it trace inclusion}. The set 
+of interaction traces $\Tr(O@S)$ of a type is the smallest set such that:
 
@@
 
-Either view of subtyping is also easy to relate to the Liskov substitution 
-principle \cite{Liskov1987} --- a type will accept every sequence of
+Either view of subtyping is comparable to the Liskov substitution 
+principle \cite{Liskov1987} --- a type must accept every sequence of
 method calls that a supertype would, therefore it is a {\it behavioural}
@@
 to study typestate inference in imperative code, rather than to 
-be practical, as it uses syntactic control of interference \cite{Reynolds1978} 
-to avoid the issue of alias control entirely - there is no means within the
-language to produce an alias to an object.
+be practical, general purpose programming language. The language uses syntactic 
+control of interference \cite{Reynolds1978} to avoid the issue of alias control 
+entirely - there is no means within the language to produce an alias to an 
+object. This focuses attention on the fundamentals of typestate inference.
 
@@
 the result type for terms which have no meaningful result.
-\item Function literals, which do not allow for implicit capture
+
+\item {\tt true} and {\tt false}, of type {\tt Bool}, which are used for
+decisions in conditional evaluation. The language contains no operations
+for manipulating booleans, to keep the calculus small, however
+such operations would be straightforward to add.
+
+\item Function literals of form $\funv{x_i}{T_i}{T_i'}{t}$, where
+$\overline{x_i}$ is the set of parameters to the function while
+$t$ is the body term. Parameters are passed by reference.
+Function literals do not allow for implicit capture
 of variables into their scopes --- everything must be passed as a parameter. 
-Tracking effects on implicitly captured references is not very well
-studied in the current literature outside of strict linear types where
+Tracking effects on implicitly captured references is not well understood
+in the current literature outside of strict linear types where
 function literals that capture linear references must themselves be linear.
-Parameters are passed by reference. 
 Requiring that all references be explicitly passed to the function allows for
@@
 on the right. Not all effect types are legal, as will be discussed later.
-\item Object literals. These are composed of a state machine description
-and a current state. The state machine description contains a set of
-distinct state labels $\overline{S_i}$, within which is a set of methods $\overline{m_{ij}}$ 
+
+\item Object literals of form 
+$O@S$, where $O = \left[ \overline{ S_i \{ m_{ij} = (v_{ij}, S_{ij}) \} } \right]$. 
+$O$ is the object protocol, and $S$ is the current state
+(where $S \in \overline{S_i}$). The object protocol contains a set of
+distinct state labels $\overline{S_i}$, within which is a set of methods 
+$\overline{m_{ij}}$ 
 which can be called in that state. The language does
@@
 in $x$. After a valid invocation, the state of the object will have changed
-as specified by the object's state machine.
+as specified by the object's protocol.
 \item A let-bind construct $\lett{x}{t}{t'}$, which creates a new variable
@@
 The example in \figref{fig:exampleprog} shows a short TS0 program
-that TODO.
+that emulates the usage of an iterator. The object {\tt it}
+provides an iterator-like contract, starting in the state {\tt CHECK} where
+only the method {\tt hasNext} can be called, which returns a boolean value.
+This method changes the state of the object to {\tt READ}, where both
+{\tt hasNext} and {\tt next} can be called. Calling {\tt next} changes the
+state of the object back to {\tt CHECK}. While the language does not
+support {\it conditional transitions} that would be necessary to express
+the restriction that {\tt next} should only be callable after {\tt hasNext}
+returns {\tt true}, it does still provide more protection than a language
+without typestate would.
 
@@
 \begin{lstlisting}
-TODO
+let it = [ CHECK { hasNext = (true,READ) } 
+          READ  { hasNext = (true,READ);
+                  next = (unit,CHECK) }
+        ]@CHECK in
+let console = [S { print = (unit,S) }]@S in
+let f = \(x).(if x.hasNext 
+              then (x.next ; true) 
+              else false) in
+if f(it) then console.print else unit
 \end{lstlisting}
@@
 \end{figure}
+
+The object {\tt console} is effectively stateless, allowing the method
+{\tt print} to be called without restriction.
+
+The function {\tt f} will call {\tt hasNext}, and if it returns true then
+will call {\tt next}. It returns a boolean value indicating whether
+{\tt next} was called.
+
+Finally, the object {\tt it} is passed to {\tt f} and {\tt print} will
+be called on {\tt console} if {\tt f} returns true. Given the fixed return
+values for the methods on {\tt it}, {\tt print} will always be called.
+
+The most interesting part of the program is the function literal assigned
+to {\tt f}. What is the effect type for the parameter {\tt x}? 
+This is the only piece of type information that cannot be determined
+easily, and is key to determining whether {\tt f} would use {\tt it} in
+a safe manner.
 
@@
 may contain mappings for variables which are not in the context as shown in
-rule \textsc{st\_extra}.
+rule \textsc{st\_extra}. Such mappings are candidates for garbage collection,
+or an explicit deallocation instruction could be injected after the
+evaluation of the body of a let-binding.
 
@@
 
-\subsection{Type rules and soundness}
+\subsection{Type system}
 
@@
 relation.
+
+The type system of TS is sound (it has both progress and type preservation
+properties), which is proven in \thmref{thm:soundness} in 
+\appref{app:soundness}.
 
@@
 
-Corresponding to the values of the language, there are four main kinds
-in the type system: Unit, Bool, function types and object types. 
+There are four kinds in the type system: Unit, Bool, function types and 
+object types.
 Two modifications are made to the definition of object protocols described
@@
 exactly it is in --- conditional evaluation readily produces this situation.
-With an if-then-else statement where the branches
+With an if-then-else expression where the branches
 interact with an object differently, it is common for the state of the object
 at the end of each branch to be different. As
-knowing precisely which branch will be taken is rarely possible in a
-decidable type system, we must either insist that the state of the object
+knowing precisely which branch will be taken statically is rarely possible, 
+we must either insist that the state of the object
 be the same upon exit from both branches, or devise a means to cope with
@@
 the two state machines in the same fashion. Join is always defined for
-object types, while meet is not --- where the return types of methods are
-incompatible (i.e. no join/meet exists respectively), a join operation will 
-discard the method (it is unsafe to
-call it) while a meet operation would require that the method be callable,
-however no valid return type can be assigned to the method. Therefore
-the kind of object types is a join semi-lattice, with the object type 
-$\{ S_1 \{\} \}@S_1$ serving as the least element.
-
-The intuition here is that the only sequences of method calls which can be 
-safely invoked on an object of type $O@\overline{S}$ are those which could be 
-safely invoked on each individual $O@S$, where $S \in \overline{S}$, and that
-a least upper bound of the return types exists.
+object types, while meet is not. Consider the two object types
+$\{ S \{ m : \unitt \Rightarrow S \} \}@S$ and
+$\{ S \{ m : \boolt \Rightarrow S \} \}@S$. The meet of these two types
+requires that method $m$ is callable, but $\unitt \sqcap \boolt$ is undefined
+and therefore no return type can be assigned to the method $m$.
+
+The intuition for state sets is that the only sequences of method calls which 
+can be safely invoked on an object of type $O@\overline{S}$ are those which 
+could be safely invoked on each individual $O@S$, where $S \in \overline{S}$.
 
@@
 types in TS is a join semi-lattice, with $\top$ as the greatest element.
+The greatest element within object types exclusively is $\{ S \{\} \}@S$,
+that is, an object which allows no method calls.
 
@@
 
-Given the formal definition of an object type $O@\overline{S}$, to save
-referring to the $\Sigma$, $M$ and $\Delta$ that $O$ is composed of we shall
-use the following notational conventions:
+Given an object type $O@\overline{S}$, to save
+referring to the $\Sigma$, $M$ and $\Delta$ that formally define $O$
+the following notational conventions shall be used:
 
@@
 asserting the existence
-of method $m$ with return type $T$ in state $S$ which triggers a 
+of method $m$ with return type $T$ in state $S$, which triggers a 
 transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
@@
 of the method existence predicate to state sets, meaning that for all 
-$S \in \overline{S}$ there exists a state and type 
-$T'$ where $(S,m,T',S') \in \Delta$ and that
-$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
-and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
+$S \in \overline{S}$ there exists a state $S'$ and type 
+$T'$ where $(S,m,T',S') \in \Delta$.
+$T$ is defined as 
+$\bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$
+and $\overline{S'}$ is defined as $\{ S' | (S,m,T',S') \in \Delta \}$.
 \item $O@\overline{S} \precsim O_2@\overline{S_2}$ is the extension of the
@@
 
+\subsection{Object type isomorphism}
+
+As object types in TS are based upon deterministic finite state machines,
+there are often multiple ways to represent the same type. Object types
+$O_1@S$ and $O_2@S$ are considered {\it isomorphic} 
+(up to $\alpha$-conversion of state labels)
+if they are both simulations of each other:
+$O_1@\overline{S_1} \precsim O_2@S_2 \wedge O_2@\overline{S_2} \precsim O_1@\overline{S_1}$. 
+The
+{\it canonical} form amongst a set of isomorphic object types has a
+{\it minimal} finite state machine.
+
 \subsection{Effect types}
@@
 An ``effect type'' $T \gg T'$ describes how the type of a variable changes
-in response to function application. In the TS language, the only type
-changes that are those which change the state of an object instance; the
-values of the other types are immutable. If a boolean is passed as a parameter,
-the effect type will always be $\boolt \gg \boolt$, similarly for
-$\unitt$ and $\funt{T}{U}{V}$. An effect type of $\boolt \gg \unitt$ is not
-possible. In general, effect types are kind-preserving.
+in response to function application. In the TS language, as the values of
+all types with the exception of objects are immutable, effects are
+either identities ($T \gg T$) or object state changes 
+($O@\overline{S_1} \gg O@\overline{S_2}$). As references cannot
+be overwritten, we are guaranteed that the object instance cannot be replaced.
+Therefore an effect type $O_1@\overline{S_1} \gg O_2@\overline{S_2}$ can
+only be valid if $O_1 = O_2$.
 
@@
 \includegraphics[width=0.49\textwidth]{sm1.pdf}
-\caption{\label{fig:sm1} Operational semantics}
+\caption{\label{fig:sm1} A simple object protocol}
 \end{figure}
 
-In the case of an object parameter, not all effect types are possible
-either. Consider
+State set transformations are restricted for semantic reasons also. Consider
 the object protocol in \figref{fig:sm1}, which we will refer to as $O_1$.
@@
 state $B$ to both state $B$ and $D$.
-An effect type $O_1@{B,D} \gg O_1@{B,C}$ which has a state set on
-the left hand side is possible as there exists a sequence of method calls 
+An effect type $O_1@{B,D} \gg O_1@{B,C}$ which has more than one input state
+is possible as there exists a sequence of method calls 
 that would take an object in state $B$ to $C$ that would also
 take an object in state $D$ to $B$ (specifically, a single method
-call to $b$).
+call to $b$, which we can treat as having return type $\top = \boolt \sqcup \unitt$).
 
@@
 of form $O@\overline{S} \gg O@\overline{S_2}$ is
-legal if:
+valid if:
 
@@
 
-Checking the legality of an effect type is straightforward: by computing the
-intersection of all the states, we can associate with each state in the 
-intersection a set of states in the original machine
-it was derived from (as shown in \figref{fig:sm1}. 
-Should an associated state set contain an end state,
-if all other states in the set are also end states then this represents a
-``candidate'' end state within the intersection machine. The union of the state 
-sets associated with all candidate end states must be equal to the
-desired end state set for the effect type to be legal. This would correctly rule 
-out the effect type $O@\{B,D\} \gg O@\{B\}$ (there is no sequence of method
-calls that would take an instance in either state $B$ or $D$ and result in
-the object exclusively being in state $B$) while accepting effect types
-$O@\{B,D\} \gg O@\{B,D\}$ (corresponding to no operations occurring) and
-$O@\{B,D\} \gg O@\{B,C,D\}$ (corresponding to either no operations occuring
-or a single call to $b$).
-
-\subsection{Principal types}
-
-A principal type is one which makes the smallest commitment about 
-the structure of a value based on its usage; due to the presence of subtyping 
-in TS, a principal type is an upper bound on the type of the value.
-
-\begin{figure}
-\includegraphics[width=0.49\textwidth]{principal.pdf}
-\caption{\label{fig:principalex}}
-\end{figure}
-
-Following from this general definition, a principal object type in TS 
-is an object protocol and initial state set which permits all sequences of 
-method calls that could be generated by an associated term $t$, and no more. 
-In other words,
-the language accepted by the state machine is the smallest possible, while
-the return types of methods are as general as possible. Consider the example
-function $f$ shown in \figref{lst:principalex}. It is clear from the usage
-of $y$ that it must be of type $\boolt$ as demanded by the type judgement
-of an if-then-else statement, while $x$ is an object of some sort.
-We could ascribe any of the object types $O_1@A$, $O_2@A$ and $O_3@A$ to the parameter 
-$x$, and after evaluation the object types would be $O_1@A$, $O_2@\{E,H\}$
-and $O_2@F$ respectively.
-$O_1@A$ is not principal as it will admit many more sequences of method calls 
-than strictly required by the body of function $f$, while $O_2@A$ and $O_3@A$ 
-are principal.
-
-$O_2$ contains more states than are strictly necessary, i.e. it
-is not a {\tt minimal} state machine, while $O_3$ is minimal.
-We do not require that a principal type be minimal. If necessary, any
-principal state machine we produce can be minimised, so this can be done
-on demand where there is a specific need for it. One such need may be when
-presenting the type to a user, as a minimal state machine may be easier to 
-understand. Regardless, the state machines $O_2$ and $O_3$ can be shown to be 
-isomorphic in the language that they accept (starting at state $A$), and that 
-language is minimal, so we assert that either are acceptable principal types.
-
-Similarly, it is possible for a principal effect type to have more than
-one state in the state set of the input (left) object type. It is
-always possible to restructure such an object type to have a single initial
-state. It is not, however, possible to have both a single input state and
-single output state for a state machine accepting an arbitrary language --- 
-consider the language accepted by the regular expression $m^* | n^*$. It is 
-necessary for a state machine which accepts this language, and only this 
-language, to
-have at least three exit states. In general, it is more intuitive to have
-a single input state in the effect type but this is not required for a
-type to be considered principal.
-
-A principal type we can therefore ascribe to the function $f$
-is $\forall T_1 \ldots T_6 . (O_3@A \gg O_3@F, \boolt) \rightarrow T_6$, 
-or equivalently 
-$\forall T_1 \ldots T_6 . (O_2@A \gg O_2@\{E,H\}, \boolt) \rightarrow T_6$.
-
-There are occasional advantages to not always minimising principal state
-machines. The additional structure of $O_2$ retains additional information
-in the effect type which can be helpful when interpreting an effect type
-when a subtype is passed to the function, as will be described in the next
-section.
-
-\subsection{Interpreting effect types}
+Checking the validity of an effect type is straightforward: given an
+effect type $O@\overline{S_1} \gg O@\overline{S_2}$, we first compute
+the intersection of all states $S \in \overline{S_1}$. In computing this
+intersection, we can associate with each new state in the object protocol
+representing the intersection to those states in the original protocol it
+was derived from. Wherever one of these associated sets contains a state
+$S \in \overline{S_2}$, all of the states in that set must be in 
+$\overline{S_2}$. Finally, all end states $S \in \overline{S_2}$ must
+exist in one of the associated sets.
+
+An example of this is shown in \figref{fig:sm1} for the protocol $O_1$.
+If the input of the effect type is $O_1@\{B,D\}$ then the only valid
+outputs are either $O_1@\{B,D\}$ (no calls are made),
+$O_1@\{B,C\}$ (exactly one call to $b$ is made), or
+$O_1@\{B,C,D\}$ (the union of those two options). The effect type
+$O_1@\{B,D\} \gg O_1@\{B\}$ would be correctly identified as invalid.
+
+\subsection{Interpreting effect types with subtyping}
 
@@
 be a subtype of the input portion of the associated effect type.
-Determining a precise upper bound for the type of the parameter after the
+Determining the type of the parameter after after the
 function call, in particular where the parameter is an object, 
-requires some thought.
-
-Consider a function call $f(x)$ where
-$f : (O_2@\overline{S_2} \gg O@\overline{S_4}) \rightarrow T$ and
-$x : O_1@\overline{S} <: O_2@\overline{S_2}$, where $O_1$ and $O_2$
-are the state machines from \figref{fig:principalex}, 
-$\overline{S} = \{ A \}$, $S_2 = \{ A \}$ and $S_4 = \{E, H\}$. 
-Based on the effect type,
-we can assert that the function will have interacted with the $x$ using
-one of the possible sequences that would be accepted by $O_2@\overline{S_2}$,
-ending with the object in the state set $\overline{S_4}$.
-
-A straightforward but questionable choice would 
-be to replace the type of the variable
-with that of the output portion of the effect type after the call. However,
+requires some thought to avoid losing type information unnecessarily.
+
+Consider a function call $f(x)$ where \\
+$f : (O_2@A \gg O_2@\{E,H\}) \rightarrow \top$ and
+$x : O_1@A$, where $O_1$ and $O_2$
+are the state machines from \figref{fig:principalex}.
+It is clear that $O_1@A <: O_2@A$, therefore by definition any possible
+sequence of method calls that $f$ could call on $x$ would be valid.
+What should be the type of $x$ after the call, given the information that
+the effect type gives?
+
+A very naive but safe choice would be to choose $O_2@\{E,H\}$. However,
 this will often result in an unacceptable loss of type information,
 reducing the number of valid programs which can be typed within the system.
-
-It is clear that as $\ot{O}{S} \precsim \ot{O_2}{S_2}$ there must exist some
-$\overline{S_3}$ such that $\ot{O}{S_3} \precsim \ot{O_2}{S_4}$.
+Based on the effect type,
+we know that the function will interact with the $x$ using
+one of the possible sequences of methods calls from $O_2@A$ to 
+$O_2@E$ or $O_2@H$. There must exist some
+$\overline{S_3}$ such that $\ot{O_1}{S_3} \precsim O_2@\{E,H\}$.
 One possible solution would be to choose
-$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. This is not specific
-enough however --- given the object protocols from \figref{fig:principalex},
-$\overline{S_3} = \{ A, B, C, D \}$ as there are no possible method calls
-on states $E$ or $H$ of $O_2$.
+$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. 
+This is still liable to lose type information however --- given the object 
+protocols from \figref{fig:principalex}, this would produce
+$\overline{S_3} = \{ A, B, C, D \}$, as there are no possible method calls
+on states $E$ or $H$ of $O_2$ (all object types trivially simulate the
+empty object type $\{ S\{\} \}@S$). This chosen solution also ignores the
+path taken to $O_2@E$ or $O_2@H$, and so may contain states
+that are actually unreachable from $O_1@A$.
 
 It is clear that our choice of $S_3$ must relate the set of possible method
-call sequences between $\ot{O_2}{S_2}$ and $\ot{O_2}{S_4}$. For this we
+call sequences between $O_2@A$ and $O_2@\{E,H\}$. For this we
 define a relation between pairs of object types
-$(O@\overline{S}, O_2@\overline{S_2} \lesssim 
-(O@\overline{S'}, O_2@\overline{S_2'})$, which we call the
-{\it simulation predecessor relation}. The definition of this relation
-is given in \figref{fig:simreach}. Essentially, if $O@S \precsim O_2@S_2$ and there
-exists a sequence of method calls that would transition an object value of type
-$O_2@\overline{S_2}$ to $O_2@\overline{S_4}$, then this sequence of method calls
-would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
-such that $O@\overline{S_3} \precsim O@\overline{S_4}$. This is abbreviated to
-$(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.
-
-Using this relation, we can choose $\overline{S_3} = 
-\bigcup \{ \overline{S'} | 
-  (\ot{O}{S} , \ot{O_2}{S_2}) \lesssim (\ot{O}{S'}, \ot{O_2}{S_4})
-\}$. The solution using the object protocols from \figref{fig:principalex}
-will therefore be $\overline{S_3} = \{ A \}$.
+$(O_1@\overline{S}, O_2@\overline{S_2} \lesssim 
+(O_1@\overline{S_3}, O_2@\overline{S_4})$, which we call the
+{\it path relation}, which defines equivalent paths through the
+respective object protocols. The definition of the path relation
+is given in \figref{fig:simreach}.
 
@@
 \begin{array}{c}
-\\
+
+\infer[\lesssim\textsc{-refl}]
+  {(O_1@\overline{S_1}, O_2@\overline{S_2}) \lesssim (O_1@\overline{S_1}, O_2@\overline{S_2})}
+  {O_1@\overline{S_1} \precsim O_2@\overline{S_2}}
+\\\\
 \infer[\lesssim\textsc{-adj}]
-  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})}
+  {(O_1@\overline{S_1}, O_2@\overline{S_2}) \lesssim (O_1@\overline{S_3}, O_2@\overline{S_4})}
   {
     \begin{array}{cc}
-    O@\overline{S} \precsim O_2@\overline{S_2}
-    \\
-    m : T_i \Rightarrow \overline{S_4} \in O@\overline{S_2}
-    \\
-    m : T_i' \Rightarrow \overline{S_3} \in O@\overline{S}
+    \multicolumn{2}{c}{O_1@\overline{S_1} \precsim O_2@\overline{S_2}}
+    \\
+    m : T_i \Rightarrow \overline{S_4} \in O_2@\overline{S_2}
+    &
+    m : T_i' \Rightarrow \overline{S_3} \in O_1@\overline{S_1}
+    \end{array}
+  }
+
+\\\\
+\infer[\lesssim\textsc{-trans}]
+  {
+  (O_1@\overline{S_1}, O_2@\overline{S_2}) \lesssim (O_1@\overline{S_5}, O_2@\overline{S_6})
+  }
+  {
+  \begin{array}{c}
+  (O_1@\overline{S_1}, O_2@\overline{S_2}) \lesssim (O_1@\overline{S_3}, O_2@\overline{S_4})
+  \\
+  (O_1@\overline{S_3}, O_2@\overline{S_4}) \lesssim (O_1@\overline{S_5}, O_2@\overline{S_6})
   \end{array}
@@
 \\\\
-\infer[\lesssim\textsc{-refl}]
-  {(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S}, O_2@\overline{S_2})}
-  {O@\overline{S} \precsim O_2@\overline{S_2}}
-\\\\
-\infer[\lesssim\textsc{-trans}]
+\infer[\lesssim\textsc{-branch}]
 	{
-  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
+	(\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S_7}, \ot{O_2}{S_8})
 	}
 	{
-  \begin{array}{c}
-  (O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O_2@\overline{S_4})
-  \\
-  (O@\overline{S_3}, O_2@\overline{S_4}) \lesssim (O@\overline{S_5}, O_2@\overline{S_6})
+	\begin{array}{cc}
+	\multicolumn{2}{c}{
+	(\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S_3}, \ot{O_2}{S_4})
+	}
+	\\
+	\multicolumn{2}{c}{
+	(\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S_5}, \ot{O_2}{S_6})
+	}
+	\\
+	\overline{S_7} = \overline{S_3} \cup \overline{S_5}
+	&
+	\overline{S_8} = \overline{S_4} \cup \overline{S_6}
 	\end{array}
 	}
+
 \end{array}
@@
 
-The type system of TS uses a partial function named {\it remap}, defined in
-\figref{fig:remaprules} to formalise
-this selection of the type of a parameter after a function call. The cases
-for Bool, Unit and function types are trivial as these types are immutable,
-while for objects the simulation predecessor relation is used.
+The type system of TS defines a partial function named {\it remap}, 
+formalised in \figref{fig:remaprules}, to choose the type of a passed parameter
+after a function call. For object types, the path relation is used to choose
+the correct state set. For our example, the type of $x$ after the call
+$f(x)$ will therefore be $O_1@A$, as
+$(O_1@A, O_2@A) \lesssim (O_1@A, O_2@{E,H})$.
 
@@
 
-\subsection{Soundness}
-
-The soundness of the type system is stated formally in the following
-theorem:
-
-\begin{thm}[Progress and Preservation]
-Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
-with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
-there exist $t'$ and $\mu'$ such that
-$\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
-$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
-\end{thm}
-
-\begin{proof}
-in appendix \appref{app:soundness}.
-\end{proof}
+\subsection{Principal typings \& types}
+
+A typing of the term $t$ is the triple $\tau = (\Gamma,T,\Gamma')$ such that
+$\typerule{\Gamma}{t}{T}{\Gamma'}$. Typings can be ordered using the subtyping
+relation. Given two typings $\tau_1 = (\Gamma_1,T_1,\Gamma_1')$ and
+$\tau_2 = (\Gamma_2,T_2,\Gamma_2')$ of $t$ we say that $\tau_2$ is 
+{\it stronger} than $\tau_1$, written $\tau_1 \leq \tau_2$, if:
+
+\begin{enumerate}
+\item $\Gamma_2$ contains fewer variables: 
+      $dom(\Gamma_1) \supset dom(\Gamma_2)$
+\item If $dom(\Gamma_1) = dom(\Gamma_2)$, 
+	  then the variables of $\Gamma_2$ have less specific types: \\
+      $\forall x \in dom(\Gamma_1), \Gamma_1(x) <: \Gamma_2(x)$
+\item If $\Gamma_1 = \Gamma_2$, then $\Gamma_2'$ contains more variables:
+      $dom(\Gamma_1') \subset dom(\Gamma_2)$
+\item If $dom(\Gamma_1') = dom(\Gamma_2')$, then $\Gamma_2'$ contains
+      more specific types: \\
+      $\forall x \in dom(\Gamma_1), \Gamma_2'(x) <: \Gamma_1'(x)$
+\item If $\Gamma_1' = \Gamma_2'$, then $T_2 <: T_1$.
+\end{enumerate}
+
+The relation is a partial order. The principal typing of $t$ is the strongest 
+typing of $t$, if it exists: $\forall \tau' \exists \tau . \tau' \leq \tau$.
+Given such a principal typing
+$\tau = (\Gamma,T,\Gamma')$, the type $T$ 
+is referred to as the {\it principal type} of the term $t$.
+The type of a variable $x \in dom(\Gamma)$ shall also be referred to as a 
+{\it principal variable input type} of $x$ in $t$. Similarly, the type
+of $x \in dom(\Gamma')$ shall be referred to as the
+{\it principal variable output type} of $x$ in $t$.
+
+\begin{figure}
+\includegraphics[width=0.49\textwidth]{principal.pdf}
+\caption{\label{fig:principalex} Valid types of function $f$}
+\end{figure}
+
+Given a term $t$ which interacts with an object variable $o$, the
+principal variable input type of $o$ is an object type which would permit
+all sequences of method calls to $o$ that could occur in $t$, and no more.
+Equivalently, the language accepted by the object type $o$ is the smallest 
+possible, while the return types of the methods of the states of $o$
+are as general as possible.
+
+Consider the body term of example function $f$ shown in 
+\figref{fig:principalex}. It is clear from the usage
+of $y$ that its principal variable input type must be $\boolt$ as demanded by the 
+type judgement of an if-then-else statement, while $x$ is an object.
+We could ascribe any of the object types $O_1@A$, $O_2@A$ and $O_3@A$ to the 
+variable $x$, and after evaluation the object types would be $O_1@A$, $O_2@\{E,H\}$
+and $O_2@F$ respectively.
+$O_1@A$ is not a principal variable input type for $x$ as it will permit many more 
+sequences of method calls 
+than strictly required, and also has a more
+specific return type for the method $n$ than is necessary. $O_2@A$ 
+is the principal variable input type for $x$, as is the type $O_3@A$, which is
+isomorphic. It is arguable that $O_3@A$ is a preferable solution to
+$O_2@A$, as it has a minimal object protocol. However as they are isomorphic
+it is of little consequence which is chosen, and if the latter is preferable
+for display to a user it can be computed on demand.
+
+% It is possible for a principal effect type to have more than
+% one state in the state set of the input (left) object type. It is
+% always possible to restructure such an object type to have a single initial
+% state. It is not, however, possible to have both a single input state and
+% single output state for a state machine accepting an arbitrary language --- 
+% consider the language accepted by the regular expression $m^* | n^*$. It is 
+% necessary for a state machine which accepts this language, and only this 
+% language, to
+% have at least three exit states. In general, it is more intuitive to have
+% a single input state in the effect type but this is not required for a
+% type to be considered principal.
+
+The principal typing of the body of function $f$ is \\ 
+$(\{x \mapsto O_3@A, y \mapsto \boolt \}, \top, \{ x \mapsto O_3@F, y \mapsto \boolt \})$.
+From this, it follows that the principal type of $f$ itself is \\
+$(O_3@A \gg O_3@F, \boolt \gg \boolt) \rightarrow \top$.
+
+Unfortunately, principal typings do not exist for all typeable terms.
+Consider the term $t = $ {\tt f(x)}. \\
+Let $\Gamma_1 = \{f \mapsto (\boolt \gg \boolt) \rightarrow \top, x \mapsto \boolt \}$. \\
+Let $\Gamma_2 = \{f \mapsto (\unitt \gg \unitt) \rightarrow \top, x \mapsto \unitt \}$. \\
+It follows that $\tau_1 = (\Gamma_1, \top, \Gamma_1)$
+and $\tau_2 = (\Gamma_2, \top, \Gamma_2)$ are typings of $t$.
+
+Assume a principal typing $\tau_3 = (\Gamma_3, T_3, \Gamma_3')$ exists for $t$.
+Therefore, $\tau_1 \leq \tau_3$ and $\tau_2 \leq \tau_3$.
+
+The domains of $\Gamma_1$ and $\Gamma_2$ are minimal, as the term $t$
+cannot be typed without both $f$ and $x$ in the input context by rule \tfunc.
+It follows that we must have
+$\Gamma_1(f) <: \Gamma_3(f)$ and $\Gamma_2(f) <: \Gamma_3(f)$. 
+Let $T' = \Gamma_3(f)$. For $\tau_3$ to be a typing, it must be the case that
+$T' = (U \gg V) \rightarrow W$ by rule \tfunc. By rule \subfn, it follows that
+$U <: \boolt$ and $U <: \unitt$. Therefore, $U <: \boolt \sqcap \unitt$.
+The type $\boolt \sqcap \unitt$ is not defined.
+Therefore, $\tau_3$ cannot exist.
+
+Principal typings do, however, exist for all
+terms with no free variables that are used as functions. This is proven
+in \thmref{thm:principal} in \appref{app:principal}.
 
@@
 
-Throughout much of the preceding text we have avoided writing object
-types using the full syntax, instead preferring to use diagrams. The full
-syntax for object types is verbose and difficult for a user to parse.
+The full syntax for object types is verbose and difficult for a user to parse.
 Having to write out object types in full would therefore obfuscate the
-intent of the code, particularly when there is such intuitively a
+intent of the code, particularly when there is a
 strong correlation between the code and the principal object types of the
-variables it interacts with. As such, the ability to infer principal object
-types is essential for any language which includes typestate, especially
-where structural rather than nominal subtyping is employed.
-
-The main source of type verbosity in TS0 is the requirement for
-type annotations on function literals. 
-Despite there being no need for alias control
-annotations, the effect declarations for each parameter are still verbose
-and the ability to define type aliases would only help partially.
-
-We therefore desire a general type inference strategy that can infer a
-minimal $\Gamma$, $\Gamma'$ and $\alpha$ for a term $t$ such that 
-$\typerule{\Gamma}{t}{\alpha}{\Gamma'}$. If $t$ is the body of a function
-literal, we can extract the effect types for the named parameters (of which
-the free variables of $t$ should be a subset) from $\Gamma$ and $\Gamma'$.
-
-Our approach to type inference is inspired by \cite{Pottier2005} which advocates
-separate constraint generation and constraint solving phases for
-type inference. Our constraint grammar is shown in \figref{fig:congrammar}.
+variables it interacts with. In a language such as TS where nominal typing
+is not used, typestate inference is essential.
+
+The main source of avoidable verbosity in TS is in the
+type annotations on function literals: one would hope that knowledge of
+the function body would be sufficient to determine the effect types for
+each parameter passed to the function. Indeed, if a principal typing can
+be derived for the function body, then a principal typing for the function
+literal can be directly constructed from this as shown in
+\thmref{thm:principal}.
+
+We therefore desire a general type inference strategy that can 
+deduce a principal typing for a term $t$, if one exists. Our approach to type 
+inference is inspired by \cite{Pottier2005}, using separate
+constraint generation and constraint solving phases 
+to deduce the principal typing of the root term.
+
+The constraint grammar is shown in \figref{fig:congrammar}.
 
@@
 
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% thm:soundness %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% checked 02/01/2013
+%%% depends on: valuectx, upgrade, remap
+
 \begin{thm}[Progress and Preservation]
+\label{thm:soundness}
 Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
 with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
-there exist $t'$ and $\mu'$ such that
+there exists a $t'$ and $\mu'$ such that
 $\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
@@
 \begin{proof}
-by induction on the structure of $t$. Assume $t$ is not a value. It is
-therefore of one of the following forms:
-
-\begin{itemize}
-\item $t = \lett{x}{t''}{t'}$. 
-It follows by rule \tlet that there exist $\Gamma_1, T', T''$
-such that
-$\typerule{\Gamma}{t''}{T'}{\Gamma_1}$ and that
-$\typerule{\Gamma_1, x : T'}{t'}{T}{\Gamma', x : T''}$. There are two
-possibilities:
-
-  \begin{itemize}
-  \item If $t''$ is some value $v$, then by \lemref{lem:valuectx} 
-  we have that $\Gamma = \Gamma_1$ and reduction can occur
-  by \rletv, such that 
-  $\oprule{\lett{x}{v}{t'}}{\mu}{t'}{\mu [ x \mapsto v ]}$.
-  Let $\mu' = \mu [ x \mapsto v ]$. Let $\Gamma'' = \Gamma, x : T'$.
-  It follows that that $\Gamma'' \vdash \mu'$, with
-  $\typerule{\Gamma''}{t'}{T}{\Gamma', x : T'}$.
-  The requirements of rule \tdrop are satisfied by this, therefore we can 
-  type $t'$ such that $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
-
-  \item If $t''$ is a term, 
-  by induction we have that $\oprule{t''}{\mu}{t'''}{\mu'}$, with
-  some $\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'''}{T'}{\Gamma_1}$.
-  This satisfies the requirements of \rlett, meaning $t$ itself can reduce
-  such that $\oprule{\lett{x}{t''}{t'}}{\mu}{\lett{x}{t'''}{t'}}{\mu'}$.
+by induction on the typing derivation of $t$. Assume $t$ is not a value. It has
+therefore been typed by one of the following rules:
+
+\begin{itemize}
+\item \tlet. It follows that $t = \lett{x}{t_v}{t_b}$, with
+$\typerule{\Gamma}{t_v}{T_v}{\Gamma_1}$ and
+$\typerule{\Gamma_1, x : T_v}{t_b}{T}{\Gamma', x : T_v'}$ for some
+$\Gamma_1$, $T_v$ and $T_v'$.
+There are two cases to consider:
+
+  \begin{itemize}
+  \item $t_v$ is a value.
+  Reduction can occur by \rletv such that 
+  $\oprule{\lett{x}{t_v}{t_b}}{\mu}{t_b}{\mu [ x \mapsto t_v ]}$.
+  Let $\mu' = \mu [ x \mapsto t_v ]$. Let $\Gamma'' = \Gamma_1, x : T_v$.
+  It follows that $\Gamma'' \vdash \mu'$ and by rule \tdrop that
+  $\typerule{\Gamma''}{t_b}{T}{\Gamma'}$.
+
+  \item If $t_v$ is not a value, 
+  by induction we have that $\oprule{t_v}{\mu}{t_v'}{\mu'}$, with
+  some $\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t_v'}{T_v}{\Gamma_1}$.
+  This satisfies the requirements of \rlett such that $t$ itself can be reduced:
+  $\oprule{\lett{x}{t_v}{t_b}}{\mu}{\lett{x}{t_v'}{t_b}}{\mu'}$.
   Additionally, the conditions of \tlet are satisfied such that
-  $\typerule{\Gamma''}{\lett{x}{t'''}{t'}}{T}{\Gamma'}$.
-
-  \end{itemize}
-
-\item $t = t_l ; t_r$. It follows by rule
-\tseq that $\typerule{\Gamma}{t_l}{T_l}{\Gamma_{mid}}$ and
-$\typerule{\Gamma_{mid}}{t_r}{T}{\Gamma'}$.
-
-  \begin{itemize}
-  \item If $t_l$ is some value $v$, then
-  reduction can occur by \\ 
+  $\typerule{\Gamma''}{\lett{x}{t_v'}{t_b}}{T}{\Gamma'}$.
+
+  \end{itemize}
+
+\item \tseq. It follows that $t = t_l ; t_r$ with
+$\typerule{\Gamma}{t_l}{T_l}{\Gamma_{mid}}$ and
+$\typerule{\Gamma_{mid}}{t_r}{T}{\Gamma'}$ for some $\Gamma_{mid}$ and $T_l$.
+There are two cases to consider:
+
+  \begin{itemize}
+  \item $t_l$ is a value. Reduction can therefore occur by
   \rseqlv such that
-  $t_l; t_r \mid \mu \longrightarrow t_r \mu$. By \lemref{lem:valuectx}, 
-  $\Gamma'' = \Gamma = \Gamma_{mid}$
-  and $\mu' = \mu$, therefore $\Gamma'' \vdash \mu'$ and 
-  $\typerule{\Gamma''}{t_r}{T}{\Gamma'}$.
-
-  \item If $t_l$ is a term, by induction $t_l \mid \mu \longrightarrow t_l' \mid \mu'$
+  $\oprule{t_l ; t_r}{\mu}{t_r}{\mu}$. 
+  Let $\Gamma'' = \Gamma_{mid}$, and $\mu' = \mu$.
+  Directly, $\typerule{\Gamma''}{t_r}{T}{\Gamma'}$.
+  By \lemref{lem:ctxsubset} it follows that $dom(\Gamma'') \subseteq dom(\Gamma)$,
+  and by \lemref{lem:valuectx} it follows that 
+  $\forall y \in dom(\Gamma'') . \Gamma(y) <: \Gamma''(y)$.
+  Therefore $\Gamma'' \vdash \mu'$.
+
+  \item If $t_l$ is a term, by induction $\oprule{t_l}{\mu}{t_l'}{\mu'}$
   with some $\Gamma'' \vdash \mu'$ and
@@
   This satisfies the requirements of \rseqlt
-  such that $t_l; t_r \mid \mu \longrightarrow t_l'; t_r \mid \mu'$.
+  such that $\oprule{t_l ; t_r}{\mu}{t_l' ; t_r}{\mu'}$.
   Additionally, the requirements of \tseq are satisfied such that
@@
 
-\item $t = x ( \overline{x_i} )$. It follows by rule \tfunc that
+\item \tfunc. It follows that $t = x ( \overline{x_i} )$ with
 $\Gamma = \Gamma_1, \overline{x_i : T_i}$ 
-with $\Gamma_1(x) = \funt{U_i}{V_i}{T}$
-such that $\overline{T_i <: U_i}$ 
+and $\Gamma_1(x) = \funt{U_i}{V_i}{T}$
+such that $\forall i . T_i <: U_i$ 
 and that $\Gamma' = \Gamma_1, \overline{x_i : T_i'}$
-where $T_i' = \mathbf{remap} ( T_i, U_i >> V_i )$.
+where $T_i' = \remapapp{T_i}{U_i}{V_i}$.
 
@@
 where $\typerule{\overline{y_i : U_i'}}{t'}{T'}{\overline{y_i : V_i'}}$,
-with $\funt{U_i'}{V_i'}{T'} <: \funt{U_i}{V_i}{T}$. It
-follows by \subfn that for each $i$, $U_i <: U_i'$, $V_i' <: V_i$
+with $\funt{U_i'}{V_i'}{T'} <: \funt{U_i}{V_i}{T}$. 
+Therefore by \subfn for each $i$, $U_i <: U_i'$, $V_i' <: V_i$
 and $T' <: T$.
@@
 The term $t$ can be reduced by \rfunc such that
-$x ( \overline{x_i} ) \mid \mu \longrightarrow t_{sub} \mid \mu$.
+$\oprule{t}{\mu}{t_{sub}}{\mu}$.
 By the substitution lemma, 
 $\typerule{\overline{x_i : U_i'}}{t_{sub}}{T'}{\overline{x_i : V_i'}}$.
-By lemma \lemref{lem:upgrading}, it follows that
-for each $i$, there exists a $V_i'' = \remapapp{T_i''}{U_i'}{V_i'}$
+As $U_i <: U_i'$ for each $i$, by repeated application of lemma 
+\lemref{lem:upgrade}, it follows that
+there exists a $T'' <: T'$ and $V_i'' <: V_i'$ for each $i$
 such that 
-$\typerule{\overline{x_i : U_i}}{T_{sub}}{T'}{\overline{x_i : V_i''}}$.
-By \lemref{lem:remap} for each $i$ that $V_i'' <: V_i'$.
+$\typerule{\overline{x_i : U_i}}{T_{sub}}{T''}{\overline{x_i : V_i''}}$.
 
 Therefore, by given that $T_i <: U_i$ and that $V_i'' <: V_i' <: V_i$
-and that $T' <: T$ we can apply \tnarrow, \twiden and \tsub
-such that $\typerule{\overline{x_i : T_i}}{t'}{T}{\overline{x_i : V_i}}$.
+and that $T'' <: T' <: T$ we can apply \tnarrow, \twiden and \tsub
+such that $\typerule{\overline{x_i : T_i}}{t_{sub}}{T}{\overline{x_i : V_i}}$.
 
@@
 $\Gamma'' \vdash \mu'$. By applying the weakening lemma, we can type 
-$t_{sub}$ such that $\typerule{\Gamma''}{t''}{T}{\Gamma'}$.
-
-\item $t = x.m$. It follows by rule \tmethc that
-$\Gamma = \Gamma_1, x : O@\overline{S}$ and that 
-$\Gamma' = \Gamma_1, x : O@\overline{S'}$, with
-$\overline{S'} = \{ S_n | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \}$.
+$t_{sub}$ such that $\typerule{\Gamma''}{t_{sub}}{T}{\Gamma'}$.
+
+\item \tmethc. It follows that $t = x.m$ with
+$\Gamma = \Gamma_1, x : O@\overline{S}$ and
+$\Gamma' = \Gamma_1, x : O@\overline{S'}$, where
+$m : T \Rightarrow \overline{S'} \in O@\overline{S}$.
+
 As
-$\Gamma \vdash \mu$, we must have $\mu = \mu_1, x \mapsto o'@S_c$ with
-$\Gamma_1 \vdash \mu_1$ and where
+$\Gamma \vdash \mu$, we must have $\mu(x) = o'@S_c$ with
 $\typerule{\emptyset}{o'@S_c}{O'@S_c}{\emptyset}$ such that
@@
 
-By the definition of subtyping for object types,
-we must have that
-$o' = [ S' \{ m = (v,S_n) ; ... \} ... ]$, where $O'@S_n <: O@\overline{S'}$. 
-Where the value $v$ is concerned, $\typerule{\emptyset}{v}{T'}{\emptyset}$
-for some $T'$ such that $T' <: T = \bigsqcup{ \{ T_r | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \} }$.
+We must have that
+$o' = [ S_c \{ m = (v,S_n) ; ... \} ... ]$, where $O'@S_n <: O@\overline{S'}$. 
+It follows that $\typerule{\emptyset}{v}{T'}{\emptyset}$
+for some $T'$ such that $T' <: T$.
 
 Therefore reduction can occur by \rmethc such that
-$x.m \mid \mu_1, x \mapsto o@S \longrightarrow v \mid \mu_1, x \mapsto o@S_n$.
-Let $\mu' = \mu_1, x \mapsto o@S'$. Let $\Gamma'' = \Gamma'$. It follows that
+$\oprule{t}{\mu}{v}{\mu [x \mapsto o@S_n]}$.
+Let $\mu' = \mu [ x \mapsto o@S']$.
+
+Let $\Gamma'' = \Gamma'$. By the appropriate value typing rule for $v$, 
+it follows that $\typerule{\Gamma''}{v}{T'}{\Gamma'}$.
+By application of rule \tsub, it follows that
+$\typerule{\Gamma''}{v}{T}{\Gamma'}$.
+Additionally, it follows that
 $\Gamma'' \vdash \mu'$, as $O'@S_n <: O@\overline{S'}$.
-By application of the weakening lemma and rule \tsub, it follows that
-$\typerule{\Gamma''}{v}{T}{\Gamma'}$.
-
-\item $t = \ift{t_c}{t_t}{t_f}$. It follows by rule \tif that there exists
-$\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$ and $T_f$ such that
-$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
-$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
-$\typerule{\Gamma_2}{t_f}{T_f}{\Gamma_3}$ with
+
+\item \tif. It follows that $t = \ift{t_c}{t_t}{t_f}$ with
+$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and
+$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and
+$\typerule{\Gamma_2}{t_f}{T_f}{\Gamma_3}$ where
 $\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$. There are
@@
   \item $t_c = \truev$. By \lemref{lem:valuectx} we have that
-  $\Gamma = \Gamma_1$. Reduction can occur by \riftrue such that
+  $\forall x \in dom(\Gamma_1) . \Gamma(x) <: \Gamma_1(x)$, and
+  by \lemref{lem:ctxsubset} that $dom(\Gamma_1) \subseteq dom(\Gamma)$. 
+  Reduction can occur by \riftrue such that
   $\oprule{t}{\mu}{t_t}{\mu}$. By \tsub we have that
@@
   $\typerule{\Gamma_1}{t_t}{T}{\Gamma'}$.
-  Let $\Gamma'' = \Gamma$, $\mu' = \mu$ and 
-  $t' = t_t$. Therefore, $\typerule{\Gamma''}{t'}{T}{\Gamma'}$ with
-  $\Gamma'' \vdash \mu'$.
+  Let $\Gamma'' = \Gamma_1$ and $\mu' = \mu$.
+  It follows that $\Gamma'' \vdash \mu'$
+  and that $\typerule{\Gamma''}{t_t}{T}{\Gamma'}$.
 
   \item $t_c = \falsev$. By \lemref{lem:valuectx} we have that
-  $\Gamma = \Gamma_1$. Reduction can occur by \riffalse such that
+  $\forall x \in dom(\Gamma_1) . \Gamma(x) <: \Gamma_1(x)$, and
+  by \lemref{lem:ctxsubset} that $dom(\Gamma_1) \subseteq dom(\Gamma)$.
+  Reduction can occur by \riffalse such that
   $\oprule{t}{\mu}{t_f}{\mu}$. By \tsub we hhave that
@@
   repeatedly applying \twiden for all $x$, we have that
-  $\typerule{\Gamma_1}{t_f}{T}{\Gamma'}$. Let $\Gamma'' = \Gamma$,
-  $\mu' = \mu$ and $t' = t_f$. Therefore, 
-  $\typerule{\Gamma''}{t'}{T}{\Gamma'}$ with $\Gamma'' \vdash \mu'$.
+  $\typerule{\Gamma_1}{t_f}{T}{\Gamma'}$. Let $\Gamma'' = \Gamma_1$
+  and $\mu' = \mu$. It follows that $\Gamma'' \vdash \mu'$ and that
+  $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
 
@@
 
+\item \tdrop. It follows that there exists a $\Gamma'''$ such that
+$\typerule{\Gamma}{t}{T}{\Gamma', \Gamma'''}$. By induction, there exists
+a $t'$, $\Gamma''$ and $\mu'$ such that
+$\oprule{t}{\mu}{t'}{\mu'}$ with
+$\typerule{\Gamma''}{t'}{T}{\Gamma', \Gamma'''}$ where $\Gamma'' \vdash \mu'$.
+By application of \tdrop for each $x \in dom(\Gamma''')$, it follows that
+$\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
+
+\item \tsub. It follows that there exists a $T' <: T$ such that
+$\typerule{\Gamma}{t}{T'}{\Gamma'}$. By induction, there exists a $t'$,
+$\Gamma''$ and $\mu'$ such that 
+$\oprule{t}{\mu}{t'}{\mu'}$ with
+$\typerule{\Gamma''}{t'}{T'}{\Gamma'}$ where
+$\Gamma'' \vdash \mu'$. By application of \tsub, it follows that
+$\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
+
+\item \twiden. It follows that $\Gamma' = \Gamma_1, x : U$ and that there
+exists a $T <: U$ such that $\typerule{\Gamma}{t}{T}{\Gamma_1, x : T}$.
+By induction, there exists a $t'$, $\Gamma''$ and $\mu'$ such that
+$\oprule{t}{\mu}{t'}{\mu'}$ with $\typerule{\Gamma''}{t'}{T}{\Gamma_1, x : T}$
+where $\Gamma'' \vdash \mu'$. By application of \twiden, it follows that
+$\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
+
+\item \tnarrow. It follows that $\Gamma = \Gamma_1, x : T$ and that there
+exists a $U$ such that $T <: U$ and that $\typerule{\Gamma_1, x : U}{t}{T}{\Gamma'}$.
+By induction, there exists a $t'$, $\Gamma''$ and $\mu'$ such that
+$\oprule{t}{\mu}{t'}{\mu'}$ with $\typerule{\Gamma''}{t'}{T}{\Gamma'}$ where
+$\Gamma'' \vdash \mu'$.
+
 \end{itemize}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:ctxsubset %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% last revised and checked: 2012/12/13
+
+\begin{lem}
+\label{lem:ctxsubset}
+Given an arbitrary term $t$, it follows that for any judgement
+$\typerule{\Gamma}{t}{T}{\Gamma'}$ that $dom(\Gamma') \subseteq dom(\Gamma)$.
+\end{lem}
+\begin{proof}
+by induction on the typing derivation of $t$.
+
+\begin{itemize}
+\item $t$ is typed by one of \tunit, \ttrue, \tfalse,
+\tobj or \tfundef. It follows that $\Gamma = \Gamma'$. Trivially,
+$dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tlet. Therefore $t = \lett{x}{t_v}{t_b}$
+such that $\typerule{\Gamma}{t_v}{T_v}{\Gamma''}$ and
+$\typerule{\Gamma'', x : T_v}{t_b}{T}{\Gamma''', x : T_v'}$
+with $\Gamma' = \Gamma'''$.
+By induction, $dom(\Gamma'') \subseteq dom(\Gamma)$ and $dom(\Gamma''') \subseteq dom(\Gamma'')$.
+Therefore, $dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tseq. Therefore $t = t_a ; t_b$ such that
+$\typerule{\Gamma}{t_a}{T_a}{\Gamma''}$ and
+$\typerule{\Gamma''}{t_b}{T}{\Gamma'}$. By induction,
+$dom(\Gamma'') \subseteq dom(\Gamma)$ and $dom(\Gamma') \subseteq dom(\Gamma'')$. Therefore
+$dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tfunc. It follows that $dom(\Gamma') = dom(\Gamma)$,
+therefore $dom(\Gamma') \subseteq dom(\Gamma)$
+
+\item $t$ is typed by \tmethc. It follows that $dom(\Gamma') = dom(\Gamma)$,
+therefore $dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tif. If follows that $t = \ift{t_c}{t_t}{t_f}$
+with $\typerule{\Gamma}{t_c}{\boolt}{\Gamma_c}$
+and $\typerule{\Gamma_c}{t_t}{T_t}{\Gamma_t}$
+and $\typerule{\Gamma_c}{t_f}{T_f}{\Gamma_f}$ such that
+$\Gamma' = \Gamma_t \sqcup \Gamma_f$.
+
+By induction, $dom(\Gamma_c) \subseteq dom(\Gamma)$ and
+$dom(\Gamma_t) \subseteq dom(\Gamma_c)$ and
+$dom(\Gamma_f) \subseteq dom(\Gamma_c)$. $\Gamma' = \Gamma_t \sqcup \Gamma_f$
+is only defined where $dom(\Gamma_t) = dom(\Gamma_f)$ therefore
+$dom(\Gamma') = dom(\Gamma_t)$. It follows that
+$dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tdrop. Directly, $dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\item $t$ is typed by \tsub, \twiden or \tnarrow. In each case
+$dom(\Gamma') = dom(\Gamma)$. Trivially, $dom(\Gamma') \subseteq dom(\Gamma)$.
+
+\end{itemize}
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:unused %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}
+\label{lem:unused}
+Given an arbitrary term $t$ such that 
+$\typerule{\Gamma, x : T}{t}{T}{\Gamma', x : V}$ where $x$ is unused in
+$t$, $T <: V$.
+\end{lem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:valuectx %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% last revised and checked: 2012/12/13
 
@@
 \label{lem:valuectx}
-If $v$ is a value and $\typerule{\Gamma}{v}{T}{\Gamma'}$, then $\Gamma = \Gamma'$.
+Given an arbitrary value $v$, it follows that for any judgement
+$\typerule{\Gamma}{v}{T}{\Gamma'}$ that 
+$\forall x \in \Gamma' . \Gamma(x) <: \Gamma'(x)$.
 \end{lem}
 \begin{proof}
-Directly from the typing judgements. $v$ is either {\tt unit}, {\tt true},
-{\tt false}, a function literal or an object literal.
-Correspondingly, by rules \tunit,
-\ttrue, \tfalse, 
-\tfundef and \tobj, $\Gamma = \Gamma'$.
+by induction on the typing derivation of $v$.
+
+\begin{itemize}
+\item $v$ is typed by one of \tunit, \ttrue, \tfalse,
+\tobj or \tfundef. It follows that $\Gamma' = \Gamma$. Trivially,
+$\forall x \in \Gamma' . \Gamma(x) <: \Gamma'(x)$.
+
+\item $v$ is typed by \tdrop. It follows that $\typerule{\Gamma}{v}{T}{\Gamma''}$
+where $\Gamma'' = \Gamma', \Gamma'''$. By induction,
+$\forall x \in \Gamma'' . \Gamma(x) <: \Gamma''(x)$.
+By definition, $\forall x \in \Gamma' . \Gamma'(x) = \Gamma''(x)$.
+Therefore, $\forall x \in \Gamma' . \Gamma(x) <: \Gamma'(x)$.
+
+\item $v$ is typed by \tsub. It follows that $\Gamma = \Gamma'$. Trivially,
+$\forall x \in \Gamma' . \Gamma(x) <: \Gamma'(x)$.
+
+\item $v$ is typed by \twiden. It follows that $\typerule{\Gamma}{v}{T}{\Gamma'''}$
+where $\Gamma''' = \Gamma'', x : T$ and $\Gamma' = \Gamma'', x : U$ with
+$T <: U$. By induction, $\forall x \in \Gamma''' . \Gamma(x) <: \Gamma'''(x)$.
+By definition, $\forall x \in \Gamma' . \Gamma'''(x) <: \Gamma'(x)$. Transitively,
+$\forall x \in \Gamma' . \Gamma(x) <: \Gamma'(x)$.
+
+\item $v$ is typed by \tnarrow. It follows that $\typerule{\Gamma''}{v}{T}{\Gamma'}$,
+where $\Gamma'' = \Gamma''', x : U$ and $\Gamma = \Gamma''', x : T$ where
+$T <: U$. By induction, $\forall x \in \Gamma' . \Gamma''(x) <: \Gamma'(x)$.
+By definition, $\forall x \in \Gamma . \Gamma(x) <: \Gamma''(x)$ and
+$dom(\Gamma'') = dom(\Gamma)$. Therefore
+$\forall x \in \Gamma . \Gamma(x) <: \Gamma'(x)$.
+
+\end{itemize}
 \end{proof}
 
-\begin{lem}[Object state sets]
-Given an object state set $\overline{S}$ and a state $S'$, if
-$S' \in \overline{S}$ then $O@S' <: O@\overline{S}$.
-\end{lem}
-\begin{proof}
-$O@\overline{S}$ is defined as equivalent to
-$\bigsqcup_{S \in \overline{S}}{O@S}$. Given that $S \in \overline{S}$,
-we must have that $\overline{S} = \overline{S'} \cup \{ S \}$ for
-some $\overline{S'}$.
-it follows that $O@\overline{S} = O@S' \sqcup \left( 
-\bigsqcup_{S \in \overline{S'}}{O@S} \right)$.
-By the definition of join, 
-$O@S' <: O@S' \sqcup \bigsqcup_{S \in \overline{S}}{O@S}$.
-\end{proof}
-
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% last revised and checked: 2012/12/13
 
@@
 \begin{itemize}
-\item $V = \unitt$ or $V = \boolt$. By either \remapunit or \remapbool
+\item $V = \unitt$ or $V = \boolt$. By either rule \remapunit or \remapbool
 it must be the case that $U = V$. It must also be the case that
@@
 
-\item $V = \funv{V_i}{V_i'}{V'}$ for some $\overline{V_i}$, $\overline{V_i'}$
-and $V'$. By \remapfunc, $T <: U$ and $T = \remapapp{T}{U}{V}$. Therefore,
-$T <: V$.
-
-\item $V = O@\overline{S'}$ for some $O$ and $\overline{S'}$. By
-\remapobj, $O'@\overline{S'''} = \remapapp{T}{U}{V}$ where 
-$T = O'@\overline{S''}$ and $U = O@\overline{S}$ and
-$\overline{S'''} = \{ S | O'@S \precsim O@\overline{S'}, S \in O' \}$ 
-for some $O'$, $\overline{S}$ and $\overline{S''}$. By \subobj,
-$T = O'@\overline{S'''} <: O@\overline{S'} = V$ as for each $S \in \overline{S'''}$,
-$O'@S <: O@\overline{S'}$.
+\item $V = \funv{V_i}{V_i'}{V'}$. By \remapfunc, 
+$T <: U$, $U = V$ and $T = \remapapp{T}{U}{V}$. Therefore, $T <: V$.
+
+\item $V = \ot{O_1}{S_3}$. By
+\remapobj, $T = \ot{O_2}{S_2}$ and $U = \ot{O_1}{S_1}$
+where $T <: U$. It follows that $\ot{O_1}{S_4} = \remapapp{T}{U}{V}$
+where 
+$\overline{S_4} = \bigcup \{ \overline{S} \mid (\ot{O_2}{S_2}, \ot{O_1}{S_1}) \lesssim (\ot{O_2}{S}, \ot{O_1}{S_3}) \}$. By definition, $\ot{O_1}{S_4} <: \ot{O_2}{S_3}$.
+
 \end{itemize}
@@
 \label{lem:remapcv}
-Given types $T, T', U$ and $V$ such that $T <: T'$, 
+Given types $T, T', U$ and $V$ such that $T <: T'$ and
+$\remapapp{T}{U}{V}$ is defined,
 $\remapapp{T}{U}{V} <: \remapapp{T'}{U}{V}$.
 \end{lem}
+\begin{proof}
+by case analysis of type $U$.
+
+\begin{itemize}
+\item $U = \boolt$ or $\unitt$. By \remapunit or \remapbool, it follows
+that $T = \remapapp{T}{U}{V}$. 
+As $\boolt$ and $\unitt$ have no subtypes other than
+themselves, it follows that $T' = T$. Trivially,
+$\remapapp{T'}{U}{V} = \remapapp{T}{U}{V}$.
+
+\item $U = \funt{U_i}{U_i'}{U'}$. By \remapfunc it follows that
+$\remapapp{T'}{U}{V} = T' <: T = \remapapp{T}{U}{V}$.
+
+\item $U = \ot{O_2}{S_2}$, with $T = \ot{O_1}{S_1}$ and $V = \ot{O_2}{S_4}$
+and $\ot{O_1}{S_3} = \remapapp{T}{U}{V}$. Let $T' = \ot{O_3}{S_5}$,
+and $\ot{O_3}{S_6} = \remapapp{T'}{U}{V}$. By definition,
+$\overline{S_3} = \bigcup \{ \overline{S} \mid (\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S}, \ot{O_2}{S_4}) \}$, and
+$\overline{S_6} = \bigcup \{ \overline{S'} \mid (\ot{O_3}{S_5}, \ot{O_2}{S_2}) \lesssim (\ot{O_3}{S}, \ot{O_2}{S_4}) \}$.
+In both cases, each $\overline{S}$ and $\overline{S'}$ extracted from
+the path relation must be such that 
+$(\ot{O_3}{S_5}, \ot{O_1}{S_1}) \lesssim (\ot{O_3}{S'}, \ot{O_1}{S})$.
+Therefore, $\ot{O_3}{S_6} <: \ot{O_1}{S_3}$.
+
+\end{itemize}
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_in_contra %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}
+\label{lem:remap_in_contra}
+Given types $T, U, U'$ and $V$ such that $U' <: U$ and that
+both $\remapapp{T}{U'}{V}$ and $\remapapp{T}{U}{V}$ are defined,
+$\remapapp{T}{U}{V} <: \remapapp{T}{U'}{V}$.
+\end{lem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_out_cv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}
+\label{lem:remap_out_cv}
+Given types $T, U, V$ and $V'$ such that $V' <: V$ and that
+both $\remapapp{T}{U}{V}$ and $\remapapp{T}{U}{V'}$ are defined,
+$\remapapp{T}{U}{V'} <: \remapapp{T}{U}{V}$.
+\end{lem}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:lesssim_same %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}
+\label{lem:lesssim_same}
+Let $\overline{S'}$ and $\overline{S''}$ be state sets such that
+$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S'}, \ot{O}{S''})$. It follows
+that $\overline{S'} = \overline{S''}$.
+\end{lem}
+
+\begin{proof}
+TODO (coinductive?)
+\end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\begin{lem}
+\label{lem:remapsame}
+Given a type $T$, $T = \remapapp{T}{T}{T}$.
+\end{lem}
+
+\begin{proof}
+by case analysis of the type $T$.
+
+\begin{itemize}
+\item $T = \unitt$ or $T = \boolt$. 
+Directly by rule \remapunit or \remapbool, $T = \remapapp{T}{T}{T}$
+
+\item $T = \funt{T_i}{T_i'}{T'}$. Directly by rule
+\remapfunc, $T = \remapapp{T}{T}{T}$.
+
+\item $T = \ot{O}{S_1}$. Let $\ot{O}{S_2} = \remapapp{T}{T}{T}$. By definition,
+$\overline{S_2} = \bigcup \{ \overline{S} \mid (\ot{O}{S_1}, \ot{O_1}{S_1}) \lesssim (\ot{O_1}{S}, \ot{O_1}{S_1}) \}$. By \lemref{lem:lesssim_same}, it follows that
+$\overline{S} = \overline{S_1}$ for all $\overline{S}$. Therefore,
+$\overline{S_2} = \overline{S_1}$.
+\end{itemize}
+\end{proof}
 
@@
 \label{lem:remapeff}
-Given types $T$, $U$ and $V$ such that $U <: V$ and $T = \remapapp{U}{V}{V}$, 
-it follows that $U <: T$. \\
-(TODO: object portion needs reworked due to new definition of remap)
+Given types $T$, $U$ and $V$ such that $U <: V$.
+It follows that $U <: \remapapp{U}{V}{V}$.
 \end{lem}
@@
 \begin{itemize}
-\item $V = \unitt$ or $V = \boolt$. 
-It must be the case that $V = U$ also, as $\unitt$ and $\boolt$
-have no subtypes other than themselves by rule \subrefl.
-Therefore, by rule \remapunit or \remapbool, $T = V$,
-and $U <: T$ by rule \subrefl.
-
-\item $V = \funv{T_i}{T_i'}{T_r}$ for some $\overline{T_i}$, $\overline{T_i'}$
-and $T_r$. As $U <: V$, it follows by rule \remapfunc that
-$T = U$, and $U <: T$ by \subrefl.
-
-\item $V = O@\overline{S}$ for some $O$ and $\overline{S}$. 
-It follows that $U = O'@\overline{S'}$ 
-for some $O'$ and $\overline{S'}$ 
-such that $O'@\overline{S'} \precsim O@\overline{S}$ 
-by rule \subobj. 
-By rule \remapobj, it follows that $T = O'@\overline{S''}$ 
-such that $\overline{S''} = \{ S | O'@S \precsim O@\overline{S}, S \in O' \}$.
-By \lemref{lem:ssetinc} it follows that 
-$O'@\overline{S'} <: O'@\overline{S''}$.
+\item $V = \unitt$ or $V = \boolt$. By rule \remapunit or \remapbool,
+$\remapapp{U}{V}{V} = U$. By \subrefl $U <: U$.
+
+\item $V = \funt{T_i}{T_i'}{T_r}$. By rule \remapfunc
+$\remapapp{U}{V}{V} = U$. By \subrefl $U <: U$.
+
+\item $V = \ot{O_2}{S_2}$, with $U = \ot{O_1}{S_1}$.
+By rule \remapobj, $\remapapp{\ot{O_1}{S_1}}{\ot{O_2}{S_2}}{\ot{O_2}{S_2}} = \ot{O_1}{S_3}$ where
+$\overline{S_3} = \bigcup \{ \overline{S} \mid (\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S}, \ot{O_2}{S_2}) \}$.
+By rule $\lesssim$\textsc{-refl},
+$(\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S_1}, \ot{O_2}{S_2})$,
+therefore $\overline{S_1} \subseteq \overline{S_3}$. It follows that
+$\ot{O_1}{S_1} <: \ot{O_1}{S_3}$ by \lemref{lem:ssetinc}.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{cor}
-\label{lem:remapsame}
-Given a type $T$, $T <: \remapapp{T}{T}{T}$. \\
-(TODO: in fact, with the new definition of remap it should be possible to
-state that $T = \remapapp{T}{T}{T}$).
-\end{cor}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:upgrade %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{lem}[remap]
+%%% Revised and checked 2012/12/13
+%%% DEPENDS ON: unused, weakening, remap
+
+\begin{lem}
 \label{lem:upgrade}
@@
 $\typerule{\Gamma, x : T}{t}{U}{\Gamma', x : V}$, and if
-$T' <: T$, then there exists a $V' <: \remapapp{T'}{T}{V}$ 
-such that $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
+$T' <: T$, then there exists a $V' <: V$
+and $U' <: U$
+such that $\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
 \end{lem}
 \begin{proof}
-by induction on the structure of $t$.
+by induction on the typing derivation of $t$.
 The term $t$ is of one of the following forms:
@@
 \begin{itemize}
-\item $t$ is a value, so by \lemref{lem:valuectx} it follows that
-$\Gamma = \Gamma'$ and $T = V$. 
-The environment can be arbitrary in the typing rules for values,
+\item $t$ is typed by one of \tunit, \ttrue, \tfalse, \tobj or \tfundef.
+Therefore, $T = V$.
+The environment can be arbitrary in these typing rules,
 therefore we can type $t$ such that 
 $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : T'}$.
-Let $V' = T'$.
-By \lemref{lem:remapeff}, we have that $V' <: \remapapp{T'}{T}{V}$.
-Therefore, $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-
-\item $t = \lett{y}{t'}{t''}$. It follows from the rule \tlet that
-$x \neq y$ (as variables cannot be rebound) and that there exists 
-$\Gamma_1, \Gamma_2, T_2, T_y, T_y'$ such that
-$\typerule{\Gamma, x : T}{t'}{T_y}{\Gamma_1, x : T_2}$ and that
-$\typerule{\Gamma_1, x : T_2, y : T_y}{t''}{U}{\Gamma', x : V, y : T_y'}$.
-By induction, it follows that there exists a type $V_2 <: \remapapp{T'}{T}{T_2}$ such that
-$\typerule{\Gamma, x : T'}{t'}{T_y}{\Gamma_1, x : V_2}$.
-By \lemref{lem:remap} we have that $V_2 <: T_2$.
-Therefore it also follows by induction that
-there exists a type $V_3 <: \remapapp{V_2}{T_2}{V}$
+Let $V' = T'$, and $U' = U$.
+
+\item $t$ is typed by \tlet. Therefore $t = \lett{y}{t'}{t''}$ and
+$x \neq y$ (as variables cannot be rebound), with
+$\typerule{\Gamma, x : T}{t'}{Y}{\Gamma_1, x : T_2}$ and
+$\typerule{\Gamma_1, x : T_2, y : Y}{t''}{U}{\Gamma', x : V, y : Y_2}$
+for some $\Gamma_1, T_2, Y$ and $Y_2$.
+
+By induction, it follows that there exists a $T_3 <: T_2$ 
+and $Y_3 <: Y$
 such that
-$\typerule{\Gamma_1, x : V_2, y : T_y}{t''}{U}{\Gamma', x : T_3, y : V_3}$.
-Let $V' = V_3$. The rules of \tlet are satisfied such that
+$\typerule{\Gamma, x : T'}{t'}{Y_3}{\Gamma_1, x : T_3}$.
+
+By induction, it follows that there exists a type $T_4 <: V$
+and $U_2 <: U$
+such that
+$\typerule{\Gamma_1, x : T_3, y : Y}{t''}{U_2}{\Gamma', x : T_4, y : Y_2}$.
+By a second application of the induction hypothesis, it follows that there
+exists a $Y_4 <: Y_2$
+and $U_3 <: U_2$ such that
+$\typerule{\Gamma_1, x : T_3, y : Y_2}{t''}{U_3}{\Gamma', x : T_4, y : Y_4}$.
+
+Let $V' = T_4$. Let $U' = U_3$. By rule \tlet
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
+
+\item $t$ is typed by \tfunc. Therefore $t = y ( \overline{z_i} )$.
+There are three cases to consider:
+
+\begin{itemize}
+	\item $x \neq y$ and $x \notin \overline{z_i}$, meaning that
+	$x$ is unused in this function call. By \lemref{lem:unused}
+	it follows that $T <: V$, and that
+	$\typerule{\Gamma}{t}{U}{\Gamma'}$.
+	By \lemref{lem:weakening},
+	$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : T'}$.
+	Let $V' = T'$ and $U' = U$.
+	Therefore
+	$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$
+	with $V' <: V$ (transitively) and $U' <: U$ (reflexively).
+
+	\item $x = y$.
+	Therefore $T = \funt{T_i}{T_i'}{U}$, with
+	$\Gamma(z_i) = T_i$ and $\Gamma'(z_i) = T_i'$ for each $i$.
+	Additionally, $T = V$ by \tfunc.
+
+	$T'$ must therefore be of form $\funt{W_i}{W_i'}{W}$ where
+	$T_i <: W_i$ and $W_i' <: T_i'$ and $W <: U$
+	by \subfn. Let $V' = T'$. Let $U' = W$.
+	Let $\Gamma''$ be a context such that
+
+		\begin{itemize}
+		\item $dom(\Gamma'') = dom(\Gamma')$
+		\item $\Gamma''(z_i) = \remapapp{\Gamma(z_i)}{W_i}{W_i'}$ for each $z_i$.
+		\item $\forall a \notin \overline{z_i} . \Gamma''(a) = \Gamma'(a)$
+		\end{itemize}
+
+	It follows by rule \tfunc that $\typerule{\Gamma, x : T'}{t}{U'}{\Gamma'', x : T'}$.
+	By \lemref{lem:remap}, $\Gamma''(z_i) <: W_i' <: \Gamma'(z_i)$ for each
+	$z_i$. Therefore by repeated application of \twiden for each $z_i$,
+	$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$,
+	with $U' <: U$ and $V <: V$.
+
+	%%% V' = remap(T', T >> V)
+	\item $x \in \overline{z_i}$. 
+	By \tfunc it follows that $\Gamma(y) = \funt{U_i}{V_i}{U}$
+	and that $T = U_i$ and $V = V_i$ for some $i$.
+	Let $V' = \remapapp{T'}{T}{V}$ and $U' = U$. It follows that
+	$\typerule{\Gamma, x : T'}{t}{U}{\Gamma, x : V'}$,
+	with $U' <: U$ and
+	$V' <: V'$ by \lemref{lem:remap}.
+\end{itemize}
+
+%%% V' = T' or is a subtype of remap(T', T >> V)
+\item $t$ is typed by \tmethc. Therefore $t = y.m$.
+There are two cases to consider:
+
+\begin{itemize}
+	\item $x = y$. It follows that $T = \ot{O_1}{S_1}$, and that
+	$V = \ot{O_1}{S_3}$, with $m : U \Rightarrow \overline{S_3} \in \ot{O_1}{S_1}$.
+	It must therefore be the case that $T' = \ot{O_2}{S_2}$ where
+	$\ot{O_2}{S_2} <: \ot{O_1}{S_1}$. Therefore,
+	$m : U' \Rightarrow \overline{S_4} \in \ot{O_2}{S_2}$ with
+	$U' <: U$. Let $V' = \ot{O_2}{S_4}$. By the definition of object subtyping,
+	$V' <: V$.
+	By rule \tmethc, $\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
+
+	\item $x \neq y$, meaning that $x$ is unused by this method call.
+	By \lemref{lem:unused} $T <: V$. By \tmethc it follows that
+	$\typerule{\Gamma}{t}{U}{\Gamma'}$.
+
+	By \lemref{lem:weakening},
+	$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : T'}$.
+	Let $V' = T'$ and $U' = U$. Reflexively, $V' <: V$ and $U' <: U$.
+
+\end{itemize}
+
+\item $t$ is typed by \tlet. It follows that $t = t_a ; t_b$ and 
+that there exists
+$\Gamma'', T_{mid}$ and $T_a$ 
+such that $\typerule{\Gamma, x : T}{t_a}{T_a}{\Gamma'', x : T_{mid}}$
+and that $\typerule{\Gamma'', x : T_x}{t_b}{U}{\Gamma', x : V}$.
+
+By induction, it follows that there exists $T_{mid}' <: T_{mid}$
+and $T_a' <: T_a$
+such that $\typerule{\Gamma, x : T'}{t_a}{T_a'}{\Gamma'', x : T_{mid}'}$.
+
+By induction, it follows that 
+there exists $V' <: V$ 
+and $U' <: U$
+such that
+$\typerule{\Gamma'', x : T_{mid}'}{t_b}{U'}{\Gamma'', x : V'}$.
+
+By rule \tlet it follows that
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma'', x : V'}$.
+
+\item $t$ is typed by \tif. It follows that $t = \ift{t_c}{t_t}{t_f}$ 
+such that:
+
+\begin{itemize}
+\item $\typerule{\Gamma, x : T}{t_c}{\boolt}{\Gamma_1, x : V_c}$
+\item $\typerule{\Gamma_1, x : V_c}{t_t}{U_t}{\Gamma_2, x : V_t}$
+\item $\typerule{\Gamma_1, x : V_c}{t_f}{U_f}{\Gamma_3, x : V_f}$
+\end{itemize}
+
+Additionally, $\Gamma' = \Gamma_2 \sqcup \Gamma_3$, 
+$U = U_t \sqcup U_f$ and
+$V = V_t \sqcup V_f$.
+
+By induction, there exists a $V_c' <: V_c$
+and $U_c <: \boolt$
+such that $\typerule{\Gamma, x : T'}{t_c}{U_c}{\Gamma_1, x : V_c'}$.
+Trivially, $U_c = \boolt$.
+
+By induction there exists a 
+$V_t' <: V_t$ and $U_t' <: U_t$ such that
+$\typerule{\Gamma_1, x : V_c'}{t_t}{U_t'}{\Gamma_2, x : V_t'}$.
+
+By induction there exists a 
+$V_f' <: V_f$ and $U_f' <: U_f$
+such that $\typerule{\Gamma_1, x : V_c'}{t_f}{U_f'}{\Gamma_3, x : V_f'}$.
+
+Let $V' = V_t' \sqcup V_f'$, and $U' = U_t' \sqcup U_f'$.
+By the definition of type join, $V' <: V$ and $U' <: U$.
+By rule \tif it follows that
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
+
+\item $t$ is typed by \tdrop. It follows that
+$\typerule{\Gamma, x : T}{t}{U}{\Gamma''', x : V}$ where
+$\Gamma''' = \Gamma', \Gamma''$ for some $\Gamma''$. By induction,
+there exists a $V' <: V$ and $U' <: U$ such that
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma''', x : V'}$. By
+rule \tdrop, it follows that
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
+
+\item $t$ is typed by \tsub. It follows that
+$\typerule{\Gamma, x : T}{t}{W}{\Gamma', x : V}$ for some $W <: U$.
+By induction, there exists a $V' <: V$ and $U' <: W$
+such that
+$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$. Transitively,
+$U' <: U$.
+
+\item $t$ is typed by \twiden. There are two possibilities:
+
+	\begin{itemize}
+	\item $x$ is the subject of widening in the output context.
+	Therefore $\typerule{\Gamma, x : T}{t}{U}{\Gamma', x : W}$ where $W <: V$.
+	By induction, there exists a $W' <: W$ and $U' <: U$
+	$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : W'}$. Let $V' = W'$.
+
+	\item $x$ is not the subject of widening. Therefore
+	$\typerule{\Gamma, x : T}{t}{U}{\Gamma'', x : V}$ such that
+	$\Gamma'' = \Gamma''', y : Y'$ and $\Gamma' = \Gamma''', y : Y$
+	where $Y' <: Y$.
+
+	By induction, there exists a $V' <: V$ and
+	$U' <: U$ such that
+	$\typerule{\Gamma, x : T'}{t}{U'}{\Gamma'', x : V'}$.
+
+	By application of \twiden on $y$, it follows that
+	$\typerule{\Gamma}{t}{U'}{\Gamma', x : V'}$.
+
+	\end{itemize}
+
+\item $t$ is typed by \tnarrow. There are two possibilities:
+
+	\begin{itemize}
+
+	\item $x$ is the subject of narrowing in the input context.
+	Therefore $\typerule{\Gamma, x : W}{t}{U}{\Gamma', x : V}$ where $T <: W$.
+	Transitively, $T' <: W$.
+	By induction, there exists a 
+	$V' <: V$ and $U' <: U$
+	such that $\typerule{\Gamma, x : T'}{t}{U'}{\Gamma', x : V'}$.
+
+	\item $x$ is not the subject of narrowing in the input context.
+	Therefore $\typerule{\Gamma'', x : T}{t}{U}{\Gamma', x : V}$ where
+	$\Gamma'' = \Gamma''', y : Y$ and $\Gamma = \Gamma''', y : Y'$ where
+	$Y' <: Y$.
+
+	By induction, there exists a $V' <: V$ and $U' <: U$
+	such that $\typerule{\Gamma'', x : T'}{t}{U}{\Gamma', x : V'}$.
+	By application of \tnarrow to $y$, 
 	$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
 
-\item $t = y ( \overline{z_i} )$. 
-It follows from the rule \tfunc that there exists
-$\Gamma_1, \overline{T_i}, \overline{U_i}, \overline{V_i}, \overline{T_i'}$ 
-such that 
-$\Gamma, x : T = \Gamma_1, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i}$
-and that
-$\Gamma', x : V = \Gamma_1, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i'}$
-such that for each $i$, $T_i <: U_i$, $T_i' <: V_i$ 
-and $U <: U'$.
-
-There are three cases to consider:
-\begin{itemize}
-\item $x \neq y$ and $x \notin \overline{x_i}$, meaning that
-$\Gamma_1 = \Gamma_2, x : T$ for some $\Gamma_2$, that
-$\Gamma = \Gamma_2, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i}$
-and that $\Gamma' = \Gamma_2, y : \funt{U_i}{V_i}{U}, \overline{z_i : T_i'}$.
-Additionally, $T = V$. 
-It follows that $t$ can be typed without $x$
-such that $\typerule{\Gamma}{t}{U}{\Gamma'}$. Let $V' = T'$.
-By the weakening lemma, $\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.
-
-\item $x = y$, meaning $\Gamma = \Gamma_1, \overline{z_i : T_i}$ and
-$\Gamma' = \Gamma_1, \overline{z_i : T_i'}$. 
-Additionally, $T = V = \funt{U_i}{V_i}{U}$.
-$T'$ must therefore be of form $\funt{U_i'}{V_i'}{U'}$ 
-such that for each $i$, $U_i <: U_i'$, $V_i' <: V_i$, $U' <: U$ by the
-definition of subtyping. Let $V' = T'$.
-By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.
-It follows by rule \tfunc, and \tsub to replace $U'$ with $U$, that
-$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-
-\item $x \in \overline{z_i}$, meaning that $T = T_i$ and 
-$V = T_i' = \remapapp{T}{U_i}{V_i}$ for some $i$.
-Let $V' = \remapapp{T'}{U_i}{V_i}$. 
-By \lemref{lem:remapcv}, $V' <: V$.
-As $T' <: T$, we can type $t$ by rule \tfunc such that
-$\typerule{\Gamma, x : T'}{t}{U}{\Gamma, x : V'}$.
-
-\end{itemize}
-
-\item $t = y.m$. By the rule \tmethc it follows that
-$\Gamma, x : T = \Gamma_1, y : O@\overline{S}$
-and that
-$\Gamma', x : V = \Gamma_1, y : O@\overline{S'}$
-for some $\Gamma_1, O, \overline{S}$ and $\overline{S'}$
-where
-$\overline{S'} = \{ S_n | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \}$.
-Additionally, 
-$U = \bigsqcup{ \{ T_r | m : T_r \Rightarrow S_n \in O@S, S \in \overline{S} \} }$.
-There are two cases to consider:
-
-\begin{itemize}
-\item $x = y$. It follows that $\Gamma = \Gamma_1 = \Gamma'$ and
-that $T = O@\overline{S}, V = O@\overline{S'}$.
-$T'$ must be of form $O'@\overline{S}''$ such that
-$O'@\overline{S}'' \precsim O@\overline{S}$ by rule \subobj.
-
-Let $V' = O'@\overline{S}''' = \remapapp{T'}{T}{V}$
-where
-$S''' = \{ S_n' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}''\}$.
-Let $U' = \bigsqcup{ \{ T_r' | m : T_r' \Rightarrow S_n' \in O'@S'', S'' \in \overline{S}'' \} }$.
-As $O'@\overline{S}'' \precsim O@\overline{S}$, it follows that
-$O'@\overline{S}''' \precsim O@\overline{S}'$.
-Therefore, $U' <: U$ and $V' <: V$. By the rule \tmethc, and \tsub
-to replace $U'$ with $U$, it follows that
-$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-
-\item $x \neq y$. It follows that $\Gamma = \Gamma'', y : O@\overline{S}$
-and that $\Gamma' = \Gamma'', y : O@\overline{S'}$
-for some $\Gamma''$, $O$ and $\overline{S}$, and that
-$T = V$.
-By rule \tmethc we have that $\typerule{\Gamma}{t}{U}{\Gamma'}$.
-Let $V' = T'$. By \lemref{lem:remapeff}, $V' <: \remapapp{T'}{T}{V}$.
-By the weakening lemma, 
-$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-
-\end{itemize}
-
-\item $t = t' ; t''$. It follows by rule \tlet that there exists
-$\Gamma'', V_{mid}, U'$ 
-such that $\typerule{\Gamma, x : T}{t'}{U'}{\Gamma'', x : V_{mid}}$
-and that $\typerule{\Gamma'', x : V_{mid}}{t''}{U}{\Gamma', x : V}$.
-By induction, it follows that there exists $V_{mid}' <: \remapapp{T'}{T}{V_{mid}}$ 
-such that $\typerule{\Gamma, x : T'}{t'}{U'}{\Gamma'', x : V_{mid}'}$.
-By \lemref{lem:remap} we have that $V_{mid}' <: V_{mid}$.
-There it also follows by induction that 
-there exists $V_{end} <: \remapapp{V_{mid}'}{V_{mid}}{V}$ such that
-$\typerule{\Gamma'', x : V_{mid}'}{t''}{U}{\Gamma'', x : V_{end}}$.
-Let $V' = V_{end}$. By rule \tlet it follows that
-$\typerule{\Gamma, x : T'}{t' ; t''}{U}{\Gamma'', x : V'}$.
-
-\item $t = \ift{t_c}{t_t}{t_f}$. It follows from rule \tif that there
-exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $U_t$, $U_f$, $V_c$, $V_t$ and
-$V_f$
-such that
-$\typerule{\Gamma, x : T}{t_c}{\boolt}{\Gamma_1, x : V_c}$ and that
-$\typerule{\Gamma_1, x : V_c}{t_t}{U_t}{\Gamma_2, x : V_t}$ and that
-$\typerule{\Gamma_1, x : V_c}{t_f}{U_f}{\Gamma_3, x : V_f}$ with
-$\Gamma' = \Gamma_2 \sqcup \Gamma_3$, $U = U_t \sqcup U_f$ and
-$V = V_t \sqcup V_f$.
-
-By induction, there exists a $V_c' <: \remapapp{T'}{T}{V_c}$
-such that $\typerule{\Gamma, x : T'}{t_c}{\boolt}{\Gamma_1, x : V_c'}$.
-By \lemref{lem:remap} we also have that $V_c' <: V_c$.
-Therefore by induction there also exists a 
-$V_t' <: \remapapp{V_c'}{V_c}{V_t}$ and $V_f' <: \remapapp{V_c'}{V_c}{V_f}$
-such that $\typerule{\Gamma_1, x : V_c'}{t_t}{T_t}{\Gamma_2, x : V_t'}$
-and that $\typerule{\Gamma_1, x : V_c'}{t_f}{T_f}{\Gamma_3, x : V_f'}$.
-By \lemref{lem:remap} we also have that $V_t' <: V_t$ and that
-$V_f' <: V_f$.
-Let $V' = V_t' \sqcup V_f'$. By the definition of join, $V' <: V$.
-It follows by rule \tif that
-$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
-
-\end{itemize}
+	\end{itemize}
+
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+%%% Last revised and checked: 2012/12/13
 
@@
 \begin{proof}
-by induction on the structure of $t$.
-
-If $t$ is some value $v$, it follows by \lemref{lem:valuectx} that 
-$\Gamma = \Gamma'$ and that the typing derivation will still be valid as long
-as the input and output contexts are identical. Therefore,
-$\typerule{\Gamma , \Gamma''}{v}{T}{\Gamma', \Gamma''}$.
-
-If $t$ is not a value, then it is of one of the following forms:
-
-\begin{itemize}
-\item $t = \lett{x}{t'}{t''}$. It follows from rule \tlet that
+by induction on the typing derivation of $t$.
+
+\begin{itemize}
+\item $t$ is typed by one of \tunit, \ttrue, \tfalse, \tobj or \tfundef.
+It follows that $\Gamma = \Gamma'$ and the term can be typed with any
+context $\Gamma'''$ such that $\typerule{\Gamma'''}{t}{T}{\Gamma'''}$
+using the same rule. 
+Therefore, $\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \tlet. It follows that $t = \lett{x}{t'}{t''}$ and
 there exists $\Gamma_1, \Gamma_2, T', T''$ such that
 $\typerule{\Gamma}{t'}{T'}{\Gamma_1}$ and
-$\typerule{\Gamma_1, x : T'}{t''}{T}{\Gamma', x : T''}$.
-By \tlet we have that $dom(\Gamma') = dom(\Gamma)$.
+$\typerule{\Gamma_1, x : T'}{t''}{T}{\Gamma', x : T''}$,
+with $dom(\Gamma') = dom(\Gamma)$.
+
 By induction, $t'$ can be typed such
@@
 and $t''$ can be typed such that
-$\typerule{\Gamma_1, \Gamma'', x : T'}{t''}{T}{\Gamma', \Gamma'', x : T''}$,
-as we can assume by Barendregt's convention that $x$ can be made distinct from
+$\typerule{\Gamma_1, \Gamma'', x : T'}{t''}{T}{\Gamma', \Gamma'', x : T''}$.
+We can assume by Barendregt's convention that $x$ can be made distinct from
 any variable names in $\Gamma''$ by relabeling $x$.
-Therefore,
-$\typerule{\Gamma, \Gamma''}{\lett{x}{t'}{t''}}{T}{\Gamma', \Gamma''}$.
-
-\item $t = x ( \overline{x_i} )$. It follows from the rule \tfunc that
+Therefore by rule \tlet
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \tfunc. It follows that
 there exists a $\Gamma_1, \overline{T_i}, \overline{U_i}, \overline{V_i}, \overline{T_i'}$ 
@@
 can extend the input and output contexts such that
-$\typerule{\Gamma, \Gamma''}{x ( \overline{x_i} )}{T}{\Gamma', \Gamma''}$.
-
-\item $t = x.m$. It follows from the rule \tmethc that there exists
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \tmethc. It follows that there exists
 a $\Gamma_1$ such that $\Gamma = \Gamma_1, x : O@\overline{S}$ and that 
@@
 
-\item $t = t' ; t''$. It follows from the rule \tseq that there exists
+\item $t$ is typed by \tseq. Therefore $t = t' ; t''$ and there exists
 $\Gamma_1$ and $T'$ such that $\typerule{\Gamma}{t'}{T'}{\Gamma_1}$
-and $\typerule{\Gamma_1}{t''}{T}{\Gamma'}$. By induction
+and $\typerule{\Gamma_1}{t''}{T}{\Gamma'}$. By \lemref{lem:ctxsubset}
+it follows that $dom(\Gamma') \subseteq dom(\Gamma_1) \subseteq dom(\Gamma)$.
+
+By induction
 $\typerule{\Gamma, \Gamma''}{t'}{T'}{\Gamma_1, \Gamma''}$
-as $dom(\Gamma_1) \subseteq dom(\Gamma)$ and therefore
-$\Gamma_1$ and $\Gamma''$ are disjoint.
-Similarly, as $dom(\Gamma_1) = dom(\Gamma')$, by induction
-$\typerule{\Gamma_1, \Gamma''}{t''}{T}{\Gamma', \Gamma''}$.
-Therefore $\typerule{\Gamma, \Gamma''}{t'}{T'}{\Gamma_1, \Gamma''}$.
-
-\item $t = \ift{t_c}{t_t}{t_f}$. It follows from rule \tif that there
-exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$ and $T_f$ such that
-$\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$ and that
-$\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$ and that
-$\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ with
-$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$.
-By induction each of $\Gamma_1$, $\Gamma_2$ and $\Gamma_3$ can
-be extended by $\Gamma''$ in the typing judgements for $t_c$,
-$t_t$ and $t_f$. As $\Gamma'' \sqcup \Gamma'' = \Gamma''$, it
-follows that $(\Gamma_2, \Gamma'') \sqcup (\Gamma_3, \Gamma'') = 
-\Gamma_2 \sqcup \Gamma_3, \Gamma''$ and therefore $t$ can be
-typed such that $\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
-
-\end{itemize}
-
+and 
+$\typerule{\Gamma_1, \Gamma''}{t''}{T}{\Gamma', \Gamma''}$,
+as $dom(\Gamma_1) \cap dom(\Gamma'') = \emptyset$.
+
+Therefore by \tseq, 
+$\typerule{\Gamma, \Gamma''}{t'}{T'}{\Gamma_1, \Gamma''}$.
+
+\item $t$ is typed by \tif. Therefore $t = \ift{t_c}{t_t}{t_f}$. and there
+exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$ and $T_f$ such that:
+
+\begin{itemize}
+\item $\typerule{\Gamma}{t_c}{\boolt}{\Gamma_1}$
+\item $\typerule{\Gamma_1}{t_t}{T_t}{\Gamma_2}$
+\item $\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$
+\end{itemize}
+
+Additionally, $\Gamma' = \Gamma_2 \sqcup \Gamma_3$.
+Context join is only defined where $dom(\Gamma_2) = dom(\Gamma_3)$,
+and by definition $dom(\Gamma') = dom(\Gamma_2)$.
+
+By induction, $\typerule{\Gamma, \Gamma''}{t_c}{\boolt}{\Gamma_1, \Gamma''}$.
+
+By induction, $\typerule{\Gamma_1, \Gamma''}{t_t}{T_t}{\Gamma_2, \Gamma''}$
+as by \lemref{lem:ctxsubset} we have that $dom(\Gamma_1) \subseteq dom(\Gamma)$,
+therefore $dom(\Gamma_1) \cap dom(\Gamma'') = \emptyset$.
+
+Similarly by induction, 
+$\typerule{\Gamma_1, \Gamma''}{t_f}{T_f}{\Gamma_2, \Gamma''}$.
+
+Therefore by \tif, $\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \tdrop, therefore $\typerule{\Gamma}{t}{T}{\Gamma', \Gamma'''}$
+for some $\Gamma'''$. By induction it follows that
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma'', \Gamma'''}$.
+Therefore by rule \tdrop,
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \tsub. It follows that $\typerule{\Gamma}{t}{T'}{\Gamma'}$
+for $T' <: T$. By induction, $\typerule{\Gamma, \Gamma''}{t}{T'}{\Gamma', \Gamma''}$.
+Therefore by rule \tsub, $\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+
+\item $t$ is typed by \twiden. It follows that $\typerule{\Gamma}{t}{T}{\Gamma_{out}, x : U'}$
+where $\Gamma' = \Gamma_{out}, x : U$ where $U' <: U$. 
+By \lemref{lem:ctxsubset} it must be the case that $x \in dom(\Gamma)$, there
+$x \notin dom(\Gamma'')$.
+By induction,
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma_{out}, \Gamma'', x : U'}$.
+By \twiden, 
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma_{out}, \Gamma'', x : U}$.
+
+\item $t$ is typed by \tnarrow. It follows that
+$\typerule{\Gamma_{in}, x : U}{t}{T}{\Gamma'}$
+where $\Gamma = \Gamma_{in}, x : U'$ where $U' <: U$.
+By induction,
+$\typerule{\Gamma_{in}, \Gamma'', x : U'}{t}{T}{\Gamma', \Gamma''}$.
+By rule \tnarrow,
+$\typerule{\Gamma, \Gamma''}{t}{T}{\Gamma', \Gamma''}$.
+\end{itemize}
 \end{proof}
@@
 
+\section{Existence of principal types}
+\label{app:principal}
+
+\subsection{Concatenation of effect types}
+$T \gg T' \cdot U \gg U'$ is the concatenation operation for effect types. 
+Effect concatenation is a partial binary operation that produces an 
+effect type $V \gg V'$ such that:
+
+\begin{itemize}
+\item $V <: T$
+\item $V_{mid} <: \remapapp{V}{T}{T'}$
+\item $V_{mid} <: U$
+\item $V' <: \remapapp{V_{mid}}{U}{U'}$
+\end{itemize}
+
+The intuition behind effect concatenation is to find a variable input and
+output type for $x$ in the term {\tt f(x) ; g(x) } given 
+$f : (T \gg T') \rightarrow \top$ and
+$g : (U \gg U') \rightarrow \top$. The typing
+$\tau = (\Gamma, \top, \Gamma')$ of term $t$ where
+$\Gamma = \Gamma_f, x : V$,
+$\Gamma' = \Gamma_f, x : V''$ and
+$\Gamma_f = f : (T \gg T') \rightarrow, g : (U \gg U') \rightarrow \top$.
+If $\tau$ exists, then $\tau$ is the principal typing for $t$, given
+the assumed types of $f$ and $g$.
+
+$T \gg T' \cdot U \gg U'$ is defined such that:
+
+\begin{itemize}
+\item The concatenation of identical effects where the input and
+output types are equal is idempotent: $T \gg T \cdot T \gg T = T \gg T$.
+\item Where function types are concerned, the meet of the function types
+is used. Let $T = \funt{T_i}{T_i'}{T'}$, and
+$U = \funt{U_i}{U_i'}{U'}$. If $V = T \sqcap U$ is defined, then
+$T \gg T \cdot U \gg U = V \gg V$. Otherwise
+$T \gg T \cdot U \gg U$ is undefined.
+\item The concatenation of effects concerning object types is the most
+interesting case.
+Consider the effect types $\ot{O_1}{S_1} \gg \ot{O_1}{S_2}$ and
+$\ot{O_2}{S_3} \gg \ot{O_2}{S_4}$. Each effect type defines a
+deterministic finite state machine: $O_1 = (\Sigma_1, M_1, \Delta_1)$
+defines the state labels and transitions, while $\overline{S_1}$ 
+defines the input state set and $\overline{S_2}$ defines the output
+state set. Therefore, the well-understood process of
+deterministic finite state machine concatentation can be used to produce
+a new finite state machine $(O_3, \overline{S_5}, \overline{S_6})$ from
+the two machines $\mathcal{M}_1 = (O_1, \overline{S_1}, \overline{S_2})$ and
+$\mathcal{M}_2 = (O_2, \overline{S_3}, \overline{S_4})$. The effect type
+$\ot{O_3}{S_5} \gg \ot{O_3}{S_6}$ can then be extracted from this.
+
+The function defining the concatenation operation for state machines is
+not always defined, due to the return types of methods.
+If $\mathcal{M}_1$ contains a transition out of an end state $m : T$
+while the second contains a transition out of the initial state 
+$m : U$, then the concatenation is only defined if $T \sqcap U$ is defined.
+This would then be the return type of the method $m$ in the concatenated
+machine, where the two methods would overlap.
+\end{itemize}
+
+\subsection{Concatenation of contexts}
+
+Let $t$ and $t'$ be arbitrary typeable terms. If $(\Gamma_1, T, \Gamma_1')$
+is a typing of $t$ and $(\Gamma_2, U, \Gamma_2')$ is a typing of $t'$, we
+desire a means of deriving a typing for the term {\tt t ; t'} if one exists.
+Using effect concatentation, the partial function
+$\mathit{seqctx}(\Gamma_1, \Gamma_1', \Gamma_2, \Gamma_2')$ can be defined
+which fufills this function, with some restrictions.
+
+Let $\Gamma_1, \Gamma_1', \Gamma_2, \Gamma_2'$ be contexts such
+that $dom(\Gamma_1) = dom(\Gamma_1')$ and $dom(\Gamma_2) = dom(\Gamma_2')$.
+
+Let $\overline{x} = dom(\Gamma_1) - dom(\Gamma_2)$.
+Let $\Gamma_x = \overline{x : \Gamma_1(x)}$.
+and $\Gamma_x' = \overline{x : \Gamma_1'(x)}$.
+
+Let $\overline{y} = dom(\Gamma_2) - dom(\Gamma_1)$.
+Let $\Gamma_y = \overline{y : \Gamma_2(y)}$.
+and $\Gamma_y' = \overline{y : \Gamma_2'(y)}$.
+
+Let $\overline{z} = dom(\Gamma_1) \cap dom(\Gamma_2)$. For each
+$z_i \in \overline{z}$, let $T_i \gg V_i = 
+\Gamma_1(z_i) \gg \Gamma_1'(z_i) \cdot \Gamma_2(z_i) \gg \Gamma_2'(z_i)$,
+which may be undefined.
+
+Where it is defined, let $U_i = \remapapp{T_i}{\Gamma_1(z_i)}{\Gamma_1'(z_i)}$.
+By the definition of effect concatenation, we have that:
+
+\begin{itemize}
+\item $T_i <: \Gamma_1(z_i)$
+\item $U_i <: \Gamma_2(z_i)$
+\item $V_i <: remap(U_i, \Gamma_b(z_i) \gg \Gamma_b'(z_i))$
+\end{itemize}
+
+We define $\mathit{seqctx}$ such that
+$\mathit{seqctx}(\Gamma_1, \Gamma_1', \Gamma_2, \Gamma_2') = (\Gamma, \Gamma_{mid}, \Gamma')$,
+where:
+
+\begin{itemize}
+\item $\Gamma = \Gamma_x, \Gamma_y, \overline{z_i : T_i}$.
+\item $\Gamma_{mid} = \Gamma_x', \Gamma_y, \overline{z_i : U_i}$.
+\item $\Gamma' = \Gamma_x', \Gamma_y', \overline{z_i : V_i}$.
+\end{itemize}
+
+It follows that $\Gamma$, $\Gamma_{mid}$ and $\Gamma'$ have the following
+properties:
+
+\begin{itemize}
+\item $dom(\Gamma) = dom(\Gamma') = dom(\Gamma_1) \cup dom(\Gamma_2)$
+\item $\forall x \in \Gamma_1 . \Gamma(x) <: \Gamma_1(x)$
+\item $\forall x \in \Gamma_1' . \Gamma_{mid}(x) <: \Gamma_1'(x)$
+\item $\forall x \in \Gamma_2 . \Gamma_{mid}(x) <: \Gamma_2(x)$
+\item $\forall x \in \Gamma_2' . \Gamma'(x) <: \Gamma_2'(x)$
+\end{itemize}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:principal_nofun %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%% BROKEN, needs remap fixes
+
+\subsection{Proofs \& Supporting Lemmas}
+
+\begin{lem}
+\label{lem:principal_nofun}
+Principal typings exist for all terms $t$ which do not contain function calls.
+\end{lem}
+
+\begin{proof}
+by induction on the structure of $t$.
+
+\begin{itemize}
+\item $t = \unitv$. 
+Trivially, the typing $(\emptyset, \unitt, \emptyset)$ is principal.
+
+\item $t = \truev$ or $t = \falsev$. 
+Trivially, the typing $(\emptyset, \boolt, \emptyset)$ is principal.
+
+\item $t = \funv{x_i}{T_i}{U_i}{t'}$. 
+As $t$ is typeable, it must be the case that $t'$ is typeable also.
+By induction, the term $t'$ has a principal typing
+$(\Gamma, T, \Gamma')$. It follows that for each $i$, if $x_i \notin dom(\Gamma)$
+then $x_i$ is not used in $t'$. Let $\overline{x_j} \subseteq \overline{x_i}$
+be the set of such unused variables.
+Let $\Gamma_2 = \Gamma, \overline{x_j : \top}$ and 
+$\Gamma_2' = \Gamma', \overline{x_j : \top}$.
+By \lemref{lem:weakening}, $(\Gamma_2, T, \Gamma_2')$ is a typing of $t'$.
+
+If the effect $T_i \gg U_i$ is specified for parameter $x_i$ then it
+must be the case that $T_i <: \Gamma_2(x_i)$ for each $i$ so that 
+$t$ is typeable by rule \tfunc.
+Let $\Gamma_3 = \overline{x_i : T_i'}$ such that:
+
+\begin{itemize}
+\item $T_i' = T_i$ if $T_i$ is specified
+\item $T_i' = \Gamma_2(x_i)$ otherwise.
+\end{itemize}
+
+By repeated application of \lemref{lem:upgrade}, there exists a $\Gamma_3'$ and 
+$T'$ such that
+$(\Gamma_3,T',\Gamma_3')$ is a typing of $t'$, where $T' <: T$,
+$dom(\Gamma_3') = dom(\Gamma_2')$ and 
+$\forall x \in \Gamma_3' . \Gamma_3'(x) <: \Gamma_2'(x)$.
+
+Let $U_i' = U_i$ if $U_i$ is specified or $U_i' = \Gamma_3'(x)$ otherwise.
+It must be the case that $\Gamma_3'(x) <: U_i'$ for each $i$ so that
+$t$ is typeable by rule \tfunc.
+Let $\Gamma_4 = \overline{x_i : U_i}$.
+By repeated application of rule \twiden, 
+$(\Gamma_3,T',\Gamma_4)$ is a typing of $t'$.
+
+By rule \tfundef it follows that
+$\tau = (\emptyset, \funt{T_i'}{U_i'}{T'}, \emptyset)$ is a typing of $t$. 
+There can be no stronger typing than this, as $T_i'$ has an upper bound
+determined by either the principal typing of $t'$ or the user specified
+type $T_i$. Similarly, $U_i'$ has a lower bound determined by
+modification of the principal typing of $t'$ or the user specified
+$U_i$. Therefore $\tau$ is the principal type of $t$.
+
+\item $t = \left[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij},S_{ij}') } \} } \right]@S$.
+As $t$ is typeable, each $v_{ij}$ must be typeable also.
+By induction, each $v_{ij}$ has a principal typing $(\emptyset, T_{ij}, \emptyset)$,
+as values can be typed with arbitrary input contexts, and principal typings
+have the smallest input context domain possible.
+
+Trivially, $\tau = (\emptyset,\{ \overline{ S_i \{ \overline{ m_{ij} = T_{ij} \Rightarrow S_ij } \} } \}@S,\emptyset)$ is
+the principal typing of $t$.
+
+\item $t = \lett{x}{t_v}{t_b}$.
+
+As $t$ is typeable, $t_v$ and $t_b$ must be typeable also.
+By induction, $t_v$ has a principal typing $(\Gamma_v,T_v,\Gamma_v')$, and
+$t_b$ has a principal typing $(\Gamma_b,T_b,\Gamma_b')$.
+As $t$ is typeable, by rule \tlet it must be the case that 
+$x \notin dom(\Gamma_v)$.
+
+Let $\Gamma_b'' = \Gamma_b - \{x\}$ and
+$\Gamma_b''' = \Gamma_b' - \{x\}$.
+
+Let $(\Gamma_{in}, \Gamma_{mid}, \Gamma_{out}) = seqctx(\Gamma_v, \Gamma_v', \Gamma_b'', \Gamma_b''')$.
+If this is undefined, then there exists a $y \in dom(\Gamma_v) \cap dom(\Gamma_b'')$
+that cannot be assigned a type that supports the usage of $y$ in $t_v$
+followed by $t_b$. In this case, $t$ itself would be untypeable. 
+Therefore, $seqctx(\Gamma_v, \Gamma_v', \Gamma_b'', \Gamma_b''')$
+is defined.
+
+By \lemref{lem:upgrade} and the definition of $\mathit{seqctx}$, 
+there exists a $T_v' <: T_v$ such that
+$(\Gamma_{in}, T_v', \Gamma_{mid})$ is a typing of $t_v$.
+
+If $x \in \Gamma_b$, let $T_x = \Gamma_b(x)$ and $T_x' = \Gamma_b'(x)$.
+If $x \notin \Gamma_b$ (i.e. $x$ is not used in the body of the let binding), 
+let $T_x = T_x' = \top$. Either directly or by \lemref{lem:weakening},
+$((\Gamma_b'', x : T_x), T_b, (\Gamma_b''', x : T_x'))$ is a typing of $t'$.
+
+In order for $t$ to be typeable, it must be the case that $T_v' <: T_x$.
+Let $\Gamma_{mid}' = \Gamma_{mid}, x : T_v'$.
+By \lemref{lem:upgrade}, there exists a $T_b' <: T_b$ and $T_x'' <: remap(T_v', T_x \gg T_x')$ such that
+$(\Gamma_{mid}', T_b', (\Gamma_{out}, x : T_x''))$ is a typing of $t_b$.
+
+By rule \tlet it follows that $(\Gamma_{in}, T_b', \Gamma_{out})$ is a
+typing of $t$. This typing must be principal.
+
+%%% TODO: proof of principality from construction?
+
+\item $t = x.m$.
+
+Let $O = \{ S \{ m : \top \Rightarrow S' \} S' \{ \} \}$.
+
+Trivially, $\tau = (x : O@S, \top,  x : O@S')$
+is a principal typing of $t$.
+
+\item $t = t_a; t_b$.
+
+As $t$ is typeable, it must be the case that $t_a$ and $t_b$ are typeable.
+By induction, $t_a$ has a principal typing $(\Gamma_a, T_a, \Gamma_a')$ and
+$t_b$ has a principal typing $(\Gamma_b, T_b, \Gamma_b')$.
+
+Let $(\Gamma, \Gamma_{mid}, \Gamma') = \mathit{seqctx}(\Gamma_a, \Gamma_a', \Gamma_b, \Gamma_b')$.
+If this is not defined, then there exists a $y \in dom(\Gamma_a) \cap dom(\Gamma_b)$
+that cannot be assigned a type that supports the usage of $y$ in $t_a$
+followed by the usage in $t_b$. 
+In this case, $t$ itself would be untypeable. Therefore, it must be defined.
+
+By the definition of $\mathit{seqctx}$, we have that for each
+$x \in dom(\Gamma_a)$ that $\Gamma(x) <: \Gamma_a(x)$, and
+that $\Gamma_{mid}(x) <: \remapapp{\Gamma(x)}{\Gamma_a(x)}{\Gamma_a'{x}}$.
+
+%%% TODO: Prove that
+It follows that $(\Gamma, T_a', \Gamma_{mid})$ is a typing of
+$t_a$.
+
+%%% TODO: prove that
+It follows that $(\Gamma_{mid}, T_b', \Gamma_{mid})$ is a typing of
+$t_b$.
+
+It follows by rule \tseq that $(\Gamma_{in}, T_b', \Gamma_{out})$ is a typing
+of $t$.
+
+%%% TODO: show this is minimal.
+
+\item $t = \ift{t_c}{t_t}{t_f}$.
+
+As $t$ is typeable, it follows that $t_c$, $t_t$ and $t_f$ must be typeable.
+By induction, $t_c$ has a principal typing $\tau_c = (\Gamma_c, T_c, \Gamma_c')$,
+$t_t$ has a principal typing $\tau_t = (\Gamma_t, T_t, \Gamma_t')$ and
+$t_f$ has a principal typing $\tau_f = (\Gamma_f, T_f, \Gamma_f')$.
+
+In order for $t$ to be typeable, it must be the case that a typing
+$(\Gamma_c'', \boolt, \Gamma_c''')$ exists for $t_c$. It follows that as
+$\tau_c$ is a principal typing for $t_c$, that for each $x \in \Gamma_a$ that 
+$\Gamma_a''(x) <: \Gamma_a(x)$ and that $\Gamma_a'(x) <: \Gamma_a'''(x)$,
+with $\boolt <: T_c$.
+
+%%% seqctx for T_c -> T_t branch
+
+%%% seqctx for T_c -> T_f branch
+
+%%% \sqcap of all types in respective \Gamma_{in} of each seqctx to
+%%% derive overall typing.
+
+%%% show this is principal.
+
+\end{itemize}
+\end{proof}
+
+\begin{thm}
+\label{thm:principal}
+Principal typings exist for all terms $t$ which
+are typeable when a type can be assumed for all variables used as functions.
+\end{thm}
+
+\begin{proof}
+by induction on the structure of $t$.
+
+\end{proof}
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 \end{document}
+
+%--------------------------------------
+% abandoned stuff
+%--------------------------------------
+
+% \begin{thm}
+% A closed term $t$ which is typeable always has a principal typing.
+% \end{thm}
+
+% \begin{proof}
+% Let $t$ be an arbitrary closed term. $t$ must therefore be contain a
+% $\lett{x_i}{t_i}{t_i'}$ sub-term for every variable that is used within
+% it. We may therefore assume the existence of a type for
+% any variable used by a given subterm $t'$.
+
+% We shall prove by induction on the structure of such closed terms $t'$.
+% For each $t'$, we will assume the existence of $\Gamma$, an input context
+% which contains all necessary
+
+% \begin{itemize}
+% \item $t = \unitv$. The typing $(\emptyset, \unitt, \emptyset)$ is principal.
+
+% \item $t = \truev$ or $t = \falsev$. The typing $(\emptyset, \boolt, \emptyset)$ is principal.
+
+% \item $t = \funv{x_i}{T_i}{U_i}{t'}$. By induction, the term $t'$ has a principal typing
+% $(\Gamma, T, \Gamma')$. It follows that for each $i$, if $x_i \notin dom(\Gamma)$
+% then $x_i$ is not used in $t'$. Let $\overline{x_j} \subseteq \overline{x_i}$
+% be the set of such unused variables.
+% Let $\Gamma_2 = \Gamma, \overline{x_j : \top}$ and 
+% $\Gamma_2' = \Gamma', \overline{x_j : \top}$. It is therefore the case
+% that $dom(\Gamma_2) = dom(\Gamma_2') = \overline{x_i}$.
+% By \lemref{lem:weakening}, $(\Gamma_2, T, \Gamma')$ is a typing of $t'$.
+% As $t$ is typeable, it follows that for each $i$ that if the effect type
+% $T_i \gg U_i$ is declared, that $\Gamma_2(x) <: T_i$ and that 
+% $U_i <: \Gamma_2'(x)$. 
+% Let $T_i' = T_i$ if $T_i$ is declared, or $T_i' = \Gamma_2(x_i)$ otherwise.
+% Let $U_i' = _Ui$ if $U_i$ is declared, or $U_i' = \Gamma_2'(x_i)$ otherwise.
+
+% By rule $\tfundef$ it follows that
+% $\tau = (\emptyset, \funt{T_i'}{U_i'}{T}, \emptyset)$ is a typing of $t$. Further,
+% there can be no stronger typing than this, as for each $i$ the type 
+% $U_i'$ is either bounded by the user defined $U_i$, the principal variable 
+% output type of $x_i$ in $t'$, or the assigned type $\top$ which is the greatest 
+% element of the type lattice. Therefore, $\tau$ is the principal type of $t$.
+
+% \item $t = \left[ \overline{S_i \{ \overline{ m_{ij} = (v_{ij},S_{ij}') } \} }]@S$.
+
+% \item $t = \lett{x}{t_v}{t_b}$.
+
+% \item $t = \lett{x}{t_v}{x.m}$.
+
+% \item $t = x(\overline{x_i})$.
+
+% \item $t = t'; t''$.
+
+% \item $t = \ift{t_c}{t_t}{t_f}$.
+
+% \end{itemize}
+% \end{proof}

### 2013-01-04 formalism/paper.tex
fixed compile error, minor layout problems in progress/preservation proof

@@
 \label{thm:soundness}
-Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
-with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
-there exists a $t'$ and $\mu'$ such that
-$\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
-$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
+Given a non-value term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
+and a store $\mu$ such that $\Gamma \vdash \mu$, then:
+
+\begin{itemize}
+\item there exists a $t'$ and $\mu'$ such that
+$\oprule{t}{\mu}{t'}{\mu'}$
+\item there exists a $\Gamma''$ such that
+$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$
+\end{itemize}
 \end{thm}
@@
 \begin{proof}
-by induction on the typing derivation of $t$. Assume $t$ is not a value. It has
-therefore been typed by one of the following rules:
+by induction on the typing derivation of $t$. As $t$ is not a value,
+$\typerule{\Gamma}{t}{T}{\Gamma'}$ must have been derived by one of the
+following rules:
 
@@
   Reduction can occur by \rletv such that 
-  $\oprule{\lett{x}{t_v}{t_b}}{\mu}{t_b}{\mu [ x \mapsto t_v ]}$.
+  $\oprule{t}{\mu}{t_b}{\mu [ x \mapsto t_v ]}$.
   Let $\mu' = \mu [ x \mapsto t_v ]$. Let $\Gamma'' = \Gamma_1, x : T_v$.
@@
   \begin{itemize}
-  \item $t_l$ is a value. Reduction can therefore occur by
+  \item $t_l$ is a value. Reduction can therefore occur by \\
   \rseqlv such that
@@
 	$\typerule{\Gamma, x : T'}{t}{U}{\Gamma', x : V'}$.
+
+	\end{itemize}
 

### 2013-01-06 formalism/paper.tex
removed unnecessary / unwritten proofs for remap, all typing proofs reviewed and fixed. principal typing proof still broken.

@@
 \newcommand{\os}[2]{\omega_{#1}@\sigma_{#2}}
+
+\newcommand{\pathrefl}{$\lesssim${\sc-refl}}
+\newcommand{\pathadj}{$\lesssim${\sc-adj}}
+\newcommand{\pathtrans}{$\lesssim${\sc-trans}}
+\newcommand{\pathbranch}{$\lesssim${\sc-branch}}
 
@@
 \]
-\caption{\label{fig:simreach} Definition of the simulation predecessor relation}
+\caption{\label{fig:simreach} Definition of the path relation}
 \end{figure}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% last revised and checked: 2013/01/04
 
@@
 Given an arbitrary term $t$ such that 
-$\typerule{\Gamma, x : T}{t}{T}{\Gamma', x : V}$ where $x$ is unused in
+$\typerule{\Gamma, x : T}{t}{U}{\Gamma', x : V}$ where $x$ is unused in
 $t$, $T <: V$.
 \end{lem}
+
+\begin{proof}
+Straightforward observation from the typing rules.
+\end{proof}
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% revised and checked 2013/01/04
 
@@
 As $\boolt$ and $\unitt$ have no subtypes other than
-themselves, it follows that $T' = T$. Trivially,
-$\remapapp{T'}{U}{V} = \remapapp{T}{U}{V}$.
+themselves, it follows that $T' = T$. By \subrefl,
+$\remapapp{T'}{U}{V} <: \remapapp{T}{U}{V}$.
 
 \item $U = \funt{U_i}{U_i'}{U'}$. By \remapfunc it follows that
-$\remapapp{T'}{U}{V} = T' <: T = \remapapp{T}{U}{V}$.
-
-\item $U = \ot{O_2}{S_2}$, with $T = \ot{O_1}{S_1}$ and $V = \ot{O_2}{S_4}$
-and $\ot{O_1}{S_3} = \remapapp{T}{U}{V}$. Let $T' = \ot{O_3}{S_5}$,
-and $\ot{O_3}{S_6} = \remapapp{T'}{U}{V}$. By definition,
-$\overline{S_3} = \bigcup \{ \overline{S} \mid (\ot{O_1}{S_1}, \ot{O_2}{S_2}) \lesssim (\ot{O_1}{S}, \ot{O_2}{S_4}) \}$, and
-$\overline{S_6} = \bigcup \{ \overline{S'} \mid (\ot{O_3}{S_5}, \ot{O_2}{S_2}) \lesssim (\ot{O_3}{S}, \ot{O_2}{S_4}) \}$.
+$\remapapp{T'}{U}{V} = T'$ and $\remapapp{T}{U}{V} = T$, therefore
+$\remapapp{T'}{U}{V} <: \remapapp{T}{U}{V}$.
+
+\item $U = \ot{O_1}{S_1}$. By \remapobj it follows that 
+$V = \ot{O_1}{S_2}$ and
+$T = \ot{O_2}{S_3}$ with
+$\remapapp{T}{U}{V} = \ot{O_2}{S_4}$. Similarly,
+$T' = \ot{O_3}{S_5}$
+with $\remapapp{T'}{U}{V} = \ot{O_3}{S_6}$.
+
+By definition,
+$\overline{S_4} = \bigcup \{ \overline{S} \mid (\ot{O_2}{S_3}, \ot{O_1}{S_1}) \lesssim (\ot{O_2}{S}, \ot{O_1}{S_2}) \}$, and
+$\overline{S_6} = \bigcup \{ \overline{S'} \mid (\ot{O_3}{S_5}, \ot{O_1}{S_1}) \lesssim (\ot{O_3}{S'}, \ot{O_1}{S_2}) \}$.
 In both cases, each $\overline{S}$ and $\overline{S'}$ extracted from
 the path relation must be such that 
-$(\ot{O_3}{S_5}, \ot{O_1}{S_1}) \lesssim (\ot{O_3}{S'}, \ot{O_1}{S})$.
-Therefore, $\ot{O_3}{S_6} <: \ot{O_1}{S_3}$.
+$(\ot{O_3}{S_5}, \ot{O_2}{S_3}) \lesssim (\ot{O_3}{S'}, \ot{O_2}{S})$
+as $T' <: T$.
+Therefore, $\ot{O_3}{S_6} <: \ot{O_2}{S_4}$.
 
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_in_contra %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{lem}
-\label{lem:remap_in_contra}
-Given types $T, U, U'$ and $V$ such that $U' <: U$ and that
-both $\remapapp{T}{U'}{V}$ and $\remapapp{T}{U}{V}$ are defined,
-$\remapapp{T}{U}{V} <: \remapapp{T}{U'}{V}$.
-\end{lem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remap_out_cv %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\begin{lem}
-\label{lem:remap_out_cv}
-Given types $T, U, V$ and $V'$ such that $V' <: V$ and that
-both $\remapapp{T}{U}{V}$ and $\remapapp{T}{U}{V'}$ are defined,
-$\remapapp{T}{U}{V'} <: \remapapp{T}{U}{V}$.
-\end{lem}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:lesssim_same %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% revised and checked: 2013/01/04
 
@@
 \begin{proof}
-TODO (coinductive?)
+by induction on the path relation. Assume 
+$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S'}, \ot{O}{S''})$. It follows that
+this will have been derived by one of the following rules:
+
+\begin{itemize}
+
+\item \pathrefl. Directly, $\overline{S} = \overline{S'} = \overline{S''}$.
+
+\item \pathadj. It follows that there exists a method $m$ such that
+$m : T \Rightarrow \overline{S'} \in \ot{O}{S}$. There can be no other
+target state set for this method, therefore $\overline{S'} = \overline{S''}$.
+
+\item \pathtrans. It follows that there exists state sets $\overline{S_1}$
+and $\overline{S_2}$ such that
+$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_1}, \ot{O}{S_2})$ and
+$(\ot{O}{S_1}, \ot{O}{S_2}) \lesssim (\ot{O}{S'}, \ot{O}{S''})$. By induction,
+it follows that $\overline{S_1} = \overline{S_2}$. Therefore, also by induction,
+$\overline{S'} = \overline{S''}$.
+
+\item \pathbranch. It follows that there exists state sets
+$\overline{S_1}$, $\overline{S_2}$, $\overline{S_3}$, $\overline{S_4}$ such
+that 
+$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_1}, \ot{O}{S_2})$ and
+$(\ot{O}{S}, \ot{O}{S}) \lesssim (\ot{O}{S_3}, \ot{O}{S_4})$ where
+$\overline{S'} = \overline{S_1} \cup \overline{S_3}$ and
+$\overline{S''} = \overline{S_2} \cup \overline{S_4}$.
+By induction, $\overline{S_1} = \overline{S_2}$ and 
+$\overline{S_3} = \overline{S_4}$. Therefore, $\overline{S'} = \overline{S''}$.
+
+\end{itemize}
 \end{proof}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% revised and checked: 2013/01/04
 
@@
 $\overline{S} = \overline{S_1}$ for all $\overline{S}$. Therefore,
-$\overline{S_2} = \overline{S_1}$.
+$\overline{S_2} = \overline{S_1}$, meaning
+$T = \remapapp{T}{T}{T}$.
 \end{itemize}
 \end{proof}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%% lem:remapsame %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% revised and checked 2013/01/04
 
@@
 \label{lem:remapeff}
-Given types $T$, $U$ and $V$ such that $U <: V$.
-It follows that $U <: \remapapp{U}{V}{V}$.
+Let $U <: V$. It follows that $U <: \remapapp{U}{V}{V}$.
 \end{lem}
@@
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%% revised and checked: 2013/01/05
 
@@
 \begin{proof}
-by induction on the structure of $t$.
-
-If $t$ is some value $v$, then $\Gamma = \Gamma'$, therefore
-$\subst{x_i}{y_i}{\Gamma} = \subst{x_i}{y_i}{\Gamma'}$
-and substitution has no effect on the value, which has no specific requirements
-of either $\Gamma$ or $\Gamma'$ other than that they are equal.
+by induction on the typing derivation of $t$. $\typerule{\Gamma}{t}{T}{\Gamma'}$
+must have been derived by one of the following rules:
+
+\begin{itemize}
+
+\item \tunit, \ttrue, \tfalse, \tobj or \tfundef. For each of these rules,
+$\Gamma = \Gamma'$. Therefore, $\subst{x_i}{y_i}{\Gamma} = \subst{x_i}{y_i}{\Gamma'}$.
+and substitution has no effect on the value. The rule used for derivation
+has no specific requirements of either $\Gamma$ or $\Gamma'$ other than that 
+they are equal.
 Therefore, $\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{v}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
 
-If $t$ is a term, then it is of one of the following forms:
-
-\begin{itemize}
-\item $t = \lett{x}{t''}{t'}$. 
+\item \tlet. It follows that $t = \lett{x}{t_v}{t_b}$. 
 Substitution is defined on this term such that \\
 $\subst{x_i}{y_i}{t} \equiv 
-\lett{x}{( \subst{x_i}{y_i}{t''} )}{(\subst{x_j}{y_j}{t'})}$
+\lett{x}{( \subst{x_i}{y_i}{t_v} )}{(\subst{x_j}{y_j}{t_b})}$
 , where $\overline{y_j} = \overline{y_i} - \{ x \}$.
@@
 that there exists $T', T'', \Gamma_1$ such that
-$\typerule{\Gamma}{t''}{T''}{\Gamma_1}$ and that
-$\typerule{\Gamma_1, x : T''}{t''}{T}{\Gamma', x : T'}$.
+$\typerule{\Gamma}{t_v}{T''}{\Gamma_1}$ and that
+$\typerule{\Gamma_1, x : T''}{t_b}{T}{\Gamma', x : T'}$.
 Additionally, $x \notin dom(\Gamma)$.
 
-By induction, $t''$ can be substituted such that 
+By induction, $t_v$ can be substituted such that 
 $\typerule
 {\subst{x_i}{y_i}{\Gamma}}
-{\subst{x_i}{y_i}{t''}}
+{\subst{x_i}{y_i}{t_v}}
 {\subst{x_i}{y_i}{\Gamma_1}}
 $.
-Additionally, $t'$ can be substituted such that
-$(\subst{x_j}{y_j}{\Gamma_1}) , x : T'' \triangleright \subst{x_j}{y_j}{t''} \triangleleft \{ (\overline{x_j / y_j} \} \Gamma') , x : T'$ as it is guaranteed that $x \notin \overline{y_j}$.
+Additionally, $t_b$ can be substituted such that
+$(\subst{x_j}{y_j}{\Gamma_1}) , x : T'' \triangleright \subst{x_j}{y_j}{t_b} \triangleleft \{ (\overline{x_j / y_j} \} \Gamma') , x : T'$ as it is guaranteed that $x \notin \overline{y_j}$.
 As $dom(\Gamma') = dom(\Gamma)$, and that $x \notin dom(\Gamma)$, it follows that $\subst{x_i}{y_i}{\Gamma_1} = (\subst{x_j}{y_j}{\Gamma_1})$ and
@@
 
-\item $t = t_l ; t_r$.
+\item \tseq. It follows that $t = t_l ; t_r$,
+and there exists $\Gamma'', T'$ such that
+$\typerule{\Gamma}{t_l}{T'}{\Gamma''}$ and
+$\typerule{\Gamma''}{t_r}{T}{\Gamma'}$.
 Substitution is defined on this term such that
-$\{ \overline{x_i / y_i} \} ( t_l ; t_r ) \equiv \subst{x_i}{y_i}{t_l} ; \subst{x_i}{y_i}{t_r}$.
-
-By rule \tseq, there exists $\Gamma'', T'$ such that
-$\typerule{\Gamma}{t_l}{T'}{\Gamma''}$ and
-$\typerule{\Gamma''}{t_r}{T}{\Gamma'}$. By induction,
+$\subst{x_i}{y_i}{(t_l ; t_r)} \equiv \subst{x_i}{y_i}{t_l} ; \subst{x_i}{y_i}{t_r}$.
+
+By induction, $t_l$ and $t_r$ can be typed such that
 $\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t_l}}{T'}{\subst{x_i}{y_i}{\Gamma''}}$ and
 $\typerule{\subst{x_i}{y_i}{\Gamma''}}{\subst{x_i}{y_i}{t_r}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
-Therefore, $t$ can by typed \tseq after the substitution.
-
-\item $t = x_0 ( \overline{x_k} )$.
+Therefore, $t$ can by typed using \tseq such that
+
+$\typerule{\subst{x_i}{y_i}{\Gamma}}{\subst{x_i}{y_i}{t}}{T}{\subst{x_i}{y_i}{\Gamma'}}$.
+
+\item \tfunc. It follows that $t = z ( \overline{z_k} )$ and
+that $\Gamma = \Gamma_1, \overline{z_k : T_k}$,
+$\Gamma' = \Gamma_1, \overline{z_k : T_k'}$
+where $\Gamma_1(z) = \funt{U_k}{V_k}{T}$ with for each $k$,
+$T_k <: U_k$ and $T_k' = \remapapp{T_k}{U_k}{V_k}$. Each
+variable in $\overline{z_k} \cup \{ z \}$ is distinct.
+
 Substition is defined on this term such that
-$\subst{x_i}{y_i}{x_0} ( \overline{x_k} ) \equiv y_0 ( \overline{y_k} )$
-where for each $x \in \{ x_0, \overline{x_k} \}$,
-$x = y_j \implies y = x_j$, otherwise $y$ is the original value.
-
-For each substitution that occurs in the term, the variable must exist in
-both the input and output contexts by rule \tfunc, and will be substituted
-for the same name. Therefore, $t$ can be typed by \tfunc after the
-substitution.
-
-\item $t = x.m$.
+$\subst{x_i}{y_i}{z ( \overline{z_k} )} \equiv z' ( \overline{z_k'} )$
+where $z' = z$ if $z \notin \overline{y_i}$ or $z = x_i$ for the $x_i$ such
+that $z = y_i$, and similarly for all $z_i$.
+
+The uniqueness of each substitution pair ensures
+that no variable will be passed twice as part of the function call.
+Therefore, the conditions of \tfunc are satisfied such that
+
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
+
+\item \tmethc. It follows that $t = x.m$ with
+$\Gamma = \Gamma_1, x : \ot{O}{S}$ and
+$\Gamma' = \Gamma_1, x : \ot{O}{S'}$ such
+that $m : T \Rightarrow \overline{S'} \in \ot{O}{S}$.
+
 Substition is defined on this term such that
-$\subst{x_i}{y_i}{x.m} \equiv y.m$
-where $y = x_j$ if $x = y_j$ for some $y_j \in \overline{y_i}$.
-Otherwise, $y = x$.
-
-Rule \tmethc requires that $x$ exist in both $\Gamma$ and $\Gamma'$ and
-therefore any substitution of x in the term will be matched with the
-same substitution in the contexts. Therefore, $t$ can be typed by
-\tmethc after the substitution.
-
-\item $t = \ift{t_c}{t_t}{t_f}$. Substitution is defined on this term such
-that \\
-$\subst{x_i}{y_i}{(\ift{t_c}{t_t}{t_f})} \equiv 
-\ift{\substxy{t_c}}{\substxy{t_t}}{\substxy{t_f}}$.
-By rule \tif there exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$, $T_f$
+$\subst{x_i}{y_i}{x.m} \equiv z.m$
+where $z = x_j$ if $x = y_j$ for some $y_j \in \overline{y_i}$.
+Otherwise, $z = x$.
+
+The conditions of \tmethc are satisfied such that
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
+
+\item \tif. It follows that $t = \ift{t_c}{t_t}{t_f}$,
+where there exists $\Gamma_1$, $\Gamma_2$, $\Gamma_3$, $T_t$, $T_f$
 such that
@@
 $\typerule{\Gamma_1}{t_f}{T_f}{\Gamma_3}$ where
-$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$. By induction
+$\Gamma' = \Gamma_2 \sqcup \Gamma_3$ and $T = T_t \sqcup T_f$.
+
+Substitution is defined on this term such
+that \\
+$\subst{x_i}{y_i}{(\ift{t_c}{t_t}{t_f})} \equiv 
+\ift{\substxy{t_c}}{\substxy{t_t}}{\substxy{t_f}}$.
+
+By induction
 substitution can be performed on $t_c$, $t_t$ and $t_f$ such that
@@
 By the definition of substitution on contexts, 
-$\substxy{\Gamma_2} \sqcup \substxy{\Gamma_3} = \substxy{\Gamma'}$. Therefore,
-by rule \tif, 
+$\substxy{\Gamma_2} \sqcup \substxy{\Gamma_3} = \substxy{\Gamma'}$.
+
+The conditions of the rule \tif are satisfied such that
 $\typerule{\substxy{\Gamma}}{\substxy{\ift{t_c}{t_t}{t_f}}}{T}{\substxy{\Gamma'}}$.
+
+\item \tdrop. It follows that there exists a $\Gamma''$ such that
+$\typerule{\Gamma}{t}{T}{\Gamma', \Gamma''}$. By induction,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma', \Gamma''}}$.
+By the definition of substitution on contexts,
+$\substxy{\Gamma', \Gamma''} = \substxy{\Gamma'}, \substxy{\Gamma''}$.
+
+Therefore by rule \tdrop,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
+
+\item \tsub. It follows that
+$\typerule{\Gamma}{t}{T'}{\Gamma'}$ for some $T' <: T$. By induction,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T'}{\substxy{\Gamma'}}$.
+By rule \tsub,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
+
+\item \twiden. It follows that $\Gamma' = \Gamma_1, x : U$
+and there exists a $U' <: U$ such that
+$\typerule{\Gamma}{t}{T'}{\Gamma_1, x : U'}$. 
+By induction,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma_1, x : U'}}$.
+By rule \twiden,
+$\typerule{\substxy{\Gamma}}{\substxy{t}}{T}{\substxy{\Gamma_1, x : U}}$.
+
+\item \tnarrow. It follows that $\Gamma = \Gamma_1, x : U'$
+and there exists a $U$ such that $U' <: U$ and
+$\typerule{\Gamma_1, x : U}{t}{T'}{\Gamma'}$.
+By induction,
+$\typerule{\substxy{\Gamma_1, x : U}}{\substxy{t}}{T}{\substxy{\Gamma'}}$.
+By rule \tnarrow,
+$\typerule{\substxy{\Gamma_1, x : U'}}{\substxy{t}}{T}{\substxy{\Gamma'}}$
+
 \end{itemize}
@@
 \item $t = \funv{x_i}{T_i}{U_i}{t'}$. 
-As $t$ is typeable, it must be the case that $t'$ is typeable also.
+As $t$ is typeable, it must be the case that $t'$ is typeable also. Trivially
+$t'$ will contain no function calls as it is part of $t$.
 By induction, the term $t'$ has a principal typing
-$(\Gamma, T, \Gamma')$. It follows that for each $i$, if $x_i \notin dom(\Gamma)$
+$(\Gamma, T, \Gamma')$. In order for $t$ to be typeable, it must be the case
+that $dom(\Gamma) \subseteq \overline{x_i}$ --- the function body cannot use
+any variables which are not explicitly passed as a parameter.
+
+For each $i$, if $x_i \notin dom(\Gamma)$
 then $x_i$ is not used in $t'$. Let $\overline{x_j} \subseteq \overline{x_i}$
@@
 
-If the effect $T_i \gg U_i$ is specified for parameter $x_i$ then it
-must be the case that $T_i <: \Gamma_2(x_i)$ for each $i$ so that 
-$t$ is typeable by rule \tfunc.
-Let $\Gamma_3 = \overline{x_i : T_i'}$ such that:
-
-\begin{itemize}
-\item $T_i' = T_i$ if $T_i$ is specified
-\item $T_i' = \Gamma_2(x_i)$ otherwise.
-\end{itemize}
-
-By repeated application of \lemref{lem:upgrade}, there exists a $\Gamma_3'$ and 
-$T'$ such that
-$(\Gamma_3,T',\Gamma_3')$ is a typing of $t'$, where $T' <: T$,
-$dom(\Gamma_3') = dom(\Gamma_2')$ and 
-$\forall x \in \Gamma_3' . \Gamma_3'(x) <: \Gamma_2'(x)$.
-
-Let $U_i' = U_i$ if $U_i$ is specified or $U_i' = \Gamma_3'(x)$ otherwise.
-It must be the case that $\Gamma_3'(x) <: U_i'$ for each $i$ so that
-$t$ is typeable by rule \tfunc.
-Let $\Gamma_4 = \overline{x_i : U_i}$.
-By repeated application of rule \twiden, 
-$(\Gamma_3,T',\Gamma_4)$ is a typing of $t'$.
-
-By rule \tfundef it follows that
-$\tau = (\emptyset, \funt{T_i'}{U_i'}{T'}, \emptyset)$ is a typing of $t$. 
-There can be no stronger typing than this, as $T_i'$ has an upper bound
-determined by either the principal typing of $t'$ or the user specified
-type $T_i$. Similarly, $U_i'$ has a lower bound determined by
-modification of the principal typing of $t'$ or the user specified
-$U_i$. Therefore $\tau$ is the principal type of $t$.
+The effect $T_i \gg U_i$ for parameter $x_i$ must be such that 
+$T_i <: \Gamma_2(x_i)$ and $\Gamma_2'(x_i) <: U_i$ in order for
+$t$ to be typeable. 
+Trivially, $(\emptyset, \funt{T_i}{U_i}{T}, \emptyset)$ is the principal
+typing of $T$.
+
+It is worth noting that the explicit effects for each parameter are not
+actually necessary, and can be extracted directly from $\Gamma_2$
+and $\Gamma_2'$. By choosing $T_i' = \Gamma_2(x_i)$ and $U_i' = \Gamma_2'(x_i)$,
+$(\emptyset, \funt{T_i'}{U_i'}{T}, \emptyset)$ is a typing of the modified
+function value $\lambda (\overline{x_i}) . t'$, which can safely omit the
+effect types for each $x_i$.
+
+%! extended proof of construction of principal types when some annotations
+%! provided, and some are missing
+%
+% so that 
+% $t$ is typeable by rule \tfunc.
+% Let $\Gamma_3 = \overline{x_i : T_i}$.
+
+% By repeated application of \lemref{lem:upgrade}, there exists a $\Gamma_3'$ and 
+% $T'$ such that
+% $(\Gamma_3,T',\Gamma_3')$ is a typing of $t'$, where $T' <: T$,
+% $dom(\Gamma_3') = dom(\Gamma_2')$ and 
+% $\forall x \in \Gamma_3' . \Gamma_3'(x) <: \Gamma_2'(x)$.
+
+% Let $U_i' = U_i$ if $U_i$ is specified or $U_i' = \Gamma_3'(x)$ otherwise.
+% It must be the case that $\Gamma_3'(x) <: U_i'$ for each $i$ so that
+% $t$ is typeable by rule \tfunc.
+% Let $\Gamma_4 = \overline{x_i : U_i}$.
+% By repeated application of rule \twiden, 
+% $(\Gamma_3,T',\Gamma_4)$ is a typing of $t'$.
+
+% By rule \tfundef it follows that
+% $\tau = (\emptyset, \funt{T_i'}{U_i'}{T'}, \emptyset)$ is a typing of $t$. 
+% There can be no stronger typing than this, as $T_i'$ has an upper bound
+% determined by either the principal typing of $t'$ or the user specified
+% type $T_i$. Similarly, $U_i'$ has a lower bound determined by
+% modification of the principal typing of $t'$ or the user specified
+% $U_i$. Therefore $\tau$ is the principal type of $t$.
 
@@
 As $t$ is typeable, each $v_{ij}$ must be typeable also.
-By induction, each $v_{ij}$ has a principal typing $(\emptyset, T_{ij}, \emptyset)$,
-as values can be typed with arbitrary input contexts, and principal typings
-have the smallest input context domain possible.
-
-Trivially, $\tau = (\emptyset,\{ \overline{ S_i \{ \overline{ m_{ij} = T_{ij} \Rightarrow S_ij } \} } \}@S,\emptyset)$ is
+By induction, each $v_{ij}$ has a principal typing, which will be of
+form $(\emptyset, T_{ij}, \emptyset)$
+as values can be typed with no requirements on the input context.
+
+Directly, 
+$\tau = (\emptyset,\{ \overline{ S_i \{ \overline{ m_{ij} = T_{ij} \Rightarrow S_ij } \} } \}@S,\emptyset)$ is
 the principal typing of $t$.
@@
 
-Let $(\Gamma_{in}, \Gamma_{mid}, \Gamma_{out}) = seqctx(\Gamma_v, \Gamma_v', \Gamma_b'', \Gamma_b''')$.
+Let $(\Gamma, \Gamma_{mid}, \Gamma') = seqctx(\Gamma_v, \Gamma_v', \Gamma_b'', \Gamma_b''')$.
 If this is undefined, then there exists a $y \in dom(\Gamma_v) \cap dom(\Gamma_b'')$
@@
 
-By \lemref{lem:upgrade} and the definition of $\mathit{seqctx}$, 
-there exists a $T_v' <: T_v$ such that
-$(\Gamma_{in}, T_v', \Gamma_{mid})$ is a typing of $t_v$.
+By the definition of $\mathit{seqctx}$ it must be the case that
+$(\Gamma, T_v, \Gamma_{mid})$ is a typing of 
+$t_v$. By \lemref{lem:upgrade} there exists a $T_v' <: T_v$ such that
+$(\Gamma, T_v', \Gamma_{mid})$ is a typing of $t_v$.
 
@@
 Let $\Gamma_{mid}' = \Gamma_{mid}, x : T_v'$.
-By \lemref{lem:upgrade}, there exists a $T_b' <: T_b$ and $T_x'' <: remap(T_v', T_x \gg T_x')$ such that
+By \lemref{lem:upgrade}, there exists a $T_b' <: T_b$ and $T_x'' <: T_x'$ such that
 $(\Gamma_{mid}', T_b', (\Gamma_{out}, x : T_x''))$ is a typing of $t_b$.
@@
 By rule \tlet it follows that $(\Gamma_{in}, T_b', \Gamma_{out})$ is a
-typing of $t$. This typing must be principal.
-
-%%% TODO: proof of principality from construction?
+typing of $t$. This typing is principal.
+
+%%% TODO: parts of this solution are existentially quantified, and not shown
+%%% to be the smallest / largest types possible to get the principal typing 
+%%% result.
 
@@
 followed by the usage in $t_b$. 
-In this case, $t$ itself would be untypeable. Therefore, it must be defined.
-
-By the definition of $\mathit{seqctx}$, we have that for each
-$x \in dom(\Gamma_a)$ that $\Gamma(x) <: \Gamma_a(x)$, and
-that $\Gamma_{mid}(x) <: \remapapp{\Gamma(x)}{\Gamma_a(x)}{\Gamma_a'{x}}$.
-
-%%% TODO: Prove that
-It follows that $(\Gamma, T_a', \Gamma_{mid})$ is a typing of
-$t_a$.
-
-%%% TODO: prove that
-It follows that $(\Gamma_{mid}, T_b', \Gamma_{mid})$ is a typing of
-$t_b$.
-
-It follows by rule \tseq that $(\Gamma_{in}, T_b', \Gamma_{out})$ is a typing
-of $t$.
-
-%%% TODO: show this is minimal.
+In this case, $t$ itself would be untypeable. Therefore, 
+$(\Gamma, \Gamma_{mid}, \Gamma')$ is defined.
+
+By the definition of $\mathit{seqctx}$, the types assigned to each variable
+in $\Gamma$ are the least specific that can support the usage of each variable
+in $t_a$ followed by $t_b$. It must therefore be the case 
+that $(\Gamma, T_a, \Gamma_{mid})$ and $(\Gamma_{mid}, T, \Gamma')$ are typings
+of $t_a$ and $t_b$ respectively. Therefore, by rule \tseq,
+$(\Gamma, T, \Gamma')$ is a typing of $t$.
+
+This typing has the least specific
+possible input context and most specific possible output context by the 
+definition of $\mathit{seqctx}$ and its
+application to the principal typings of $t_a$ and $t_b$. The type $T$ is
+not necessarily the most specific type that can be ascribed to the term $t$
+given this context however: for each $x \in dom(\Gamma_b)$, by the definition
+of $\mathit{seqctx}$ we have that $\Gamma_{mid}(x) <: \Gamma_b(x)$.
+
+It follows by \lemref{lem:upgrade} that there exists a $T' <: T$
+such that $(\Gamma_{mid}, T', \Gamma')$ is a typing of
+$t_b$, therefore $(\Gamma, T', \Gamma')$ is a typing of $T$. 
+It follows that $(\Gamma, T', \Gamma')$
+is the principal typing of $t$.
+
+%%% DODGY need better way to construct type T' type or bound it.
 
@@
 
-In order for $t$ to be typeable, it must be the case that a typing
-$(\Gamma_c'', \boolt, \Gamma_c''')$ exists for $t_c$. It follows that as
-$\tau_c$ is a principal typing for $t_c$, that for each $x \in \Gamma_a$ that 
-$\Gamma_a''(x) <: \Gamma_a(x)$ and that $\Gamma_a'(x) <: \Gamma_a'''(x)$,
-with $\boolt <: T_c$.
+In order for $t$ to be typeable, it must be the case that a 
+$T_c <: \boolt$.
 
@@
 \end{thm}
-
-\begin{proof}
-by induction on the structure of $t$.
-
-\end{proof}
 

### 2013-01-07 src/main/scala/AST.scala
resync source, building TODO of all remaining tasks to wrap this work up

@@ -21,34 +21,67 @@ trait SourceElement extends Attributable with Positioned
 sealed abstract class Term extends SourceElement
 sealed abstract class Value extends Term
 
-case class UnitValue() extends Value
-case class TrueValue() extends Value
-case class FalseValue() extends Value
+case class UnitValue() extends Value { override def toString = "unit" }
+case class TrueValue() extends Value { override def toString = "true" }
+case class FalseValue() extends Value { override def toString = "false" }
 case class ObjValue(states : Seq[StateDef], state : String) extends Value {
   val stateMap =
     (states.foldLeft
       (Map.empty[String,StateDef])
       ((m, s) => m + (s.name -> s))
     )
+
+  override def toString = states.mkString("{", " ", "}") + "@" + state
 }
 
-case class FunValue(params : Seq[ParamDef], body : Term) extends Value
+case class FunValue(params : Seq[ParamDef], body : Term) extends Value {
+  override def toString = params.mkString("\\(", ", ", ").") + "(" + body + ")"
+}
 
-case class LetBind(varName : String, value : Term, body : Term) extends Term
-case class Update(varName : String, body : Term) extends Term
-case class MethCall(objVarName : String, methName : String) extends Term
-case class Sequence(left : Term, right : Term) extends Term
-case class FunCall(funName : String, paramNames : Seq[String]) extends Term
-case class If(condition : Term, whenTrue : Term, whenFalse : Term) extends Term
+case class LetBind(varName : String, value : Term, body : Term) extends Term {
+  override def toString = 
+    "let " + varName + " = (" + value + ") in (" + body + ")"
+}
+
+case class Update(varName : String, body : Term) extends Term {
+  override def toString =
+    varName + ":= (" + body + ")"
+}
+
+case class MethCall(objVarName : String, methName : String) extends Term {
+  override def toString = objVarName + "." + methName
+}
+
+case class Sequence(left : Term, right : Term) extends Term {
+  override def toString = left + " ; " + right
+}
+
+case class FunCall(funName : String, paramNames : Seq[String]) extends Term {
+  override def toString = funName + paramNames.mkString("(", ", ", ")")
+}
+
+case class If(condition : Term, whenTrue : Term, whenFalse : Term) extends Term {
+  override def toString = 
+    "if (" + condition + ") then (" + whenTrue + ") else (" + whenFalse + ")"
[…]

### 2013-01-07 src/main/scala/ConstraintGenerator.scala

@@ -287,7 +287,7 @@ object ConstraintGenerator {
   def ifConstraints(t : If) =
     (ConstraintSet() +
       EqualityConstraint(VarTE(t.condition->typeVar), BoolTE) +
-      EqualityConstraint(VarTE(t->typeVar), JoinTE(VarTE(t.whenTrue->typeVar), VarTE(t.whenFalse->typeVar))) +
+      EqualityConstraint(VarTE(t->typeVar), SeparateJoinTE(VarTE(t.whenTrue->typeVar), VarTE(t.whenFalse->typeVar))) +
       ContextConstraint(t.condition->inContextVar, sameAs(t->inContextVar)) +
       ContextConstraint(t.whenTrue->inContextVar, sameAs(t.condition->outContextVar)) +
       ContextConstraint(t.whenFalse->inContextVar, sameAs(t.condition->outContextVar)) +

### 2013-01-07 src/main/scala/ObjectConstraintSolver.scala

@@ -23,26 +23,11 @@ class ObjectConstraintSolver(private[ts] val constraints : Seq[TypeConstraint]) 
   private[ts] var objects = new VarSolutionSet[StateGraph]()
   private[ts] var states = new VarSolutionSet[Set[String]]()
 
-  // solving cases:
-  // 1. solve subtype constraint where upper bound solved
-  // 2. solve remap where effect is solved
-  // 3. solve equality
-  // 4. solve join / meet
-
-  // we assume all constraints have been type-normalised
-  // so that wherever a type is referred to it is always
-  // of form O@S. Specifically, don't want
-  // a = O@S join O'@S'
-  // instead want
-  // O@S = O'@S' join O''@S''
-
-  def solve() {
-    val partition = partitionConstraints()
-    resolveDirectObjectEqualities(partition.objectEqualities)
-    resolveDirectStateEqualities(partition.stateEqualities)
-
-    partition.definitions.foreach(solveDefinition(_))
-  }
+  def getCurrentSolution(o : ObjectTE) 
+      : Option[(StateGraph, Option[Set[String]])] =
+    objects.
+      getSolution(o.objVar).
+      map(os => (os, states.getSolution(o.stateVar)))
 
   def getSolutionFor(o : ObjectTE) : Option[SolvedObjectTE] = {
     val objGraphOpt = objects.getSolution(o.objVar)
@@ -53,90 +38,189 @@ class ObjectConstraintSolver(private[ts] val constraints : Seq[TypeConstraint]) 
     ))
   }
 
+  def solve() : Boolean = {
+    val (objectEqualities, stateEqualities, otherConstraints) = 
+      extractEqualities()
+
+    resolveDirectObjectEqualities(objectEqualities)
+    resolveDirectStateEqualities(stateEqualities)
+
+    val constraintsByObject = partitionConstraintsByObject(otherConstraints)
+
+    val solveOrder = deriveSolveOrder(constraintsByObject)
+
+    if(solveOrder.isEmpty) return false;
+
+    solveOrder.get.foreach(o => solveConstraints(constraintsByObject(o)))
+
+    return true;
+  }
+
   type DirectEquality = Pair[TypeVar, TypeVar]
-  type Definition = Pair[ObjectTE, TypeExpr]
 
-  case class ConstraintPartition(
-    objectEqualities : Seq[DirectEquality],
-    stateEqualities : Seq[DirectEquality],
-    subtypes : Seq[SubtypeConstraint],
-    definitions : Seq[Definition]
-  )
-
[…]

### 2013-01-07 src/main/scala/Relations.scala

@@ -11,7 +11,7 @@
 
 package uk.ac.gla.dcs.ts
 
-class Relation[T,U](private val relation : Set[(T,U)]) {
+class Relation[T,U](val relation : Set[(T,U)]) {
 
   def findRightEquivs(left : T) : Set[U] =
     relation.filter(_._1 == left).map(_._2)

### 2013-01-07 src/main/scala/TypeExpressions.scala

@@ -38,9 +38,19 @@ case class ObjectTE(objVar : TypeVar, stateVar : TypeVar) extends TypeExpr {
 
 case class SolvedObjectTE(
   graph : StateGraph, 
-  states : Set[String]) extends TypeExpr
+  states : Set[String]) extends TypeExpr {
+}
+
+object SolvedObjectTE {
+  def apply(graph : StateGraph, state : String) : SolvedObjectTE = 
+    SolvedObjectTE(graph, Set(state))
+}
 
 case class JoinTE(left : TypeExpr, right : TypeExpr) extends TypeExpr {
+  override def toString = left + " ⊻ " + right
+}
+
+case class SeparateJoinTE(left : TypeExpr, right : TypeExpr) extends TypeExpr {
   override def toString = left + " ⊔ " + right
 }
 
@@ -80,46 +90,53 @@ object TypeExprUtil {
   type TVEquivalence = Equivalence[TypeVar,TypeVar]
 
   def isomorphic(
-    t1 : TypeExpr, 
-    t2 : TypeExpr, 
-    equivVars : TVEquivalence) 
-    : Boolean = {
+      t1 : TypeExpr, 
+      t2 : TypeExpr, 
+      equivVars : TVEquivalence) 
+      : Boolean =
+    checkIsomorphic(t1, t2, Bijection(equivVars.toMap)).isDefined
 
+  def checkIsomorphic(
+      t1 : TypeExpr,
+      t2 : TypeExpr,
+      equivVars : Bijection[TypeVar, TypeVar]) 
+      : Option[Bijection[TypeVar, TypeVar]] =
     (t1, t2) match {
-      case (UnitTE, UnitTE) => true
-      case (VarTE(v1), VarTE(v2)) => equivVars.addEquivalence(v1, v2)
+      case (UnitTE, UnitTE) => Some(equivVars)
+      case (VarTE(v1), VarTE(v2)) => equivVars.addEquivalence(v1,v2)
       case (FunTE(f1params, f1ret), FunTE(f2params, f2ret)) =>
         functionsIsomorphic(f1params, f1ret, f2params, f2ret, equivVars)
       case (ObjectTE(o1v, o1s), ObjectTE(o2v, o2s)) =>
-        (equivVars.addEquivalence(o1v, o2v) &&
-          equivVars.addEquivalence(o1s, o2s))
[…]

### 2013-01-07 src/main/scala/TypeVarEquivalence.scala

@@ -23,13 +23,50 @@ class Equivalence[T, U](private var equivMap : Map[T, U]) {
     }
   }
 
+  // TODO: what about the reverse, t2 is equivalent to something else?
   def checkEquivalent(t1 : T, t2 : U) : Option[Boolean] =
     equivMap.get(t1).map(_ == t2)
 
   def getRightEquivalent(t1 : T) : Option[U] = equivMap.get(t1)
 
+  def toMap = equivMap
+
 }
 
 object Equivalence {
   def empty[T, U] = new Equivalence(Map.empty[T, U])
+}
+
+class Bijection[T,U](private val t2u : Map[T,U], private val u2t : Map[U,T]) {
+  
+  def addEquivalence(t : T, u : U) : Option[Bijection[T,U]] =
+    (checkEquivalent(t,u).
+      flatMap(if(_) Some(this) else None).
+      orElse { Some(new Bijection(t2u + (t -> u), u2t + (u -> t))) }
+    )
+
+  def checkEquivalent(t : T, u : U) : Option[Boolean] = {
+    // as the mappings are symmetric, if (t -> u) does not exist in t2u
+    // then there must not be a mapping (u -> v) for some v in u2t, otherwise
+    // they are not equivalent.
+    t2u.get(t).map(_ == u) orElse (u2t.get(u).map(_ => false))
+  }
+
+  def getRightEquivalent(t : T) : Option[U] = t2u.get(t)
+  def getLeftEquivalent(u : U) : Option[T] = u2t.get(u)
+}
+
+object Bijection {
+  def empty[T,U] = new Bijection[T,U](Map.empty, Map.empty)
+  def apply[T,U](t2u : Map[T,U]) : Bijection[T,U] = {
+    val u2t = (t2u.foldLeft
+      (Map.empty[U,T])
+      { case (revMap, (t,u)) =>
+        if(revMap contains u) throw new IllegalArgumentException()
+        revMap + (u -> t)
+      }
+    )
+
+    new Bijection(t2u, u2t)
+  }
 }

### 2013-01-07 src/main/scala/VarSolutionSet.scala

@@ -24,6 +24,9 @@ class VarSolutionSet[T] {
     solutions(v)
   }
 
+  def getCanonicalVars() =
+    solutions.values.foldLeft(Set.empty[TypeVar])(_ + _()._1)
+
   def getCanonicalEquiv(v : TypeVar) =
     getProxy(v).get()._1
 
@@ -33,7 +36,7 @@ class VarSolutionSet[T] {
   def getSolutionOrFail(v : TypeVar) : T = getSolution(v).get
 
   def areEquivalent(v1 : TypeVar, v2 : TypeVar) =
-    getProxy(v1).get()._1 == getProxy(v2).get()._2
+    getProxy(v1).get()._1 == getProxy(v2).get()._1
 
   def makeEquivalent
       (v1 : TypeVar, v2 : TypeVar) {

### 2013-01-07 src/main/scala/sm/StateGraphUtils.scala

@@ -179,6 +179,7 @@ object StateGraphUtils {
       : (StateGraph, StateNameEquiv) = {
 
     val (combinedGraph, g2RelabeledEquiv) = includeInto(g1, g2)
+    val g2RelabeledNames = g2RelabeledEquiv.relation.map(_._2)
 
     val sg = new StateGenerator(g1.nodes.toNodeInSet)
 
@@ -193,14 +194,18 @@ object StateGraphUtils {
     var stateMap : Map[Set[_ <: StateGraph#NodeT], State] = 
       Map(startSet -> g1startInner.value)
 
-    var g2OverlayEquiv = Set.empty[(String, String)]
+    var g2OverlayEquiv = g2RelabeledNames.map(n => (n,n))
     var overlayGraph = combinedGraph
 
     visit(startSet)(stateSet => {
       val partition = partitionEdges(stateSet)
 
-      partition.disjoint.foreach(e => 
-        overlayGraph += stateMap(stateSet).value ~> e.edge.to.value by e.edge.m)
+      val g2State = stateSet.filter(g2RelabeledNames contains _.value.name).head
+      g2OverlayEquiv += (g2State.value.name -> stateMap(stateSet).value.name)
+
+      partition.disjoint.foreach { e => 
+        overlayGraph += stateMap(stateSet).value ~> e.edge.to.value by e.edge.m
+      }
 
       val visitList = (partition.shared map { case (methodName, edgeSet) => {
 
@@ -217,9 +222,33 @@ object StateGraphUtils {
       Right(visitList)
     })
 
+    // remove all nodes from g2 which are not reachable from the start point
+    val (cleanedGraph, removedStates) = 
+      removeUnreachable(overlayGraph, g1start, g2RelabeledNames)
+
+    // remove all dead entries from the equivalence relation
+    g2OverlayEquiv = g2OverlayEquiv.filterNot(removedStates contains _._2)
+
     val g2Equiv = g2RelabeledEquiv.compose(Relation(g2OverlayEquiv))
 
-    (overlayGraph, g2Equiv)
+    (cleanedGraph, g2Equiv)
+  }
+
+  def removeUnreachable(
+      g : StateGraph, 
+      start : String, 
+      removeCandidates : Set[String])
+      : (StateGraph, CSet[String]) = {
+
+    val unconnected = 
+      findUnconnected(g, State(start)).filter { (s : StateGraph#NodeT) => 
[…]

### 2013-01-07 src/test/scala/ObjectConstraintSolverTest.scala

@@ -26,72 +26,87 @@ class ObjectConstraintSolverTest extends FunSuite with ShouldMatchers {
   implicit def intToTypeVar(i : Int) : TypeVar = TypeVar(i)
   def tvp(tv1 : Int, tv2 : Int) = Pair(TypeVar(tv1), TypeVar(tv2))
 
+  def getSolutionsOrFail
+      (objIds : Range)
+      (implicit solver : ObjectConstraintSolver) 
+      : Seq[SolvedObjectTE] =
+    getSolutionsOrFail(objIds map (obj(_)))
+
+  def getSolutionsOrFail
+    (objects : Seq[ObjectTE])
+    (implicit solver : ObjectConstraintSolver) : Seq[SolvedObjectTE] = {
+
+    objects map { o =>
+      val soln = solver.getSolutionFor(o)
+      assert(soln.isDefined, "solution for " + o + " not found")
+      soln.get
+    }
+  }
+
+  def obj(id : Int) = ObjectTE(objVar(id), stVar(id))
+  def objVar(id : Int) = id*2
+  def stVar(id : Int) = id*2+1
+
   test("resolve direct equalities") {
     val constraints = Seq(
-      EqualityConstraint(ObjectTE(1,2),ObjectTE(3,4)),
-      EqualityConstraint(ObjectTE(3,4),ObjectTE(5,6)),
-      EqualityConstraint(ObjectTE(7,8),ObjectTE(9,10)),
-      EqualityConstraint(ObjectTE(7,8),ObjectTE(11,12)),
-      EqualityConstraint(ObjectTE(13,14),ObjectTE(5,6))
-    )
-
-    val ocs = new ObjectConstraintSolver(constraints)
-    val partition = ocs.partitionConstraints()
-    ocs.resolveDirectObjectEqualities(partition.objectEqualities)
-    ocs.resolveDirectStateEqualities(partition.stateEqualities)
-
-    ocs.objects.getCanonicalEquiv(3) should be (intToTypeVar(1))
-    ocs.objects.getCanonicalEquiv(5) should be (intToTypeVar(1))
-    ocs.objects.getCanonicalEquiv(13) should be (intToTypeVar(1))
-
-    ocs.objects.getCanonicalEquiv(9) should be (intToTypeVar(7))
-    ocs.objects.getCanonicalEquiv(11) should be (intToTypeVar(7))
-
-    ocs.states.getCanonicalEquiv(4) should be (intToTypeVar(2))
[…]

### 2013-01-07 src/test/scala/TestUtils.scala

@@ -61,6 +61,11 @@ object TestUtils {
 
   implicit def strToUpdateHelper(str : String) = UpdateHelper(str)
 
+  implicit def typeToConstraintBuilder(typ : TypeExpr) = 
+    new ConstraintBuilder(typ)
+
+  def remap(input : TypeExpr, effect : EffectTE) = RemapTE(input, effect)
+
   /* state and method creation helpers */
 
   def s(name : String) = State(name)
@@ -70,6 +75,34 @@ object TestUtils {
   def methodGraph(m : Method) : (StateGraph, String, String) = 
     (Graph(s("A") ~> s("B") by m), "A", "B")
 
+  def call(m : Method, param : EffectTE, eff : EffectTE) 
+      : Seq[TypeConstraint] = {
+        
+    val (g, inState, outState) = methodGraph(m)
+    val (effIn, effOut) = extractObjects(eff)
+    val (paramIn, paramOut) = extractObjects(param)
+
+    Seq(
+      paramOut =-= remap(paramIn, eff),
+      effIn =-= SolvedObjectTE(g, inState),
+      effOut =-= SolvedObjectTE(g, outState)
+    )
+  }
+
+  def extractObjects(eff : EffectTE) = 
+    eff match {
+      case EffectTE(in @ ObjectTE(_,_), out @ ObjectTE(_,_)) => (in,out)
+      case _ => throw new IllegalArgumentException()
+    }
+    
+
+  val emptyGraph : StateGraph = Graph(s("A"))
+  val emptyObject = SolvedObjectTE(emptyGraph, "A")
+
+}
+
+class ConstraintBuilder(typ : TypeExpr) {
+  def =-=(other : TypeExpr) = EqualityConstraint(typ, other)
 }
 
 class TestUtilsTest extends org.scalatest.FunSuite {

### 2013-01-07 src/test/scala/TypeInferenceTest.scala

@@ -21,6 +21,7 @@ import org.kiama.util.Messaging._
 
 import TypeChecker._
 import sm._
+import TestUtils._
 
 import grizzled.slf4j.Logger
 
@@ -112,14 +113,7 @@ class TypeInferenceTest extends FunSuite with ShouldMatchers {
   // object tests
   infTest("object value, no methods",
     """[ S1 { } ]@S1""")((t, inCtx, freeVars, te, outCtx) => {
-
-    checkIsomorphic(
-      te, 
-      SolvedObjectTE(
-        Graph(State("S1")),
-        Set("S1")
-      )
-    )
+    checkIsomorphic(te, emptyObject)
   })
 
   infTest("object value, single method",

### 2013-01-07 src/test/scala/sm/PermuteTest.scala

@@ -1,0 +1,57 @@
+/*
+ * This file is part of TS.
+ * 
+ * TS - Copyright (c) 2012, Iain McGinniss
+ * All rights reserved.
+ * 
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted under the terms of the 2-clause BSD license:
+ * http://opensource.org/licenses/BSD-2-Clause
+ */
+
+package uk.ac.gla.dcs.ts.sm
+
+import org.scalatest.FunSuite
+import org.scalatest.matchers.ShouldMatchers
+
+import StateGraphUtils._
+
+class PermuteTest extends FunSuite with ShouldMatchers {
+
+  val empty = Set.empty[Int]
+  val emptyPairs = Set.empty[Set[(Int,Int)]]
+
+  test("empty sets") {
+    (emptyPairs) should equal (permuteMatch(empty, empty))
+  }
+
+  test("mismatched set sizes") {
+    (emptyPairs) should equal (permuteMatch(Set(1), Set()))
+  }
+
+  test("single value sets") {
+    (Set(Set(1 -> 2))) should equal (permuteMatch(Set(1), Set(2)))
+  }
+
+  test("two value sets") {
+    val expected = 
+      Set(Set(1 -> 3, 2 -> 4),
+          Set(1 -> 4, 2 -> 3))
+
+    val actual = permuteMatch(Set(1,2), Set(3,4))
+    (expected) should equal (actual)
+  }
+
+  test("three value sets") {
+    val expected =
+      Set(Set(1 -> 4, 2 -> 5, 3 -> 6),
+          Set(1 -> 4, 2 -> 6, 3 -> 5),
+          Set(1 -> 5, 2 -> 4, 3 -> 6),
+          Set(1 -> 5, 2 -> 6, 3 -> 4),
+          Set(1 -> 6, 2 -> 4, 3 -> 5),
+          Set(1 -> 6, 2 -> 5, 3 -> 4))
+
+    val actual = permuteMatch(Set(1,2,3), Set(4,5,6))
+    (expected) should equal (actual)
+  }
+}

### 2013-01-07 src/main/scala/AST.scala
strong update removed, top type introduced and join rewritten, more tests required

@@ -43,11 +43,6 @@ case class LetBind(varName : String, value : Term, body : Term) extends Term {
     "let " + varName + " = (" + value + ") in (" + body + ")"
 }
 
-case class Update(varName : String, body : Term) extends Term {
-  override def toString =
-    varName + ":= (" + body + ")"
-}
-
 case class MethCall(objVarName : String, methName : String) extends Term {
   override def toString = objVarName + "." + methName
 }
@@ -97,29 +92,33 @@ case class MissingState(
 
 sealed abstract class Type extends Attributable {
   def >>(outType : Type) = EffectType(this, outType)
-  def join(other : Type) : Option[Type] = (this, other) match {
-    case (UnitType(), UnitType()) => Some(UnitType())
-    case (BoolType(), BoolType()) => Some(BoolType())
+  def join(other : Type) : Type = (this, other) match {
+    case (UnitType(), UnitType()) => UnitType()
+    case (BoolType(), BoolType()) => BoolType()
     case (FunType(f1p, f1r), FunType(f2p, f2r)) => {
-      if(f1p.size != f2p.size) None
+      if(f1p.size != f2p.size) TopType()
       else {
-        val retTypeOpt = f1r join f2r
-        val paramTypesOpt = (f1p.zip(f2p).foldRight
-          (Option(Seq.empty[EffectType]))
-          ((paramPair, resOpt) => resOpt.flatMap(res => {
-            val (eff1, eff2) = paramPair
-            (eff1 join eff2) map (_ +: res)
-          }))
-        )
+        val retType = f1r join f2r
+        val paramTypesOpt = 
+          (f1p.zip(f2p).foldRight
+            (Option(Seq.empty[EffectType]))
+            ((paramPair, resOpt) => resOpt.flatMap(res => {
+              val (eff1, eff2) = paramPair
+              (eff1 join eff2) map (_ +: res)
+            }))
+          )
 
-        paramTypesOpt.flatMap(p => retTypeOpt.map(r => FunType(p, r)))
+        val typeOpt = 
+          paramTypesOpt.map(p => FunType(p, retType))
+
+        typeOpt.getOrElse(TopType())
       }
     }
     case (a @ ObjType(_, _), b @ ObjType(_, _)) =>
       a.joinObj(b)
-    case (ErrorType(),_) => Some(ErrorType())
-    case (_, ErrorType()) => Some(ErrorType())
-    case _ => None
+    case (ErrorType(),_) => ErrorType()
[…]

### 2013-01-07 src/main/scala/ConstraintGenerator.scala

@@ -94,7 +94,6 @@ object ConstraintGenerator {
       case FunValue(_,body) => allConstraints(body)
       case LetBind(_,value,body) => 
         allConstraints(value) ++ allConstraints(body)
-      case Update(_, body) => allConstraints(body)
       case Sequence(left, right) => 
         allConstraints(left) ++ allConstraints(right)
       case If(cond,thn,els) => 
@@ -132,7 +131,6 @@ object ConstraintGenerator {
       case t : ObjValue   => objectValueConstraints(t)
       case t : FunValue   => funValueConstraints(t)
       case t : LetBind    => letBindConstraints(t)
-      case t : Update     => updateConstraints(t)
       case t : Sequence   => sequenceConstraints(t)
       case t : FunCall    => funCallConstraint(t)
       case t : MethCall   => methCallConstraint(t)
@@ -198,17 +196,6 @@ object ConstraintGenerator {
         ContextRemoval(t.body->outContextVar, t.varName)) +
       EqualityConstraint(VarTE(t->typeVar), VarTE(t.body->typeVar))
     )
-
-  def updateConstraints(t : Update) = {
-    val bodyType = VarTE(t.body->typeVar)
-    (ConstraintSet() +
-      ContextConstraint(t.body->inContextVar, 
-        removeFrom(t->inContextVar, t.varName)) +
-      ContextConstraint(t->outContextVar, 
-        addTo(t.body->outContextVar, t.varName, bodyType)) +
-      EqualityConstraint(VarTE(t->typeVar), UnitTE)
-    )
-  }
 
   def sequenceConstraints(t : Sequence) =
     (ConstraintSet() +
@@ -322,8 +309,6 @@ object ConstraintGenerator {
         printInferenceTree(value, indent + 1)
         printInferenceTree(body, indent + 1)
       }
-      case Update(_, body) =>
-        printInferenceTree(body, indent + 1)
       case Sequence(left, right) => {
         printInferenceTree(left, indent + 1)
         printInferenceTree(right, indent + 1)

### 2013-01-07 src/main/scala/Parser.scala

@@ -16,7 +16,6 @@ trait Parser extends org.kiama.util.PositionedParserUtilities {
   lazy val term : PackratParser[Term] =
     (term <~ ";") ~ term                                  ^^ Sequence |
     ("let" ~> ident <~ "=") ~ (subterm <~ "in") ~ term    ^^ LetBind |
-    (ident <~ ":=") ~ subterm                             ^^ Update |
     subterm
 
   lazy val subterm : PackratParser[Term] =

### 2013-01-07 src/main/scala/TypeChecker.scala

@@ -102,7 +102,6 @@ object TypeChecker {
 
       /* terms */
       case LetBind(_,_,body) => body->ttype
-      case Update(_,_) => UnitType()
       case Sequence(l,r) => r->ttype
       case t @ MethCall(o,m) => 
         ((t->input)(o)) match { 
@@ -117,11 +116,7 @@ object TypeChecker {
       case t @ If(cond,thn,els) => {
         val t1 = thn->ttype
         val t2 = els->ttype
-        t1.join(t2).getOrElse {
-          message(t, "result types of branches are incompatible: %s and %s".
-            format(t1, t2))
-          ErrorType()
-        }
+        t1.join(t2)
       }
     }
 
@@ -144,10 +139,6 @@ object TypeChecker {
         case p @ LetBind(name,value,body) => {
           if (t eq value) p->input
           else value->output + Pair(p.varName,p.value->ttype)
-        }
-        case p @ Update(varName,_) => {
-          val ctx = (p->input) - varName
-          ctx
         }
         case p @ Sequence(left, right) => 
           (if (t eq left) p->input else p.left->output)
@@ -182,7 +173,6 @@ object TypeChecker {
     attr {
       case t : Value => t->input
       case LetBind(varName,valTerm,bodyTerm) => bodyTerm->output - varName
-      case Update(varName,body) => body->output + Pair(varName, body->ttype)
       case t @ MethCall(objName,methName) => methCallOutput(t)
       case Sequence(l,r) => r->output
       case t : FunCall => funCallOutput(t)
@@ -268,9 +258,6 @@ object TypeChecker {
           message(t, err match {
             case DifferentDomains(_, _) =>
               "output domains differ"
-            case MismatchedTypes(varName, t1, t2) =>
-              ("variable %s has incompatible types in the output of " +
-              "the branches: %s and %s") format (varName, t1, t2)
           })
         })
         (t.whenTrue->output).mapValues(_ => ErrorType())
@@ -350,10 +337,6 @@ class FullTracePrinter(term : Terminal) extends org.kiama.output.PrettyPrinter {
         val termDoc = 
           LET <+> varName <+> "=" <+> valueTermId <+> IN <+> bodyTermId
         Pair(termDoc, Seq(value, body))
-      }
-      case Update(varName, body) => { 
-        val bodyTermId = body->termId
-        Pair(varName <+> ":=" <+> (bodyTermId), Seq(body))
       }
       case MethCall(objVarName, methName) => 
         Pair(objVarName <> "." <> methName, Seq.empty)

### 2013-01-07 src/main/scala/package.scala

@@ -34,8 +34,6 @@ package object ts {
   sealed abstract class JoinError
   case class DifferentDomains(leftDiff : Set[String], rightDiff : Set[String])
     extends JoinError
-  case class MismatchedTypes(varName : String, t1 : Type, t2 : Type)
-    extends JoinError
 
   def joinContexts(c1 : Context, c2 : Context) : Either[Seq[JoinError],Context] = {
     if(c1.keySet != c2.keySet) {
@@ -44,24 +42,10 @@ package object ts {
       return Left(Seq(DifferentDomains(c1Extra, c2Extra)))
     }
 
-    val (errors, ctx) = (c1.foldLeft
-      (Pair(Seq.empty[JoinError], emptyContext))
-      ((res, p) => {
-        val errors = res._1
-        val ctx = res._2
-        val varName = p._1
-        val varType = p._2
-        val otherType = c2(varName)
-
-        (varType.join(otherType).
-          map(joinedType => Pair(errors, ctx + (varName -> joinedType))).
-          getOrElse(Pair(MismatchedTypes(varName, varType, otherType) +: errors, 
-            ctx))
-        )
-      })
-    )
-
-    if(errors.isEmpty) Right(ctx) else Left(errors)
+    Right(c1.map { case (varName, varType) => 
+      val otherType = c2(varName)
+      varName -> varType.join(otherType)
+    })
   }
 
   /** A typing context, where variables are mapped to type expressions

### 2013-01-07 src/test/scala/ParserTest.scala

@@ -98,8 +98,9 @@ class ParserTest extends FunSuite {
   }
 
   test("parse if-then-else with sequences in branches") {
-    checkTerm("if true then (x := unit; y := unit) else (y := unit; x := unit)",
-      If(truev, seq("x" := unitv, "y" := unitv), seq("y" := unitv, "x" := unitv))
+    checkTerm("if true then (x.m; y.n) else (y.n; x.m)",
+      If(truev, seq(MethCall("x", "m"), MethCall("y", "n")), 
+                seq(MethCall("y", "n"), MethCall("x", "m")))
     )
   }
 }

### 2013-01-07 src/test/scala/TestUtils.scala

@@ -35,9 +35,9 @@ object TestUtils {
   def seq(ts : Term*) : Term = {
     ts.reduceRight((t1, t2) => Sequence(t1, t2))
   }
-  def up(name : String, t : Term) = Update(name, t)
 
   /* type short-hand creation */
+  def topt = { TopType() }
   def unitt = { UnitType() }
   def boolt = { BoolType() }
   def funt(retType : Type, paramTypes : EffectType*) =
@@ -54,12 +54,6 @@ object TestUtils {
   /* type inference related helpers */
   implicit def intToVarTE(i : Int) = VarTE(TypeVar(i))
   implicit def intToTypeVar(i : Int) = TypeVar(i)
-
-  case class UpdateHelper(varName : String) {
-    def :=(t : Term) = Update(varName, t)
-  }
-
-  implicit def strToUpdateHelper(str : String) = UpdateHelper(str)
 
   implicit def typeToConstraintBuilder(typ : TypeExpr) = 
     new ConstraintBuilder(typ)

### 2013-01-07 src/test/scala/TypeCheckerTest.scala

@@ -82,22 +82,6 @@ class TypeCheckerTest extends FunSuite with ShouldMatchers {
     assert(let->output === empty)
   }
 
-  test("update term") {
-    val updateExpr = funv(unitv)
-    val update = Update("x", updateExpr)
-    val value = unitv
-    val let = LetBind("x", value, update)
-    checkType(let, unitt)
-    assert(let->input === Map.empty)
-    assert(let->output === Map.empty)
-    assert(value->ttype === unitt)
-    assert(updateExpr->input === Map.empty)
-    assert(updateExpr->ttype == funt(unitt))
-    assert(updateExpr->output === Map.empty)
-    assert(update->input === Map("x" -> unitt))
-    assert(update->output === Map("x" -> funt(unitt)))
-  }
-
   test("invalid parameter type") {
     val application = FunCall("f", Seq("x"))
     val letParam = LetBind("x", unitv, application)
@@ -116,36 +100,32 @@ class TypeCheckerTest extends FunSuite with ShouldMatchers {
     checkType(ift, boolt)
   }
 
-  test("if-then-else which changes variables") {
-    val ift = If(truev, "x" := truev, "x" := falsev)
-    val let = LetBind("x", unitv, ift)
+  test("if-then-else with different branch effects") {
+    val xObjStates = 
+      Seq(
+        StateDef("S1", 
+          Seq(MethodDef("m", unitv, "S2"), 
+              MethodDef("n", truev, "S2"))),
+        StateDef("S2", Seq.empty)
+      )
 
-    checkType(let, unitt)
-    assert(ift->ttype === unitt)
-    assert(ift.whenTrue->output === Map("x" -> boolt))
-    assert(ift.whenFalse->output === Map("x" -> boolt))
-    assert(ift->output === Map("x" -> boolt))
-  }
+    val xObjTypeStates =
+      Seq(
+        StateSpec("S1",
+          Seq(
+            MethodSpec("m", unitt, "S2"),
+            MethodSpec("n", boolt, "S2")
+          )),
+        StateSpec("S2", Seq.empty)
+      )
 
-  test("if-then-else with incompatible variable changes") {
-    val ift = If(truev, 
-      seq("x" := falsev, "y" := unitv), 
-      seq("x" := unitv, "y" := truev)
-    )
-    val let1 = LetBind("x", unitv, ift)
-    val let2 = LetBind("y", unitv, let1)
+    val xObjStart = ObjValue(xObjStates, "S1")
[…]

### 2013-01-07 src/test/scala/TypeOpTest.scala

@@ -44,7 +44,7 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
     }
   }
 
-  val joinOp = (_ : Type).join((_ : Type))
+  val joinOp = (x : Type, y : Type) => Option(x.join(y))
   val meetOp = (_ : Type).meet((_ : Type))
 
   def testJoin(expected : Type, t1 : Type, t2 : Type) {
@@ -64,7 +64,7 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
   }
 
   def testIncompatible(t1 : Type, t2 : Type) {
-    testCannotJoin(t1, t2)
+    testJoin(TopType(), t1, t2)
     testCannotMeet(t1, t2)
   }
 
@@ -94,5 +94,7 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
   testJoin(unitt, unitt, unitt)
   testJoin(boolt, boolt, boolt)
   testJoin(funt(unitt), funt(unitt), funt(unitt))
+
+  testJoin(topt, unitt, boolt)
 
 }

### 2013-01-07 src/main/scala/Parser.scala
bool and top types now parsed for explicit type annotations

@@ -48,8 +48,10 @@ trait Parser extends org.kiama.util.PositionedParserUtilities {
 
   lazy val typespec : PackratParser[Type] =
     "Unit" ^^ (_ => UnitType()) |
+    "Bool" ^^ (_ => BoolType()) |
     ("{" ~> rep1(statespec) <~ "}" <~ "@") ~ ident ^^ ObjType |
-    ("(" ~> repsep(effectType, ",") <~ "->") ~ typespec ^^ FunType
+    ("(" ~> repsep(effectType, ",") <~ "->") ~ typespec ^^ FunType |
+    ("Top" | "⊤") ^^ (_ => TopType())
 
   lazy val statespec : PackratParser[StateSpec] =
     (ident <~ "{") ~ rep(methodspec) <~ "}" ^^ StateSpec

### 2013-01-07 src/test/scala/ParserTest.scala

@@ -23,84 +23,67 @@ class ParserTest extends FunSuite {
   def checkTerm(termStr : String, expected : Term) = 
     assert(parse(termStr) === Left(expected))
 
-	test("parse unit value") {
-		checkTerm("unit", unitv)
-	}
+  def parseTest(termStr : String, expected : => Term) =
+    test("parse " + termStr) { 
+      assert(parse(termStr) === Left(expected))
+    }
 
-  test("parse zero argument function") {
-    checkTerm("\\().unit", funv(unitv))
-  }
+  // basic values
+	parseTest("unit", unitv)
+  parseTest("true", truev)
+  parseTest("false", falsev)
 
-  test("parse multi argument (no types) function") {
-    checkTerm(
-      "\\(x,y).unit",
-      funv(unitv, p("x"), p("y"))
-    )
-  }
-
-  test("parse function with typed argument") {
-    checkTerm(
-      "\\(x : Unit >> Unit).unit",
-      funv(unitv, p("x", unitt >> unitt))
-    )
-  }
-
-  test("parse function with object type argument") {
+  // function values
+  parseTest("\\().unit", funv(unitv))
+  parseTest("\\(x,y).unit", funv(unitv, p("x"), p("y")))
+  parseTest("\\(x : Unit >> Unit).unit", funv(unitv, p("x", unitt >> unitt)))
+  parseTest("\\(x : Bool >> Bool).unit", funv(unitv, p("x", boolt >> boolt)))
+  parseTest("\\(x : Top >> Top).unit", funv(unitv, p("x", topt >> topt)))
+  parseTest("\\(x : {S{a : Unit >> S}}@S >> {S{a : Unit >> S}}@S).unit", {
     val objType = 
       ObjType(Seq(StateSpec("S", Seq(MethodSpec("a", unitt, "S")))), "S")
+    funv(unitv, p("x", objType >> objType))
+  })
 
-    checkTerm(
-      "\\(x : {S{a : Unit >> S}}@S >> {S{a : Unit >> S}}@S).unit",
-      funv(unitv, p("x", objType >> objType))
-    )
-  }
-
-  test("parse minimal object value") {
-    checkTerm(
-      "[S{}]@S",
+  // object values
+  parseTest("[S{}]@S", ObjValue(Seq(StateDef("S", Seq())), "S"))
+  parseTest("[ A{ a=(unit,B) } B { b=(unit,A) } ]@A",
       ObjValue(
-        Seq(StateDef("S", Seq())),
-        "S"
[…]

### 2013-01-07 src/test/scala/ParserTest.scala
small additional check on precedence of parsing seq vs let binding

@@ -82,8 +82,8 @@ class ParserTest extends FunSuite {
     //        ^  semantically invalid but syntactically fine
 
   // parsing precedence of seq and let binding
-  parseTest("let x = \\().unit in x() ; unit",
-    LetBind("x", funv(unitv), seq(FunCall("x", Seq.empty), unitv)))
+  parseTest("let x = \\().unit in x() ; unit ; true",
+    LetBind("x", funv(unitv), seq(FunCall("x", Seq.empty), unitv, truev)))
   parseTest("(let x = unit in unit) ; unit",
     seq(LetBind("x", unitv, unitv), unitv))
 }

### 2013-01-07 src/main/scala/AST.scala
fixed parsing of object types (brought in line with paper), object values and types now checked for duplicate states, methods and missing states

@@ -31,6 +31,29 @@ case class ObjValue(states : Seq[StateDef], state : String) extends Value {
       ((m, s) => m + (s.name -> s))
     )
 
+  def validate() : Seq[ObjValidationError] = {
+    var errors = Seq.empty[ObjValidationError]
+    def checkExists(s : String, from : SourceElement) = {
+      if(!stateMap.contains(s)) errors +:= MissingState(s)(Some(from))
+    }
+
+    states.groupBy(_.name).foreach { case (state,dups) =>
+      if(dups.size > 1) errors +:= DuplicateState(state)(Some(dups.head))
+    }
+
+    checkExists(state, this)
+    states.foreach(state => {
+      
+      state.methods.groupBy(_.name).foreach { case (method, dups) =>
+        if(dups.size > 1) 
+          errors +:= DuplicateMethod(state.name, method)(Some(dups.head))
+      }
+
+      state.methods.foreach(method => checkExists(method.nextState, method))
+    })
+    errors
+  }
+
   override def toString = states.mkString("{", " ", "}") + "@" + state
 }
 
@@ -83,14 +106,17 @@ case class ParamDef(name : String, typeInfo : Option[EffectType])
 case class ErrorValue() extends Value
 
 /* validation errors */
-case class MissingState(
-  state : String, 
-  objValue : ObjValue, 
-  refPoint : SourceElement)
+sealed abstract class ObjValidationError {
+  val refPoint : Option[SourceElement]
+}
+case class DuplicateState(state : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
+case class MissingState(state : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
+case class DuplicateMethod(state : String, method : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
+
 
 /* types */
 
-sealed abstract class Type extends Attributable {
+sealed abstract class Type extends SourceElement {
   def >>(outType : Type) = EffectType(this, outType)
   def join(other : Type) : Type = (this, other) match {
     case (UnitType(), UnitType()) => UnitType()
@@ -169,12 +195,12 @@ case class FunType(params : Seq[EffectType], ret : Type) extends Type {
 case class ErrorType() extends Type
 
 case class ObjType(states : Seq[StateSpec], state : String) extends Type {
-  val stateMap : Map[String,StateSpec] =
+  lazy val stateMap : Map[String,StateSpec] =
[…]

### 2013-01-07 src/main/scala/ObjectValidator.scala

@@ -17,27 +17,16 @@ import org.kiama.attribution.Attributable
 
 object ObjectValidator {
 
-  val objectErrors : Attributable => List[MissingState] =
+  val objectErrors : Attributable => Seq[ObjValidationError] =
     attr {
-      case o : ObjValue => {
-        var errors = List.empty[MissingState]
-        if (!o.stateMap.contains(o.state))
-          errors = MissingState(o.state, o, o) :: errors
-
-        o.states.flatMap(_.methods).foreach(m => {
-          if (!o.stateMap.contains(m.nextState)) 
-            errors = MissingState(m.nextState, o, m) :: errors
-        })
-
-        errors
-      }
-      case _ => List.empty[MissingState]
+      case o : ObjValue => o.validate()
+      case o : ObjType => o.validate()
+      case _ => Seq.empty[ObjValidationError]
     }
 
-  def allObjectErrors(a : Attributable) : List[MissingState] =
+  def allObjectErrors(a : Attributable) : Seq[ObjValidationError] =
     (a.children.foldLeft
       (a->objectErrors)
       ((errs, child) => errs ++ allObjectErrors(child))
     )
-
 }

### 2013-01-07 src/main/scala/Parser.scala

@@ -54,11 +54,15 @@ trait Parser extends org.kiama.util.PositionedParserUtilities {
     ("Top" | "⊤") ^^ (_ => TopType())
 
   lazy val statespec : PackratParser[StateSpec] =
-    (ident <~ "{") ~ rep(methodspec) <~ "}" ^^ StateSpec
+    (ident <~ "{") ~ repsep(methodspec, ";") <~ "}" ^^ StateSpec
 
   lazy val methodspec : PackratParser[MethodSpec] =
-    (ident <~ ":") ~ (typespec <~ ">>") ~ ident ^^ MethodSpec
+    (ident <~ ":") ~ (typespec <~ "=>") ~ ident ^^ MethodSpec
 
   def ident: Parser[String] =
     """[a-zA-Z_]\w*""".r
+
+  def checkObj(obj : ObjType) = {
+    obj.validate
+  }
 }

### 2013-01-07 src/main/scala/TypeChecker.scala

@@ -92,8 +92,16 @@ object TypeChecker {
         if(objErrors.isEmpty) {
           ObjType(states.map(sSpec _), state)
         } else {
-          objErrors.foreach(err => 
-            message(err.refPoint, "missing state: " + err.state))
+          objErrors.foreach(err => {
+            val msg = err match {
+              case DuplicateState(state) => "duplicate state: " + state
+              case MissingState(state) => "missing state: " + state
+              case DuplicateMethod(state, method) =>
+                "duplicate method in state " + state + ": " + method
+            }
+
+            message(err.refPoint.getOrElse(o), msg)
+          })
           ErrorType()
         }
       }
@@ -389,6 +397,7 @@ class FullTracePrinter(term : Terminal) extends org.kiama.output.PrettyPrinter {
       val pDocs = params map showEffect
       parens(ssep(pDocs, ",")) <+> "->" <+> showType(ret)
     }
+    case TopType() => "Top"
     case ErrorType() => "BAD"
     case ObjType(states, state) => {
       val sDocs = states map (showStateSpec _)

### 2013-01-07 src/test/scala/ObjectValidatorTest.scala

@@ -20,7 +20,7 @@ class ObjectValidatorTest extends FunSuite with Parser {
 
   def parseTerm(termStr : String) = parseString(term, termStr)
 
-  def objTest(termStr : String, errors : List[MissingState]) = 
+  def objTest(termStr : String, errors : Seq[ObjValidationError]) = 
     test(termStr + " is " + (if (errors.isEmpty) "valid" else "invalid")) {
       val result = parseTerm(termStr)
       
@@ -29,28 +29,44 @@ class ObjectValidatorTest extends FunSuite with Parser {
           initTree(t)
           val foundErrors = allObjectErrors(t)
           assert(foundErrors.length === errors.length)
-          errors.foreach(e => {
-            assert(foundErrors.find(_.state == e.state).isDefined === true)
-          })
+          errors.foreach(e => assert(foundErrors.contains(e)))
         }
         case Right(_) => fail("unable to parse term " + termStr)
       }
     }
 
   def validObjTest(termStr : String) = 
-    objTest(termStr, Nil)
+    objTest(termStr, Seq.empty)
 
-  def invalidObjTest(termStr : String, missingStates : String*) = 
-    objTest(termStr, missingStates.map(s => MissingState(s, null, null)).toList)
+  def invalidObjTest(
+    termStr : String, 
+    expectedErrors : ObjValidationError*) =
+    objTest(termStr, expectedErrors)
 
   validObjTest("[ S {} ]@S")
   validObjTest("[ S{ m = (unit, S) } ]@S")
   validObjTest("[ S { m = (unit, S2) } S2 {} ]@S2")
 
-  invalidObjTest("[ S2 {} ]@S", "S")
-  invalidObjTest("[ S { m = (unit, S2) } ]@S", "S2")
+  invalidObjTest("[ S2 {} ]@S", MissingState("S")())
+  invalidObjTest("[ S { m = (unit, S2) } ]@S", MissingState("S2")())
 
   invalidObjTest(
     "let x = [ S { m = (unit, S2) } ]@S in [ A { n = (unit, A) } ]@B",
-    "S2", "B")
+    MissingState("S2")(), MissingState("B")())
+
+  invalidObjTest("[ S { m = (unit, S) ; m = (unit, S2) } ]@S",
+    MissingState("S2")(), DuplicateMethod("S", "m")()
+    )
+
[…]

### 2013-01-07 src/test/scala/ParserTest.scala

@@ -39,7 +39,7 @@ class ParserTest extends FunSuite {
   parseTest("\\(x : Unit >> Unit).unit", funv(unitv, p("x", unitt >> unitt)))
   parseTest("\\(x : Bool >> Bool).unit", funv(unitv, p("x", boolt >> boolt)))
   parseTest("\\(x : Top >> Top).unit", funv(unitv, p("x", topt >> topt)))
-  parseTest("\\(x : {S{a : Unit >> S}}@S >> {S{a : Unit >> S}}@S).unit", {
+  parseTest("\\(x : {S{a : Unit => S}}@S >> {S{a : Unit => S}}@S).unit", {
     val objType = 
       ObjType(Seq(StateSpec("S", Seq(MethodSpec("a", unitt, "S")))), "S")
     funv(unitv, p("x", objType >> objType))

### 2013-01-07 src/test/scala/TypeCheckerTest.scala

@@ -42,7 +42,6 @@ class TypeCheckerTest extends FunSuite with ShouldMatchers {
     })
   }
 
-
   val empty = Map.empty[String,Type]
 
   test("unit value type") {

### 2013-01-08 src/main/scala/AST.scala
fixing join/meet with respect to top type

@@ -14,6 +14,7 @@ package uk.ac.gla.dcs.ts
 import org.kiama.attribution.Attributable
 import org.kiama.util.Positioned
 
+/** standard trait for anything that appears in source code */
 trait SourceElement extends Attributable with Positioned
 
 /* terms */
@@ -105,71 +106,55 @@ case class ParamDef(name : String, typeInfo : Option[EffectType])
 
 case class ErrorValue() extends Value
 
-/* validation errors */
+/* object validation errors */
 sealed abstract class ObjValidationError {
   val refPoint : Option[SourceElement]
 }
-case class DuplicateState(state : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
-case class MissingState(state : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
-case class DuplicateMethod(state : String, method : String)(val refPoint : Option[SourceElement] = None) extends ObjValidationError
+
+case class DuplicateState(state : String)
+                         (val refPoint : Option[SourceElement] = None) 
+                         extends ObjValidationError
+
+case class MissingState(state : String)
+                       (val refPoint : Option[SourceElement] = None) 
+                       extends ObjValidationError
+
+case class DuplicateMethod(state : String, method : String)
+                          (val refPoint : Option[SourceElement] = None) 
+                          extends ObjValidationError
 
 
 /* types */
 
 sealed abstract class Type extends SourceElement {
   def >>(outType : Type) = EffectType(this, outType)
+
+  /**
+   * Calculates the join of two types. This is always defined, as the type
+   * Top (⊤) is a supertype of everything.
+   */
   def join(other : Type) : Type = (this, other) match {
+    case (TopType(), _) => TopType()
+    case (_, TopType()) => TopType()
     case (UnitType(), UnitType()) => UnitType()
     case (BoolType(), BoolType()) => BoolType()
-    case (FunType(f1p, f1r), FunType(f2p, f2r)) => {
-      if(f1p.size != f2p.size) TopType()
-      else {
-        val retType = f1r join f2r
-        val paramTypesOpt = 
-          (f1p.zip(f2p).foldRight
[…]

### 2013-01-08 src/test/scala/TypeOpTest.scala

@@ -17,7 +17,6 @@ import TestUtils._
 
 class TypeOpTest extends FunSuite with ShouldMatchers {
 
-
   def opTest(expected : Type, 
     t1 : Type, 
     t2 : Type,
@@ -64,7 +63,7 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
   }
 
   def testIncompatible(t1 : Type, t2 : Type) {
-    testJoin(TopType(), t1, t2)
+    testJoin(topt, t1, t2)
     testCannotMeet(t1, t2)
   }
 
@@ -78,23 +77,53 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
   testIncompatible(boolt, emptyObj)
   testIncompatible(funt(unitt), emptyObj)
 
+  
   // mismatched parameter lengths
   testIncompatible(funt(unitt), funt(unitt, unitt >> unitt))
 
-  // incompatible return types
-  testIncompatible(funt(unitt), funt(boolt))
+  // mismatched return types
+  testJoin(funt(topt), funt(unitt), funt(boolt))
+  testCannotMeet(funt(unitt), funt(boolt))
 
-  // incompatible output type on effect
-  testIncompatible(funt(unitt, unitt >> unitt), funt(unitt, unitt >> boolt))
+  // mismatched output type on effect
+  testJoin(funt(unitt, unitt >> topt), 
+           funt(unitt, unitt >> unitt), 
+           funt(unitt, unitt >> boolt))
 
-  // incompatible input type on effect
-  testIncompatible(funt(unitt, unitt >> unitt), funt(unitt, boolt >> unitt))
+  testCannotMeet(funt(unitt, unitt >> unitt), funt(unitt, unitt >> boolt))
+
+  // mismatched input type on effect
+  testJoin(topt, funt(unitt, unitt >> unitt), funt(unitt, boolt >> unitt))
+  testMeet(funt(unitt, topt >> unitt),
+           funt(unitt, unitt >> unitt), 
+           funt(unitt, boolt >> unitt))
+
 
   // join with self
   testJoin(unitt, unitt, unitt)
   testJoin(boolt, boolt, boolt)
   testJoin(funt(unitt), funt(unitt), funt(unitt))
+  testJoin(emptyObj, emptyObj, emptyObj)
+  testJoin(topt, topt, topt)
 
-  testJoin(topt, unitt, boolt)
[…]

### 2013-01-08 src/main/scala/FirstOrderUnification.scala
using newly learned trick to exclude counter from default equals and hashcode implementation for multi equation

@@ -37,30 +37,33 @@ case class MultiTerm(fn : Int, var args : List[TempMultiEquation]) {
   override def toString = "F" + fn + "(" + args.mkString(", ") + ")"
 }
 
-case class MultiEquation(
-  var counter : Int,
-  var s : Set[Variable],
-  var m : Option[MultiTerm]) {
+case class MultiEquation(var s : Set[Variable], var m : Option[MultiTerm])
+                        (var counter : Int) {
 
   def merge(other : MultiEquation) : MultiEquation =
     if(this == other) {
       this
     } else {
       val merged = MultiEquation(
-        this.counter + other.counter, 
         this.s ++ other.s, 
-        this.m.map(_.merge(other.m)).orElse(other.m)
-      )
+        this.m.map(_.merge(other.m)).orElse(other.m))(
+        this.counter + other.counter)
       merged.s.foreach(v => { v.m = merged })
       merged
     }
 
-  override def equals(o : Any) = o match {
-    case MultiEquation(_,s2,m2) => s == s2 && m == m2
-  }
-
   override def toString =
     "[" + counter + "] {" + s.mkString(", ") + "} = " + m.map(_.toString).getOrElse("∅")
+}
+
+object MultiEquation {
+
+  /**
+   * alternative "constructor" for case class with the old signature, to avoid
+   * changing all the existing test code
+   */
+  def apply(counter : Int, s : Set[Variable], m : Option[MultiTerm]) : MultiEquation = 
+    MultiEquation(s, m)(counter)
 }
 
 case class TempMultiEquation(

### 2013-01-08 src/main/scala/UnificationProblemBuilder.scala

@@ -88,7 +88,7 @@ class UnificationProblemBuilder {
           case (te1, te2) => {
             val (mt1, vmt1) = typeToMultiTerm(te1)
             val (mt2, vmt2) = typeToMultiTerm(te2)
-            val merged = MultiEquation(0, Set.empty, Some(mt1.merge(Some(mt2))))
+            val merged = MultiEquation(Set.empty, Some(mt1.merge(Some(mt2))))(0)
 
             log.debug("eq = " + merged)
 
@@ -133,7 +133,7 @@ class UnificationProblemBuilder {
 
     val (eq, varsInEq) = eqsByVar.getOrElse(v, emptyEq(v))
     val (multiTerm,varsInMultiTerm) = typeToMultiTerm(te)
-    val newEq = MultiEquation(0, Set(getVarSpec(v)), Some(multiTerm))
+    val newEq = MultiEquation(Set(getVarSpec(v)), Some(multiTerm))(0)
     val merged = Pair(newEq.merge(eq), varsInMultiTerm ++ varsInEq)
 
     log.debug("eq = " + merged._1)
@@ -145,7 +145,7 @@ class UnificationProblemBuilder {
   }
 
   private def emptyEq(v : TypeVar) : EqWithRefs = 
-    Pair(MultiEquation(0, Set(getVarSpec(v)), None),Set.empty[TypeVar])
+    Pair(MultiEquation(Set(getVarSpec(v)), None)(0),Set.empty[TypeVar])
 
   private def fixReferenceCount(
     vs : Set[TypeVar], 
@@ -157,11 +157,11 @@ class UnificationProblemBuilder {
         val newEq = m2.get(v) match {
           case None => 
             Pair(
-              MultiEquation(max(0, d), Set(getVarSpec(v)), None), 
+              MultiEquation(Set(getVarSpec(v)), None)(max(0, d)), 
               Set.empty[TypeVar]
             )
-          case Some(Pair(MultiEquation(c, s, m),refs)) => 
-            Pair(MultiEquation(c + d, s, m), refs)
+          case Some(Pair(meq @ MultiEquation(s, m),refs)) => 
+            Pair(MultiEquation(s, m)(meq.counter + d), refs)
         }
         pointVarsToEq(m2, newEq)
       })

### 2013-01-08 src/main/scala/FirstOrderUnification.scala
using new trick to exclude multi equation component of Variable definition from default equals and hash code.

@@ -81,14 +81,8 @@ case class TempMultiEquation(
 }
 
 /** !!! cyclic references between meqs and vars. Need to unroll somehow */
-case class Variable(val num : Int, var m : MultiEquation) {
-
+case class Variable(val num : Int)(var m : MultiEquation) {
   override def toString = "v" + num
-  override def equals(o : Any) = o match { 
-    case Variable(num2,_) => num == num2
-    case _ => false
-  }
-  override def hashCode = num.hashCode
 }
 
 /**

### 2013-01-08 src/main/scala/UnificationProblemBuilder.scala

@@ -53,7 +53,7 @@ class UnificationProblemBuilder {
 
   private def getVarSpec(tv : TypeVar) = {
     if(!vars.contains(tv)) {
-      vars = vars.updated(tv, Variable(tv.v, null))
+      vars = vars.updated(tv, Variable(tv.v)(null))
     }
     vars(tv)
   }

### 2013-01-08 src/test/scala/TestUtils.scala

@@ -47,7 +47,7 @@ object TestUtils {
   /* unification related helpers */
   val unitMT = MultiTerm(0, List.empty)
 
-  def v(tv : TypeVar) = Variable(tv.v, null)
+  def v(tv : TypeVar) = Variable(tv.v)(null)
   def vset(tvs : TypeVar*) = Set(tvs.map(tv => v(tv)) :_*)
 
 

### 2013-01-08 src/main/scala/AST.scala
state sets added to the basic type system

@@ -134,14 +134,14 @@ sealed abstract class Type extends SourceElement {
    * Top (⊤) is a supertype of everything.
    */
   def join(other : Type) : Type = (this, other) match {
+    case (ErrorType(), _) => ErrorType()
+    case (_, ErrorType()) => ErrorType()
     case (TopType(), _) => TopType()
     case (_, TopType()) => TopType()
     case (UnitType(), UnitType()) => UnitType()
     case (BoolType(), BoolType()) => BoolType()
     case (a @ FunType(_, _), b @ FunType(_, _)) => a.joinFun(b)
     case (a @ ObjType(_, _), b @ ObjType(_, _)) => a.joinObj(b)
-    case (ErrorType(), _) => ErrorType()
-    case (_, ErrorType()) => ErrorType()
     case _ => TopType()
   }
 
@@ -211,23 +211,26 @@ case class FunType(params : Seq[EffectType], ret : Type) extends Type {
 
 case class ErrorType() extends Type
 
-case class ObjType(states : Seq[StateSpec], state : String) extends Type {
+
+/**
+ * Object types, which are composed of a finite set of states with
+ * deterministic transitions, and a current state set. A real object value
+ * of this type will have precisely one current state; the state set is used
+ * to capture runtime non-determinism in the type system.
+ */
+case class ObjType(states : Seq[StateSpec], currentStateSet : Set[String]) extends Type {
+
   lazy val stateMap : Map[String,StateSpec] =
     (states.foldLeft
       (Map.empty[String,StateSpec])
       ((map,state) => map + Pair(state.name, state)))
-
-  lazy val currentState = stateMap get state
-    
-  def retType(method : String) = 
-    currentState flatMap (s => s.retType(method)) getOrElse ErrorType()
-
-  def joinObj(other : ObjType) : ObjType = {
-    // TODO: implement properly
-    if(this == other) this
-    else ObjType(Seq(StateSpec("S", Seq.empty)), "S")
-  }
-
+  
+  /**
+   * Checks that the object type is free of the following errors:
+   * - duplicate state specifications
+   * - duplicate methods within state specifications
+   * - references to missing state specifications
+   */
   def validate() : Seq[ObjValidationError] = {
     var errors = Seq.empty[ObjValidationError]
     def checkExists(s : String, from : SourceElement) = {
@@ -238,7 +241,7 @@ case class ObjType(states : Seq[StateSpec], state : String) extends Type {
[…]

### 2013-01-08 src/main/scala/Parser.scala

@@ -49,7 +49,7 @@ trait Parser extends org.kiama.util.PositionedParserUtilities {
   lazy val typespec : PackratParser[Type] =
     "Unit" ^^ (_ => UnitType()) |
     "Bool" ^^ (_ => BoolType()) |
-    ("{" ~> rep1(statespec) <~ "}" <~ "@") ~ ident ^^ ObjType |
+    ("{" ~> rep1(statespec) <~ "}" <~ "@") ~ stateset ^^ { case ~(a, b) => ObjType(a, b) } |
     ("(" ~> repsep(effectType, ",") <~ "->") ~ typespec ^^ FunType |
     ("Top" | "⊤") ^^ (_ => TopType())
 
@@ -58,6 +58,11 @@ trait Parser extends org.kiama.util.PositionedParserUtilities {
 
   lazy val methodspec : PackratParser[MethodSpec] =
     (ident <~ ":") ~ (typespec <~ "=>") ~ ident ^^ MethodSpec
+
+  lazy val stateset : PackratParser[Set[String]] =
+    ident ^^ (i => Set(i)) |
+    "{" ~> rep1sep(ident, ",") <~ "}" ^^ (_.toSet)
+
 
   def ident: Parser[String] =
     """[a-zA-Z_]\w*""".r

### 2013-01-08 src/main/scala/TypeChecker.scala

@@ -194,15 +194,10 @@ object TypeChecker {
         ErrorType()
       }
       case Some(o @ ObjType(states,_)) => {
-        val nextState = o.currentState flatMap (s => s.nextState(t.methName))
-        nextState match {
-          case None => {
-            message(t, "attempt to call unavailable method %s on receiver %s of type %s".format(t.methName, t.objVarName, o))
-            ErrorType()
-          }
-          case Some(state) => {
-            ObjType(states, state)
-          }
+        val nextStateSet = o.nextStateSet(t.methName)
+        nextStateSet.map(ObjType(states, _)).getOrElse {
+          message(t, "attempt to call unavailable method %s on receiver %s of type %s".format(t.methName, t.objVarName, o))
+          ErrorType()
         }
       }
       case Some(ErrorType()) => ErrorType()
@@ -399,10 +394,16 @@ class FullTracePrinter(term : Terminal) extends org.kiama.output.PrettyPrinter {
     }
     case TopType() => "Top"
     case ErrorType() => "BAD"
-    case ObjType(states, state) => {
+    case ObjType(states, currentStateSet) => {
       val sDocs = states map (showStateSpec _)
-      braces(space <> nest(fillsep(sDocs, space)) <> line) <> "@" <> state
+      (braces(space <> nest(fillsep(sDocs, space)) <> line) <> 
+        "@" <> showStateSet(currentStateSet))
     }
+  }
+
+  def showStateSet(states : Set[String]) = {
+    if(states.size == 1) text(states.head)
+    else braces(lsep(states.map(text(_)).toSeq, ","))
   }
 
   def showStateSpec(s : StateSpec) = {

### 2013-01-08 src/test/scala/ObjTypeTest.scala

@@ -1,0 +1,64 @@
+/*
+ * This file is part of TS.
+ * 
+ * TS - Copyright (c) 2012, Iain McGinniss
+ * All rights reserved.
+ * 
+ * Redistribution and use in source and binary forms, with or without
+ * modification, are permitted under the terms of the 2-clause BSD license:
+ * http://opensource.org/licenses/BSD-2-Clause
+ */
+
+package uk.ac.gla.dcs.ts
+
+import org.scalatest.FunSuite
+import org.scalatest.matchers.ShouldMatchers
+import TestUtils._
+
+class ObjTypeTest extends FunSuite with ShouldMatchers {
+
+  val o1states = Seq(
+      StateSpec("A", Seq(
+        MethodSpec("m", unitt, "A"),
+        MethodSpec("n", boolt, "B"),
+        MethodSpec("g", funt(boolt), "B")
+      )),
+      StateSpec("B", Seq(
+        MethodSpec("m", boolt, "B"),
+        MethodSpec("f", unitt, "A"),
+        MethodSpec("g", funt(unitt), "B")
+      ))
+    )
+
+  test("return type with single known current state") {
+    val obj = ObjType(o1states, Set("A"))
+    (obj.retType("m")) should be (unitt)
+    (obj.retType("n")) should be (boolt)
+    (obj.retType("f")) should be  (ErrorType())
+  }
+
+  test("return type with state set") {
+    val obj = ObjType(o1states, Set("A", "B"))
+    (obj.retType("m")) should be (topt)
+    (obj.retType("n")) should be (ErrorType())
+    (obj.retType("f")) should be (ErrorType())
+    (obj.retType("g")) should be (funt(topt))
+  }
+
+  test("nextStateSet with single known current state") {
+    val obj = ObjType(o1states, Set("B"))
+    (obj.nextStateSet("m")) should be (Some(Set("B")))
+    (obj.nextStateSet("n")) should be (None)
+    (obj.nextStateSet("f")) should be (Some(Set("A")))
+    (obj.nextStateSet("g")) should be (Some(Set("B")))
+  }
+
+  test("nextStateSet with state set") {
+    val obj = ObjType(o1states, Set("A", "B"))
+    (obj.nextStateSet("m")) should be (Some(Set("A", "B")))
+    (obj.nextStateSet("n")) should be (None)
+    (obj.nextStateSet("f")) should be (None)
+    (obj.nextStateSet("g")) should be (Some(Set("B")))
+  }
+
+}

### 2013-01-08 src/test/scala/TypeOpTest.scala

@@ -126,4 +126,10 @@ class TypeOpTest extends FunSuite with ShouldMatchers {
   testMeet(funt(unitt), funt(unitt), topt)
   testMeet(emptyObj, emptyObj, topt)
 
+  // error types
+  testJoin(ErrorType(), topt, ErrorType())
+  testJoin(ErrorType(), boolt, ErrorType())
+  testJoin(ErrorType(), funt(unitt), ErrorType())
+  testJoin(ErrorType(), emptyObj, ErrorType())
+
 }
